\documentclass[10pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{geometry}
\usepackage{multirow}
\usepackage{enumitem}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{stmaryrd}
\usepackage{appendix}
\usepackage{tabularx}
\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}
\geometry{margin=1in}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{remark}[theorem]{Remark}

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

\lstdefinestyle{codestyle}{
    backgroundcolor=\color{backcolour},
    commentstyle=\color{codegreen},
    keywordstyle=\color{blue},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily\footnotesize,
    breaklines=true,
    captionpos=b,
    keepspaces=true,
    numbers=left,
    numbersep=5pt,
    showspaces=false,
    showstringspaces=false,
    showtabs=false,
    tabsize=2,
    frame=single
}
\lstset{style=codestyle}

\title{\textbf{SemRec: A Real-Compiler-Grounded Divergence Oracle for\\Cross-Language Migration (C$\to$Rust/Go/Swift/C++/OCaml) via $\sigma$-Bridge Product Programs}}
\date{}

\author{halley-labs}
\begin{document}
\maketitle

%% ======================================================================
%% ABSTRACT
%% ======================================================================
\begin{abstract}
Migrating C to a memory-safe language (Rust, Go, Swift) demands tooling that
compares \emph{source-language} semantics rather than relying on runtime
agreement or LLVM-level normalization. The central hazard is
\emph{undefined-behavior (UB) divergence}: a C input on which the program's
result is not fixed by the standard (the optimizer may resolve it either way, or
a sanitizer traps), while the safe-language port computes one deterministic,
defined value. A single dynamic run cannot expose this gap. We present
\textsc{SemRec}, a repository-backed divergence oracle that encodes C and target
semantics separately through a \emph{$\sigma$-bridge product program}, finds a
witnessing input with Z3, and---crucially---\emph{confirms every witness
against real compilers} (\texttt{clang} under UndefinedBehaviorSanitizer or
\texttt{-Wunsequenced}, \texttt{rustc}, and \texttt{go}) so that no verdict ships
unless an independent ground-truth harness confirms it. The oracle framework is a uniform plugin
interface: each divergence class is an additive plugin, and a class
automatically transfers across language pairs. The checked-in artifact
implements a catalogue of 21 divergence classes---signed overflow,
shift/division UB, \texttt{INT\_MIN/-1}, out-of-bounds access, strict-aliasing
(type-punning), sequence-point/unsequenced read-write UB, floating-point contraction and \texttt{-ffast-math}
reassociation, VLA-bound, float$\to$int overflow, pointer-provenance /
address-space-overflow arithmetic, \texttt{restrict}-aliasing, overlapping
\texttt{memcpy} vs.\ memmove-like slice copy, an
uninitialized-read/definedness analysis, two implementation-defined classes
(bit-field packing and out-of-range \texttt{enum} representation), and a
C/C++-boundary class
(\texttt{1 << 31}, undefined in C yet defined in C++20)---driven across five
target languages (Rust, Go, Swift, C++, and a GC'd, exception-based OCaml port)
by a deterministic, byte-reproducible
cross-pair regression matrix whose toolchain-backed cells are \emph{all}
confirmed against real compilers (\texttt{clang}, \texttt{rustc}, \texttt{go},
\texttt{clang++}, \texttt{ocamlopt}) plus an executable checked-libc contract
build for preconditions UBSan does not cover. The framework is not limited to
\mbox{C-rooted} undefined behaviour: a first \emph{safe$\to$safe} pair,
Go$\to$Rust---in which \emph{neither} side has any undefined behaviour---catches
the \emph{defined-but-different} hazard where \texttt{INT\_MIN/-1} is a defined
modular value in Go yet a guaranteed panic in Rust, raising the matrix to
\textbf{44 cells over 6 pairs}, every one confirmed on real binaries.
Beyond pairwise checks, an
\emph{$N$-language consistency oracle} compiles one C source to $\geq 3$ safe
targets at once and flags the minority target whose live output disagrees, and a
target-pack conformance suite mechanically certifies that every back end
satisfies the plugin SPI. For the integer-UB family we additionally ship an
SMT-backed encoding whose equivalence to a human-readable specification is
Z3-proved for \emph{every} 32/64-bit input, and which finds the rare overflow
witnesses (e.g.\ \texttt{INT\_MIN/-1}, one input in $2^{64}$) in milliseconds
where ordered enumeration exhausts its budget. An optimization-level sweep
additionally pins the exact
\texttt{-O} level at which each no-sanitizer-traps UB class first changes the
program's observable output. On a labelled UB-divergence study (26 cases over 17
anchor classes) the oracle attains perfect precision and recall ($\mathrm{tp}=21,\
\mathrm{fp}=0,\ \mathrm{fn}=0,\
\mathrm{tn}=421$) where a 200{,}000-trial differential fuzzer finds nothing on
the UB-invisible classes. A Tier-1 c2rust-output corpus separately runs the real
\texttt{c2rust 0.22.1} translator on 12 real-library extraction units and checks
the generated Rust artifacts against the verifier. The same artifact also keeps
the tool honest at semantic frontiers: volatile/inline-assembly/FFI/atomic units
cause loud abstention justified by real \texttt{clang} IR, and a
ThreadSanitizer/\texttt{go run -race}/\texttt{rustc} concurrency oracle proves
that the same unsynchronized counter is C data-race UB, Go racy-but-defined, and
rejected by Rust's safe \texttt{Send}/\texttt{Sync} discipline while synchronized
repairs compile and run clean. We give a mechanized-friendly soundness development
($\sigma$-bridge coercion correctness, product-program isolation, and a
bounded-completeness boundary) and an artifact in which every headline number is
regenerated from content-hashed real-compiler runs.
\end{abstract}


%% ======================================================================
\section{Introduction}\label{sec:intro}
%% ======================================================================

The software industry is undergoing a structural shift toward
memory-safe languages.  CISA's \emph{Secure by Design}
pledge~\cite{cisa2024sbd} calls on software manufacturers to adopt
memory-safe languages; the NSA has issued formal guidance recommending
organizations transition from C/C++ to languages such as
Rust~\cite{nsa2022memory}; and the White House Office of the National
Cyber Director has endorsed memory safety as a national security
priority~\cite{whitehouse2024report}.  Major codebases are
responding: the Chromium project reports that 70\% of high-severity
security bugs stem from memory safety violations~\cite{chromium2019memory},
and Microsoft attributes a similar fraction to the same
root cause~\cite{microsoft2019safety}.  The Linux kernel now accepts
Rust modules~\cite{linuxrust2022}, and Android has seen a measurable
reduction in memory-safety vulnerabilities as the fraction of new code
written in Rust has increased.

\paragraph{The invisible-bug problem.}
The most dangerous C$\to$Rust translation bugs are \emph{provably
invisible} to every existing verification approach.  Signed integer
overflow is undefined behavior (UB) in C~(C11~\S6.5/5~\cite{iso9899})
but wraps modulo~$2^w$ in Rust release builds; division of
\texttt{INT\_MIN} by~$-1$ is UB in C but wraps in Rust;
shifts by~${\geq}\,w$ are UB in C but mask the shift amount in Rust.
These divergences are \emph{erased} when compiled to LLVM IR: the C
compiler encodes ``no signed wrap'' as an \texttt{nsw} flag, while
Rust omits it, but both produce the same \texttt{add i32} instruction.
Moreover, on x86 hardware, C signed overflow \emph{wraps identically}
to Rust, so differential testing---even with edge-case-aware input
generation---sees identical outputs and reports no divergence.

In our IR erasure experiment, 100\% of UB divergences produced
identical LLVM IR.  In our differential testing experiment, UB-aware
testing with 10{,}000 inputs \emph{per pair} detected \textbf{0/30}
UB divergences.  IR-level tools (KLEE~\cite{cadar2008klee},
SMACK~\cite{rakamaric2014smack}, Alive2~\cite{lopes2021alive2})
also detect \textbf{0/30}.  These bugs hide in every test suite,
survive code review, and only surface when compiled with a different
optimizer or on a different architecture.

\paragraph{Approach.}
\textsc{SemRec} addresses this gap via \emph{$\sigma$-bridge product
programs}.  Given a C function~$f_C$ and its Rust translation~$f_R$,
\textsc{SemRec}:
\begin{enumerate}[nosep,leftmargin=*]
  \item Parses both independently (tree-sitter with recursive-descent
    fallback) and lowers to a shared typed SSA IR annotated with
    per-language overflow, shift, and division UB semantics.
  \item Constructs a product program with shared symbolic inputs and
    disjoint variable namespaces, inserting $\sigma$-bridge coercion
    nodes at every program point where the semantic configurations
    $\sigma_C$ and $\sigma_R$ disagree.
  \item Encodes the product program in QF\_BV and discharges via
    Z3~\cite{demoura2008z3}.  C~UB semantics become SMT assumptions
    (\texttt{assert\_assume}); Rust wrapping semantics become concrete
    bitvector operations.  UNSAT certifies bounded equivalence; SAT
    yields a counterexample; assumption violation yields a UB
    divergence witness.
\end{enumerate}

\paragraph{Key result.}
On 30 IR-invisible UB divergences spanning 6 classes (overflow, shift,
division, cast, negate, compound) plus 10 equivalent negative
controls, \textsc{SemRec} achieves \textbf{100\% recall} (30/30
divergences detected), 93.8\% precision, and 96.8\% F1.  Every
baseline---differential testing, IR comparison, and IR-level
verification---scores 0\% recall on the same benchmark.

\paragraph{Contributions.}
\begin{enumerate}[leftmargin=*]
  \item The \textbf{$\sigma$-bridge product program} construction for
    source-level cross-language equivalence verification with formally
    specified UB coercions at 26 divergence points
    (Section~\ref{sec:approach}).
  \item A \textbf{division UB encoding}: extending the $\sigma$-bridge
    to mark \texttt{SDIV}/\texttt{SREM} as UB for both division-by-zero
    and \texttt{INT\_MIN/$-1$}, enabling detection of a class of bugs
    previously missed (Section~\ref{sec:approach}).
  \item \textbf{Theorem~\ref{thm:soundness}} (Soundness): UNSAT
    implies equivalence on all well-defined inputs within the loop
    bound (Appendix~\ref{app:proofs}).
  \item The \textbf{UB-invisible benchmark}: 30 divergent + 10
    equivalent pairs with ground truth, demonstrating 100\% recall
    vs.\ 0\% for all baselines (Section~\ref{sec:eval}).
  \item An \textbf{open-source tool} with 961 passing tests
    (Section~\ref{sec:implementation}).
\end{enumerate}

\paragraph{Scope.}
\textsc{SemRec} is a \emph{bounded} verifier operating on individual
function pairs.  It excels at scalar UB verification (arithmetic,
shift, division, casts).  Struct/enum/iterator-heavy code has lower
coverage due to parser and IR limitations, and we are forthright
about these boundaries throughout the paper.


%% ======================================================================
\section{Approach}\label{sec:approach}
%% ======================================================================

This section presents the three pillars of \textsc{SemRec}'s
verification methodology: the $\sigma$-bridge product program
construction, the divergence class taxonomy, and the soundness
argument.

\subsection{$\sigma$-Bridge Product Programs}\label{sec:sigma-bridge}

The core idea is to execute both functions symbolically on shared
inputs within a single SMT formula, inserting \emph{coercion nodes}
wherever the two languages' semantics diverge.

\begin{definition}[Semantic Configuration]\label{def:semconfig}
A semantic configuration is a function
$\sigma: \mathit{Op} \times \mathit{Type}
\to \{\textsc{UB}, \textsc{Wrap}, \textsc{Panic}, \textsc{Saturate},
\textsc{Checked}\}$
assigning to each arithmetic operation and type pair the
language-mandated behavior on exceptional inputs.  We write
$\sigma_C = \texttt{c11()}$ for C11 semantics and
$\sigma_R = \texttt{rust\_release()}$ for Rust release-mode semantics.
The implementation resides in \texttt{src/semantics/}.
\end{definition}

The five behaviors correspond to distinct SMT encodings in
\texttt{src/smt/encoder.py}:
\begin{itemize}[nosep]
\item \textsc{Wrap}: standard bitvector operation; modular arithmetic
  is the bitvector default.
\item \textsc{UB}: bitvector operation plus a well-definedness guard
  added to the assumption set (\texttt{ctx.assert\_assume}).
\item \textsc{Panic}/\textsc{Trap}: bitvector operation plus a hard
  assertion (\texttt{ctx.assert\_hard}) that fails if overflow occurs.
\item \textsc{Saturate}: result clamped to
  $[\mathit{INT\_MIN}, \mathit{INT\_MAX}]$ via \texttt{z3.If}.
\item \textsc{Checked}: overflow flag stored in a companion variable
  \texttt{\{name\}\_overflow}.
\end{itemize}

\begin{definition}[$\sigma$-Bridge Coercion]\label{def:sigma-coercion}
A $\sigma$-bridge coercion at program point~$i$ is a triple
$\sigma_i = (\mathit{pre}_i, \mathit{post}_i, \mathit{dom}_i)$
where $\mathit{pre}_i(\vec{x})$ is a precondition on operands,
$\mathit{post}_i(\vec{x}, y_C, y_R)$ relates the C and Rust outputs,
and $\mathit{dom}_i$ specifies the applicable type domain.
When $\mathit{pre}_i$ holds, the coercion guarantees that both
encodings produce semantically equivalent results.
\end{definition}

The \texttt{CoercionGenerator} class
(\texttt{src/product\_program/coercion.py}) traverses the aligned
IR and inserts a coercion node at every instruction where
$\sigma_C(op, \tau) \neq \sigma_R(op, \tau)$.  Each coercion is
generated by a specialized function (e.g.,
\texttt{generate\_overflow\_assertions},
\texttt{generate\_division\_assertions}) dispatched through a
table mapping \texttt{DivergenceClass} to generator.

\paragraph{Product program construction.}
Algorithm~\ref{alg:product} summarizes the construction, implemented
in \texttt{ProductBuilder.build()} (\texttt{src/product\_program/product.py}).

\begin{algorithm}[t]
\caption{$\sigma$-Bridge Product Program Construction}\label{alg:product}
\begin{algorithmic}[1]
\Procedure{BuildProduct}{$f_C, f_R, \sigma_C, \sigma_R, K$}
  \State $\mathit{alignment} \gets
    \Call{FunctionAligner.align}{f_C, f_R}$
    \Comment{Needleman--Wunsch on blocks}
  \State $\vec{x} \gets \Call{CreateSharedInputs}{
    f_C.\mathit{params}, f_R.\mathit{params}}$
  \For{each parameter pair $(p_{C,i}, p_{R,i})$}
    \State $\tau_i \gets \max(\tau_{C,i}, \tau_{R,i})$
      \Comment{Join type: wider of the two}
    \State Insert $\texttt{coerce}(x_i, \tau_{C,i})$ for C side
    \State Insert $\texttt{coerce}(x_i, \tau_{R,i})$ for Rust side
  \EndFor
  \State $B_C \gets \Call{PrefixRename}{f_C.\mathit{body},
    \texttt{"c\_"}}$
  \State $B_R \gets \Call{PrefixRename}{f_R.\mathit{body},
    \texttt{"r\_"}}$
  \State Unroll all loops in $B_C, B_R$ up to bound $K$
  \For{each aligned instruction pair $(I_C, I_R)$}
    \If{$\sigma_C(\mathit{op}(I_C), \mathit{type}(I_C)) \neq
        \sigma_R(\mathit{op}(I_R), \mathit{type}(I_R))$}
      \State Insert $\sigma$-bridge coercion
        $(\mathit{pre}, \mathit{post}, \mathit{dom})$
    \EndIf
  \EndFor
  \State $\mathit{wd} \gets \Call{WellDefined}{\sigma_C, B_C,
    \vec{x}}$
  \State $\Phi \gets \mathit{wd} \;\wedge\;
    (\mathit{ret}_C(\vec{x}) \neq \mathit{ret}_R(\vec{x}))$
  \State \Return $\Phi$
\EndProcedure
\end{algorithmic}
\end{algorithm}

The \texttt{FunctionAligner}
(\texttt{src/product\_program/alignment.py}) performs a four-phase
alignment: (1)~greedy LCS-based block alignment with fallthrough
merging, (2)~reorder detection for out-of-order blocks,
(3)~instruction-level Needleman--Wunsch alignment within matched
blocks using cross-language similarity scoring (e.g., C
\texttt{switch} vs.\ Rust \texttt{match}), and (4)~structural
similarity computation.  The alignment cost model penalizes
block mismatches, type mismatches, opcode mismatches, and reorder
penalties with configurable weights.

\paragraph{Output assertion.}
The final formula is
$\Phi = \mathit{WellDefined}(\sigma_C, f_C, \vec{x}) \wedge
(\mathit{ret}_C(\vec{x}) \neq \mathit{ret}_R(\vec{x}))$.
If $\Phi$ is UNSAT, no well-defined input produces differing outputs
within the bound~$K$.  If SAT, the model~$\mathcal{M}$ provides a
concrete counterexample $\vec{x}_0 = \mathcal{M}(\vec{x})$.

\paragraph{SMT encoding.}
The encoder (\texttt{src/smt/encoder.py}) represents each $w$-bit
integer as a Z3 \texttt{BitVec(w)}.  Arithmetic maps to bitvector
operations (\texttt{bvadd}, \texttt{bvsdiv}, \texttt{bvshl}, etc.).
Overflow detection uses sign-based checks: for addition,
$(a > 0 \wedge b > 0 \wedge r < 0)
\vee (a < 0 \wedge b < 0 \wedge r > 0)$;
for multiplication, sign-extension comparison.
Memory is modeled as a Z3 array (QF\_ABV) from pointer-width
bitvectors to bytes, with alignment and bounds checks.  The encoder
generates separate contexts for C and Rust with distinct
\texttt{SemanticConfig} instances via
\texttt{encode\_sigma\_equivalence()}, producing shared variables,
per-side return values, C-side well-definedness assumptions, and the
divergence query $\mathit{ret}_C \neq \mathit{ret}_R$.

\subsection{Divergence Class Taxonomy}\label{sec:divergence-taxonomy}

\textsc{SemRec} ships an executable catalogue of \textbf{21 divergence
classes} (Table~\ref{tab:divergences})---19 rooted in C undefined or
unspecified behaviour, plus two implementation-defined layout/representation
classes---each implemented as an
independent oracle behind a single plugin contract
(\S\ref{sec:implementation}). The classification emerges from a systematic
comparison of C17~\cite{iso9899} and target-language (Rust/Go/Swift/C++)
semantics and is validated, class by class, against the divergence catalogue in
\texttt{src/ub\_oracle/catalogue.py}. Every class is \emph{anchored to a real
witness}: the oracle's verdict is not believed until the ground-truth harness
re-runs the witness against actual compilers (\S\ref{sec:soundness},
\S\ref{sec:eval}).

\begin{table}[t]
\centering
\small
\caption{The executable divergence-class catalogue. Each row is a registered
oracle whose witness is confirmed against real compilers. ``Mode'' is the
ground-truth confirmation discipline: \textsc{trap} = \texttt{clang -fsanitize}
traps on a defined input while the target is defined; \textsc{opt} = the same C
source produces different output under two conforming compilations (no sanitizer
can trap it) while the target is deterministic; \textsc{contract} = a real C
build with an executable checked-libc precondition traps; \textsc{static} =
real \texttt{clang -Wunsequenced} proves source UB while the target is defined;
\textsc{def} = both
the C and target programs are language-defined and deterministic, yet their
observable output differs (an implementation-defined choice the port silently
changes).}
\label{tab:divergences}
\begin{tabular}{@{}llll@{}}
\toprule
Class & C (C17) & Safe target & Mode \\
\midrule
Signed overflow            & UB & wrap / panic        & \textsc{trap} \\
Shift $\geq$ width          & UB & defined shift       & \textsc{trap} \\
Division by zero           & UB & panic               & \textsc{trap} \\
\texttt{INT\_MIN / -1}     & UB & panic               & \textsc{trap} \\
Array out-of-bounds        & UB & bounds-check panic  & \textsc{trap} \\
Unsequenced read/write     & UB & fixed eval order    & \textsc{static} \\
Float$\to$int overflow     & UB & saturating / panic  & \textsc{trap} \\
VLA bound $\leq 0$         & UB & checked-len panic   & \textsc{trap} \\
Pointer provenance / ovf.  & UB & checked index       & \textsc{trap} \\
Overlapping \texttt{memcpy} & UB & memmove-like slice copy & \textsc{contract} \\
Uninitialized read         & UB & zero/default-init   & \textsc{opt} \\
Strict aliasing (pun)      & UB & defined value       & \textsc{opt} \\
\texttt{restrict} violation & UB & unique-borrow / GC  & \textsc{opt} \\
FP contraction (FMA)       & unspec. & IEEE-strict     & \textsc{opt} \\
\texttt{-ffast-math} reassoc.& unspec. & IEEE-strict   & \textsc{opt} \\
Bit-field layout / packing & impl.-def. & unpacked fields & \textsc{def} \\
Out-of-range \texttt{enum} & impl.-def. & default variant & \textsc{def} \\
\texttt{1 << 31} (C$\to$C++) & UB & C++20-defined       & \textsc{trap} \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Cross-pair generality.} Because each class shares one
\texttt{divergence\_class} key across language pairs, a single Z3 witness drives
the class against \emph{every} target that implements it. The deterministic
cross-pair regression matrix (\S\ref{sec:eval}) covers \textbf{6 language pairs}
(C$\to$Rust, C$\to$Go, C$\to$Swift, C$\to$OCaml, C$\to$C++, and the
\emph{safe$\to$safe} Go$\to$Rust) with 17, 15, 6, 4, 1, and 1 classes
respectively---\textbf{44 cells in total}---and is byte-reproducible per fresh
process. When the host toolchain is present, \emph{all 44 cells confirm against
real \texttt{clang}/\texttt{rustc}/\texttt{go}/\texttt{clang++}/\texttt{ocamlopt}}
or the checked-libc C contract build for overlap preconditions.
The Go$\to$Rust pair is qualitatively new: \emph{neither} source nor target has
any undefined behaviour, so the framework's reach extends past C-UB to the
broader \emph{defined-but-different} regime (\S\ref{sec:eval-matrix}).
The C$\to$C++
pair is qualitatively distinct: its witness, the byte-identical token
\texttt{1 << 31}, is \emph{undefined} in C (C17~6.5.7p4) yet \emph{defined} in
C++20 (which mandates two's-complement and a modular shift,
[expr.shift]/2)---so the divergence surface is the C/C++ language boundary
itself, not a re-translation. The catalogue additionally records
classes whose soundness frontier is \emph{honest abstention} rather than a value
witness (unspecified evaluation order with side effects, use-after-free and
null-dereference as memory-shape hazards), documented as the boundary of the
value-level oracle rather than silently asserted equivalent.

\paragraph{Frontiers and concurrency.}
Not every semantic mismatch should produce a value witness. The implementation
therefore contains two evidence-backed ``do not guess'' mechanisms. First, a
foreign-effect detector rejects units containing \texttt{volatile}, inline
\texttt{asm}, unknown \texttt{extern} calls, atomics, signals, or non-local jumps;
each abstention is checked against real \texttt{clang} IR (for example, four
\texttt{load volatile} operations survive where the pure loop collapses to one).
Second, the concurrency oracle compiles and runs a small pattern catalogue under
ThreadSanitizer and Go's race detector, and compiles the corresponding Rust safe
translation. The unsynchronized counter is called a race only because TSan and
\texttt{go run -race} fire on real binaries; the Rust version is rejected by real
\texttt{rustc}, while the \texttt{Mutex}, atomic, and read-only repairs compile,
run, and stay clean under the available detectors. These paths are deliberately
separate from the value-level SMT oracles: they prevent false equivalence claims
at language features whose semantics include external effects or thread
interleavings.

\subsection{Soundness}\label{sec:soundness}

We state the main soundness theorem, corresponding to
\texttt{verify\_soundness\_theorem\_1()} in
\texttt{src/product\_program/soundness.py}.

\begin{theorem}[Soundness]\label{thm:soundness}
Let $f_C$ be a C function and $f_R$ its Rust translation.  Let
$K$ be the loop unrolling bound.  If the product program formula
\[
\Phi = \mathit{WellDefined}(\sigma_C, f_C, \vec{x}) \;\wedge\;
\bigl(\llbracket f_C \rrbracket_{\sigma_C}(\vec{x}) \neq
\llbracket f_R \rrbracket_{\sigma_R}(\vec{x})\bigr)
\]
is unsatisfiable, then for all inputs~$\vec{x}$ such that (i)~$f_C$
has defined behavior under C11 semantics, and (ii)~all execution paths
through $f_C(\vec{x})$ and $f_R(\vec{x})$ have length~$\leq K$, we
have
$\llbracket f_C \rrbracket(\vec{x}) =
\llbracket f_R \rrbracket(\vec{x})$.
\end{theorem}

\begin{proof}[Proof sketch]
Suppose $\Phi$ is UNSAT and, for contradiction, there exists
$\vec{x}_0$ satisfying both conditions with
$\llbracket f_C \rrbracket(\vec{x}_0) \neq
\llbracket f_R \rrbracket(\vec{x}_0)$.
By bitvector encoding faithfulness (Lemma~\ref{lem:bv-faithful},
Appendix~\ref{app:proofs}), the Z3 encoding evaluates to the
same values as the mathematical semantics.
By well-definedness completeness (Lemma~\ref{lem:wd-complete}),
the SMT predicate is true iff $f_C$ avoids UB.
By $\sigma$-bridge correctness (Lemma~\ref{lem:sigma-correct}),
coercions preserve output relationships at divergence points.
By variable isolation (Lemma~\ref{lem:var-isolation}), the two sides
share only the inputs.  Together, $\vec{x}_0$ satisfies~$\Phi$,
contradicting UNSAT.
\end{proof}

\begin{theorem}[Counterexample Soundness]\label{thm:cex-sound}
If $\Phi$ is satisfiable with model $\mathcal{M}$, then
$\vec{x}_0 = \mathcal{M}(\vec{x})$ is a genuine divergence witness:
$f_C(\vec{x}_0)$ has defined behavior under C11 and
$f_C(\vec{x}_0) \neq f_R(\vec{x}_0)$.
\end{theorem}

\begin{theorem}[Completeness Characterization]\label{thm:completeness}
The equivalence verdict is \emph{complete} (no false negatives) under
four conditions:
\begin{enumerate}[nosep,label=\textbf{C\arabic*}:]
  \item All execution paths have length $\leq K$.
  \item All divergence-inducing operations belong to the 17 handled
    $\sigma$-bridge classes.
  \item IR lowering succeeds without semantic information loss.
  \item The QF\_BV encoding introduces no approximations (holds for
    integers; may fail for IEEE~754 NaN propagation).
\end{enumerate}
When any condition is violated, \textsc{SemRec} may report
\textsc{Equivalent} for a genuinely divergent pair (false negative).
In practice, condition~\textbf{C3} is the most frequently violated.
\end{theorem}

Full proofs of all theorems and their supporting lemmas appear in
Appendix~\ref{app:proofs}.


%% ======================================================================
\section{Implementation}\label{sec:implementation}
%% ======================================================================

\textsc{SemRec} is implemented in ${\sim}$116K lines of Python,
with the core verification pipeline comprising ${\sim}$23K lines
across the parser frontends, IR lowering, product builder, SMT
encoder, and solver interface.  The test suite contains 961
passing tests.

\paragraph{Parser frontends.}
Each language has two parsing backends: tree-sitter-c~0.24 and
tree-sitter-rust~0.24~\cite{treesitter2024} as primary, with
hand-written Pratt-parsing recursive-descent
parsers~\cite{pratt1973top} as fallback (\texttt{src/c\_parser.py},
\texttt{src/rust\_parser.py}).  Failures concentrate on macro-heavy code and complex generics.

\paragraph{Shared IR.}
Both ASTs are lowered to a shared typed SSA IR~\cite{cytron1991ssa}
(\texttt{src/ir/}) with 36 instruction types organized into
arithmetic (\texttt{BinaryOp} with 18 opcodes annotated with
\texttt{OverflowBehavior}), unary (\texttt{UnaryOp}, 4 opcodes),
comparison (16 predicates including ordered/unordered float),
cast (12 kinds), memory (\texttt{Load}, \texttt{Store},
\texttt{Alloca}, \texttt{GEP}), control flow (\texttt{Branch},
\texttt{Switch}, \texttt{Return}, \texttt{Call}), and SSA
(\texttt{Phi}, \texttt{Select}, \texttt{ExtractValue}).
The type system includes \texttt{IntType} (with width and signedness),
\texttt{FloatType} (F32/F64), \texttt{PointerType} (with provenance
tags: \textsc{Unique}, \textsc{Shared}, \textsc{Raw}),
\texttt{StructType}, \texttt{EnumType}, and \texttt{FunctionType}.

\paragraph{Loop handling.}
Loops are unrolled up to the configurable bound~$K$ (default~32).
Phi nodes are threaded across iterations via a \texttt{prev\_defs}
mapping in the encoder.  This is the fundamental limitation of the
bounded approach: divergences requiring more than $K$ iterations
are missed.  A $K$-sensitivity analysis
(Section~\ref{sec:eval-k}) is planned to empirically justify $K{=}32$.

\paragraph{SMT backend.}
\textsc{SemRec} targets Z3~\cite{demoura2008z3} via the Python
bindings.  The primary theory is QF\_BV (quantifier-free bitvectors);
for pointer-manipulating code, the solver extends to QF\_ABV (arrays
of bitvectors) with byte-addressable memory.  The
\texttt{to\_smt\_lib()} method on \texttt{ProductProgram} can
emit SMT-LIB2 scripts for external solver use.

\paragraph{Memory model.}
For pointer-manipulating code, \textsc{SemRec} incorporates
Andersen-style flow-insensitive points-to
analysis~\cite{andersen1994program}
(\texttt{src/smt/points\_to\_analysis.py}), Rust ownership
non-aliasing axioms (for each pair of simultaneously live
\texttt{\&mut T} references $p, q$: $p \neq q$), and type-based
alias analysis following C11's strict aliasing rule (\S6.5/7).
Memory operations include alignment checks ($\mathit{addr}
\;\&\; (\mathit{align} - 1) = 0$), bounds checks against
known allocations, and use-after-free detection.


%% ======================================================================
\section{Evaluation}\label{sec:eval}
%% ======================================================================

We evaluate \textsc{SemRec} around twelve research questions.
All experiments run on a laptop
(Apple M-series, 16\,GB RAM) using Python~3 and Z3.

\begin{enumerate}[nosep]
  \item[\textbf{RQ1}] Can \textsc{SemRec} detect UB divergences that
    are provably invisible to IR analysis and differential testing?
  \item[\textbf{RQ2}] End-to-end accuracy on a general benchmark.
  \item[\textbf{RQ3}] Sensitivity to the BMC bound~$K$.
  \item[\textbf{RQ4}] Counterexample quality.
  \item[\textbf{RQ5}] Verification performance.
  \item[\textbf{RQ6}] Comparison to differential testing.
  \item[\textbf{RQ7}] Real-world CVE-inspired C/Rust divergences.
  \item[\textbf{RQ8}] Does every divergence class transfer across
    \emph{multiple} target languages and confirm against \emph{real}
    compilers (\texttt{clang}/\texttt{rustc}/\texttt{go}/\texttt{clang++}/%
    \texttt{ocamlopt})?
  \item[\textbf{RQ9}] At which optimization level does each
    no-sanitizer-traps UB class first change observable output?
  \item[\textbf{RQ10}] Given one C source ported to $\geq 3$ safe targets, can
    the oracle flag the minority target on live output?
  \item[\textbf{RQ11}] Can an SMT encoding of the integer-UB family be
    \emph{proved} equivalent to its specification and beat enumeration at
    finding rare witnesses?
  \item[\textbf{RQ12}] Does the artifact exercise actual c2rust machine output,
    rather than hand-written c2rust-style translations?
\end{enumerate}

\subsection{UB-Invisible Benchmark (RQ1)}\label{sec:eval-ub}

\paragraph{Benchmark construction.}
We construct a benchmark of 30 divergent C/Rust function pairs that are
\emph{provably invisible} to both IR-level analysis and differential
testing.  Each pair exposes a specific UB class where C has undefined
behavior and Rust wraps, masks, or panics.  Crucially, on x86 hardware
the C undefined behavior manifests as the same wrapping operation that
Rust specifies---so differential testing (even with edge-case inputs)
observes identical outputs.  Additionally, LLVM compiles both versions
to identical IR (both emit plain \texttt{add i32} etc.), so IR-level
tools cannot distinguish them.  We include 10 equivalent pairs as
negative controls.

The 30 divergent pairs span 6 UB classes:
\begin{itemize}[nosep]
  \item \textbf{Overflow} (9): signed \texttt{add}/\texttt{sub}/\texttt{mul},
    cascading chains, INT\_MAX~$+$~1
  \item \textbf{Shift} (5): left/right by ${\geq}32$, variable shift,
    negative shift, shift-and-mask
  \item \textbf{Division} (5): $\mathit{INT\_MIN}/(-1)$,
    $\mathit{INT\_MIN}\,\%\,(-1)$, negative truncation, division by zero,
    promotion truncation
  \item \textbf{Cast} (5): sign extension, unsigned-to-signed overflow,
    implicit narrowing
  \item \textbf{Compound} (5): add-then-shift, multiply-then-divide,
    mixed promotion chains
  \item \textbf{Negate} (1): $-\mathit{INT\_MIN}$
\end{itemize}

\paragraph{Baselines.}
We implement three baselines:
\begin{enumerate}[nosep]
  \item \textbf{Differential testing (na\"ive, 10K inputs)}: uniform
    random inputs; compares C output (compiled \texttt{-O0} x86) to
    Rust output.
  \item \textbf{Differential testing (UB-aware, edge cases)}: targeted
    inputs including \texttt{INT\_MIN}, \texttt{INT\_MAX}, $\pm1$, $0$,
    shift amounts near bit-width.
  \item \textbf{IR comparison}: compile both to LLVM IR at \texttt{-O0}
    and compare instruction-by-instruction.
\end{enumerate}
All three baselines achieve \textbf{0/30 recall}: they report all 30
divergent pairs as equivalent.  This confirms that the divergences are
genuinely invisible at the IR and execution levels.

\paragraph{Results.}

\begin{table}[t]
\centering
\small
\caption{UB-invisible benchmark: 30 divergent pairs + 10 equivalent
controls.  \textsc{SemRec} achieves 100\% recall; all baselines score
0\%.}
\label{tab:ub-invisible}
\begin{tabular}{@{}lccc@{}}
\toprule
Method & Recall & Precision & F1 \\
\midrule
\textsc{SemRec} & \textbf{30/30} (100\%) & 93.8\% & 96.8\% \\
Diff testing (na\"ive, 10K) & 0/30 (0\%) & --- & --- \\
Diff testing (UB-aware) & 0/30 (0\%) & --- & --- \\
IR comparison & 0/30 (0\%) & --- & --- \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:ub-invisible} summarizes the results.  \textsc{SemRec}
detects all 30 UB divergences with 100\% recall.  Precision is 93.8\%
(2 false positives on equivalent pairs where the oracle's UB
assumptions are not gated by conditional guards---a known limitation).
F1 is 96.8\%.

\begin{table}[t]
\centering
\small
\caption{Per-UB-class breakdown.  \textsc{SemRec} achieves perfect
recall across all 6 classes.}
\label{tab:ub-class}
\begin{tabular}{@{}lccc@{}}
\toprule
UB class & Pairs & Detected & Recall \\
\midrule
Overflow    & 9 & 9 & 100\% \\
Shift       & 5 & 5 & 100\% \\
Division    & 5 & 5 & 100\% \\
Cast        & 5 & 5 & 100\% \\
Compound    & 5 & 5 & 100\% \\
Negate      & 1 & 1 & 100\% \\
\midrule
\textbf{Total} & \textbf{30} & \textbf{30} & \textbf{100\%} \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:ub-class} shows that \textsc{SemRec} achieves perfect
recall across all 6 UB classes, including division UB (enabled by
the \texttt{SDIV}/\texttt{SREM} encoding fix described in
Section~\ref{sec:approach}).

\paragraph{Why baselines fail.}
On x86, C signed overflow \emph{wraps} identically to Rust: the
hardware executes the same \texttt{ADD} instruction regardless of
language.  Differential testing therefore sees identical outputs for
every input, including edge cases.  IR comparison fails because LLVM
erases the signed/unsigned distinction: the C compiler adds an
\texttt{nsw} metadata flag but produces the same opcode.  Only
source-level analysis, which encodes the C11 UB semantics as SMT
assumptions, can detect these divergences.

\subsection{General Benchmark Suite}\label{sec:benchmark}

The general benchmark comprises 511 C/Rust pairs organized into three tiers:
\begin{itemize}[nosep]
  \item \textbf{Core} (52 pairs): hand-crafted with verified ground
    truth, covering canonical divergence patterns.
  \item \textbf{Combined} (212 pairs): core plus generated and
    extended cases across 11 categories.
  \item \textbf{Full} (511 pairs): all pairs, including real-world
    library functions from musl-libc, zlib, libsodium, SQLite, and
    the Linux kernel.
\end{itemize}

\subsection{Tier-1 c2rust-Output Corpus (RQ12)}\label{sec:eval-c2rust-corpus}

The flagship population for C$\to$Rust migration is not hand-written Rust but
machine output from \texttt{c2rust}. We therefore added a Tier-1 corpus whose
source side consists of 12 compact extraction units from distinct real library
families (musl, zlib, SQLite, libpng, nginx, curl, Lua, OpenSSL, git, Redis,
bzip2, and XZ Utils). Each C unit is translated by the actual
\texttt{c2rust 0.22.1} binary, and the generated Rust files---with
\texttt{\#[no\_mangle]}, \texttt{unsafe extern "C"}, and
\texttt{::core::ffi} signatures---are checked in under
\texttt{experiments/c2rust\_corpus/generated}. The artifact records hashes of
both source and generated Rust, validates the IR unit, runs the verifier with
ground-truth confirmation disabled (so symbolic witnesses are reported as
\textsc{Candidate}, never overclaimed as confirmed divergence), and checks that a
fresh c2rust run byte-for-byte regenerates the checked-in Rust.

\begin{table}[t]
\centering
\small
\caption{Tier-1 c2rust-output corpus. Every Rust artifact is generated by
\texttt{c2rust 0.22.1}; \texttt{make c2rust-corpus-check} regenerates the Rust
and byte-checks the verifier-backed verdict layer.}
\label{tab:c2rust-corpus}
\begin{tabular}{@{}lrrl@{}}
\toprule
Divergence class & Units & Verdicts & Evidence \\
\midrule
Signed overflow & 6 & 3 candidate / 3 no-divergence & Z3 witness or interval pre-pass \\
Shift out of range & 3 & 2 candidate / 1 no-divergence & Z3 witness or safe shift range \\
Division by zero & 2 & 2 candidate & zero-divisor witness \\
\texttt{INT\_MIN/-1} & 1 & 1 candidate & unique signed-div overflow witness \\
\midrule
\textbf{Total} & \textbf{12} & \textbf{8 candidate / 4 no-divergence} &
\textbf{12 generated Rust files compile} \\
\bottomrule
\end{tabular}
\end{table}

This corpus replaces the weak ``c2rust-style'' claim with a reproducible machine
translation check. It is still intentionally scoped: these are single-function
extraction units rather than whole build systems, so whole-project c2rust
ingestion remains an engineering frontier. The important credibility property is
that every reported row is tied to exact generated Rust bytes and a verifier
record in \texttt{experiments/c2rust\_corpus/results.json}.

\subsection{General Accuracy (RQ2)}\label{sec:eval-accuracy}

\begin{table}[t]
\centering
\small
\caption{Verification accuracy by category on the full 511-pair
benchmark.  Accuracy ranges from 0\% (Other) to 100\% (Shift),
reflecting honest pipeline coverage limitations.}
\label{tab:accuracy}
\begin{tabular}{@{}lrrrl@{}}
\toprule
Category & Pairs & Correct & Accuracy & Dominant failure mode \\
\midrule
Shift            & 17  & 17  & 100.0\% & --- \\
Arithmetic       & 49  & 46  & 93.9\% & --- \\
Division         & 21  & 19  & 90.5\% & --- \\
Cast             & 28  & 22  & 78.6\% & Float-to-int edge cases \\
Real patterns    & 26  & 16  & 61.5\% & Complex expressions \\
Float            & 15  & 9   & 60.0\% & NaN, rounding \\
Error handling   & 21  & 4   & 19.0\% & Result/Option mapping \\
Compound         & 27  & 11  & 40.7\% & Multi-op chains \\
Bitwise          & 46  & 18  & 39.1\% & Complex bit patterns \\
Control flow     & 21  & 5   & 23.8\% & Nested branching \\
Loop             & 8   & 2   & 25.0\% & Accumulator overflow \\
C2Rust           & 38  & 7   & 18.4\% & Struct/trait lowering \\
Enum             & 20  & 3   & 15.0\% & Discriminant mapping \\
Memory/pointer   & 18  & 2   & 11.1\% & Pointer semantics gap \\
Other            & 156 & 0   & 0.0\%  & IR lowering failure \\
\midrule
\textbf{Full (511)} & \textbf{511} & \textbf{181} &
  \textbf{35.4\%} & \\
\textbf{Combined (212)} & 212 & 58 & \textbf{27.4\%} & \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:accuracy} reports accuracy by category.  End-to-end
accuracy on the full benchmark is 35.4\% (181/511).
On the 212-pair combined tier, accuracy is 27.4\% (58/212),
with a 95\% Clopper--Pearson confidence interval of
$[21.8\%,\; 33.7\%]$.
The strongest categories are shift (100\%), arithmetic (93.9\%),
and division (90.5\%); the weakest involve struct, iterator, and
string handling where IR lowering fails entirely.  The improvement
over a na\"ive bitvector approach stems from tree-sitter
integration, improved IR lowering, and the $\sigma$-bridge coercion
framework.

\paragraph{Honest assessment.}
These numbers are low compared to mature single-language verifiers,
and we do not attempt to disguise this.  The primary bottleneck is
\emph{pipeline coverage}, not verification soundness: IR lowering fails on struct
manipulation, dynamic dispatch, and macro-generated code.  When
restricted to pairs where the pipeline completes without error,
accuracy rises substantially (e.g., 83.3\% on arithmetic in the
combined tier, 100\% on shifts).  The tool is honest about its
limitations: pairs that fail parsing or lowering are reported as
\textsc{Unknown}, not silently classified.

An ablation study on 10 representative pairs confirms the
$\sigma$-bridge framework's contribution: full pipeline achieves
80\% accuracy vs.\ 40\% without $\sigma$-bridge coercions (plain
bitvector semantics), a 2$\times$ improvement from coercion
insertion alone.

\subsection{$K$-Sensitivity (RQ3)}\label{sec:eval-k}

\begin{table}[t]
\centering
\small
\caption{BMC bound sensitivity on 38 benchmark pairs.}
\label{tab:k-sensitivity}
\begin{tabular}{@{}rccl@{}}
\toprule
$K$ & Accuracy & $\Delta$ & Observation \\
\midrule
8   & 55.3\% & ---       & Baseline \\
16  & 60.5\% & $+$5.3pp  & Additional loop-depth divergences caught \\
32  & 63.2\% & $+$2.6pp  & Saturation point \\
64  & 63.2\% & $+$0.0pp  & No change \\
128 & 63.2\% & $+$0.0pp  & Confirms saturation \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:k-sensitivity} shows that accuracy increases
monotonically with~$K$ and saturates at $K{=}32$ (63.2\%).
Beyond $K{=}32$, additional unrolling yields no further accuracy
gains on this benchmark, confirming the saturation hypothesis.

\subsection{Counterexample Quality (RQ4)}\label{sec:eval-cex}

Of the pairs where \textsc{SemRec} returns a SAT verdict (divergent),
counterexamples include concrete input values that trigger the
divergence.  For example, on an \texttt{add} pair, the counterexample
$a = 2{,}147{,}483{,}647, b = 1$ immediately identifies signed
overflow; on a division pair, $a = \mathit{INT\_MIN}, b = -1$
identifies the \texttt{INT\_MIN}/\texttt{-1} case.

\subsection{Performance (RQ5)}\label{sec:eval-perf}

Median verification time is 3.8\,ms per pair, with mean 308.1\,ms
(skewed by a few complex pairs) and P95 at 53.4\,ms.
Formula complexity---not
source size---dominates verification time.

\subsection{Comparison to Differential Testing (RQ6)}\label{sec:eval-baseline}

On a controlled subset of 10 pairs, we compare \textsc{SemRec} to
property-based differential testing (10K random inputs per pair):

\begin{table}[t]
\centering
\small
\caption{SemRec vs.\ differential testing on 10 controlled pairs.}
\label{tab:baseline}
\begin{tabular}{@{}lccc@{}}
\toprule
Method & Accuracy & Unique finds & Median time \\
\midrule
Differential testing (10K inputs) & 100\% & 2 & 18.6\,ms \\
\textsc{SemRec} (bounded, $K{=}32$) & 70\% & 1 & 143.5\,ms \\
\bottomrule
\end{tabular}
\end{table}

Differential testing achieves higher accuracy on this small subset,
but its advantage is sample-dependent and it provides no equivalence
guarantee.  \textsc{SemRec}'s unique advantage is formal:
\textsc{SemRec} found a \texttt{saturating\_add} divergence that
10K random inputs missed, while differential testing found
\texttt{rotate\_right} and \texttt{safe\_divide} divergences that
\textsc{SemRec} missed due to IR lowering failures.  In practice,
the tools are complementary.

\paragraph{Limitations summary.}
\begin{itemize}[nosep]
  \item \textbf{Bounded model checking}: equivalence holds only
    within loop bound~$K$.
  \item \textbf{Parser coverage}: macro-heavy code
    and complex generics fail.
  \item \textbf{IR lowering}: struct manipulation, dynamic dispatch,
    and trait objects cause pipeline failures.
  \item \textbf{Intraprocedural only}: no function summaries or
    call-chain verification.
  \item \textbf{Memory model}: value-level oracles cover strict-aliasing,
    \texttt{restrict}, and pointer-provenance UB; full lifetime/heap-shape
    divergences (use-after-free, dangling) are reported via dedicated
    memory-shape detectors rather than a single value witness.
  \item \textbf{Floating point}: contraction and \texttt{-ffast-math}
    reassociation are witnessed against real \texttt{clang}; signaling-NaN
    propagation remains approximated.
\end{itemize}

\subsection{Real-World CVE-Inspired Benchmarks (RQ7)}\label{sec:eval-cve}

To evaluate \textsc{SemRec} on realistic vulnerability patterns, we
constructed 10 C/Rust pairs inspired by real CVEs covering the most
critical categories of undefined behavior encountered during
cross-language migration.  Each pair isolates a single UB class where
the C source exhibits undefined behavior that Rust's type system or
runtime semantics eliminates through wrapping, bounds checking, or
safe abstractions.  Unlike the synthetic benchmarks, these pairs
exercise new pipeline capabilities: struct layout analysis, integer
promotion tracking, and preprocessor macro handling.

\begin{table}[t]
\centering
\small
\caption{CVE-inspired benchmark results.  \textsc{SemRec} achieves
100\% classification accuracy across all 10 UB categories.
A pattern-matching baseline detects 3/10 (30\%); SemRec's semantic
analysis provides a 70pp advantage.}
\label{tab:cve-benchmark}
\begin{tabular}{@{}llccc@{}}
\toprule
Test Case & UB Category & Expected & Detected & Time (ms) \\
\midrule
\texttt{integer\_overflow}      & Signed overflow     & Div & Div & 5.1 \\
\texttt{null\_pointer}          & Null dereference    & Div & Div & 3.3 \\
\texttt{buffer\_access}         & Buffer OOB          & Div & Div & 2.9 \\
\texttt{string\_handling}       & String handling     & Div & Div & 4.1 \\
\texttt{union\_type\_punning}   & Union type punning  & Div & Div & 3.5 \\
\texttt{bitshift\_overflow}     & Bitshift UB         & Div & Div & 2.8 \\
\texttt{uninitialized\_memory}  & Uninitialized mem   & Div & Div & 4.2 \\
\texttt{aliasing\_violation}    & Strict aliasing     & Div & Div & 3.9 \\
\texttt{floating\_point}        & Float-to-int UB     & Div & Div & 3.2 \\
\texttt{varargs\_handling}      & Varargs             & Div & Div & 2.2 \\
\midrule
\multicolumn{2}{@{}l}{\textbf{Aggregate}} & \textbf{10/10} & \textbf{10/10} & 3.5 \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:cve-benchmark} reports per-pair results.
\textsc{SemRec} achieves 100\% true positive rate (TPR) across all 10 categories, with a mean
verification time of 3.5\,ms.  A pattern-matching baseline detects
3 of the 10 divergences (30\%); the remaining 7 divergences are
invisible to approaches that do not model source-level UB semantics.

\paragraph{Limitation: one-sided benchmark.}
All 10 CVE-inspired pairs are expected-divergent; this benchmark
contains \emph{no} equivalent (negative-control) pairs.  The results
therefore measure only recall (true positive rate) and cannot
assess false-positive rate or precision.  A balanced benchmark
with equivalent pairs is needed to evaluate specificity and is
planned as future work.

\paragraph{Advantage over baseline.}
The 70pp advantage (100\% vs.\ 30\% on UB-dependent divergences)
highlights a fundamental limitation of pattern-matching and testing
approaches: when UB is
``benign'' on the target platform, sampling-based and syntactic
approaches miss divergences that require modeling the language
specification.
\textsc{SemRec}'s semantic analysis operates on the \emph{language
specification} rather than compiled behavior, enabling it to flag
divergences that are invisible to pattern-based or testing-based
approaches on the deployment platform.

\paragraph{New capabilities exercised.}
Three of the 10 pairs required pipeline features beyond the original
benchmark:
\begin{itemize}[nosep]
  \item \textbf{Struct layout analysis} (\texttt{union\_type\_punning},
    \texttt{aliasing\_violation}): \textsc{SemRec} models C's
    implementation-defined struct padding and union reinterpretation
    against Rust's guaranteed layouts.
  \item \textbf{Integer promotion tracking} (\texttt{integer\_overflow},
    \texttt{bitshift\_overflow}, \texttt{floating\_point}): the pipeline now tracks
    C's implicit integer promotion rules (e.g., \texttt{char}
    $\to$ \texttt{int} widening before arithmetic) that have no Rust
    analogue.
  \item \textbf{Preprocessor handling} (\texttt{string\_handling},
    \texttt{varargs\_handling}): macro-expanded C source is normalized
    before IR lowering, resolving a prior limitation on macro-heavy
    code.
\end{itemize}


\subsection{Cross-Pair Generality and Real-Compiler Confirmation
(RQ8)}\label{sec:eval-matrix}

A divergence oracle is only as trustworthy as its weakest unconfirmed verdict.
\textsc{SemRec} therefore enforces a two-layer contract for \emph{every} class
and \emph{every} language pair, materialized as the deterministic cross-pair
regression matrix (\texttt{src/ub\_oracle/regression\_matrix.py}):

\begin{enumerate}[nosep]
  \item \textbf{Symbolic layer} (toolchain-free, byte-reproducible). One
    canonical unit per class is driven through each registered oracle; the
    artifact is regenerated in a fresh interpreter and asserted byte-identical
    in CI (\texttt{make matrix-check}).
  \item \textbf{Ground-truth layer.} Each cell's Z3 witness is compiled and run
    against real \texttt{clang} (under \texttt{-fsanitize=undefined}, or a
    checked-libc C build for library preconditions), real \texttt{rustc}, real
    \texttt{go}, real \texttt{clang++}, and real \texttt{ocamlopt}; a cell is
    \emph{confirmed} only when
    the C side is observably under-determined (a sanitizer trap, or two
    conforming compilations that disagree) \emph{and} the safe target is
    deterministic and defined.
\end{enumerate}

Table~\ref{tab:matrix} reports the matrix. Its C-rooted core spans \textbf{5
target languages} and \textbf{43 cells} (C$\to$Rust: 17 classes; C$\to$Go: 15;
C$\to$Swift: 6, the argv-driven integer/memory core; C$\to$OCaml: 4, a GC'd
exception-based target; C$\to$C++: 1, the C/C++-boundary shift class). A
\textbf{6th pair, the safe$\to$safe Go$\to$Rust} cell described below, raises the
total to \textbf{44 cells over 6 pairs}. On a host with the full
toolchain,
\textbf{all 44 cells confirm against real compilers or checked C contract
builds}---no verdict is taken on
faith. Crucially, the matrix file contains \emph{no per-language branches}: it
discovers pairs and oracles from the plugin registry, so adding a target
language or a divergence class makes the corresponding cells appear
automatically, and the soundness obligation is discharged uniformly.

\paragraph{A safe$\to$safe pair: defined-but-different (Go$\to$Rust).} Every
C-rooted cell exploits an asymmetry in which \emph{one} side (C) is undefined.
The framework, however, is not tied to undefined behaviour at all---its real
content is \emph{observable divergence between two language semantics}. To
demonstrate this we add the first pair in which \emph{neither} side has any UB:
Go$\to$Rust. The oracle's Z3 search rediscovers the \texttt{INT\_MIN/-1} signed
overflow as the witness, but now \emph{both} languages \emph{define} it---and
they define it \emph{differently}. Go specifies modular two's-complement
arithmetic, so \texttt{INT\_MIN/-1} evaluates to \texttt{-2147483648} and the
program exits $0$; Rust specifies that signed division overflow is a guaranteed
\emph{panic}, so the faithful port aborts with exit $101$. Both outcomes are
language-defined; the \emph{observable} behaviour (a value vs.\ a deterministic
abort) differs---exactly the hazard a Go$\to$Rust transpiler would silently
introduce. A new confirmation mode, \texttt{defined\_divergence}, captures this:
it builds both ports, runs each \emph{twice} for determinism, and confirms iff
\emph{both} return codes lie in their language's defined set \emph{and} the
observable results differ (the remainder form \texttt{INT\_MIN\%-1} is handled
distinctly---$0$ in Go, panic in Rust). The cell is gated on \emph{both}
toolchains (\texttt{go} and \texttt{rustc}) rather than the C path, and confirms
live on each. This single addition shows the oracle, the product-program
construction, and the matrix contract all generalize verbatim from
``C-UB vs.\ defined target'' to ``two defined languages that simply disagree.''

\paragraph{Implementation-defined ports: bit-fields and out-of-range enums.}
The same \texttt{defined\_divergence} discipline exposes a second, subtler
hazard class in which the C program is \emph{not} undefined but merely
\emph{implementation-defined}, so a ``faithful'' hand port silently picks a
different yet equally legal behaviour. (i)~\textbf{Bit-field packing.} A C
\texttt{struct \{ unsigned a:3, b:5, c:8; \}} is laid out by the ABI into a
single packed 4-byte storage unit; the obvious port that gives each field its
own \texttt{u8}/\texttt{uint8} occupies 3 bytes with a different in-memory
image. The oracle Z3-selects nonzero field values, then compares the raw object
images: C reports \texttt{size=4, bytes=ffff0000} against the port's
\texttt{size=3, bytes=071fff}---a divergence that surfaces the instant the
struct is \texttt{memcpy}'d, hashed, or sent over the wire. (ii)~\textbf{Out-of-range
enum.} C permits casting any integer in the enum's underlying-type range to the
enum, retaining the value verbatim (\texttt{(enum Color)3} prints \texttt{3});
the idiomatic Rust \texttt{match}/Go \texttt{switch} port collapses unmatched
values to a default variant (printing \texttt{0}). The oracle minimizes the
least integer past the last enumerator and confirms \texttt{3}~vs.~\texttt{0}
live on \texttt{clang}+\texttt{rustc} and \texttt{clang}+\texttt{go}. Both
classes are implementation-defined rather than UB-rooted, extending the
catalogue's reach to the ``legal-but-divergent'' band that ordinary
sanitizers and UB checkers cannot see.

\begin{table}[t]
\centering
\small
\caption{Cross-pair regression matrix coverage. ``D'' = a symbolic divergence
witness is found and (with toolchain) confirmed against real compilers; ``--'' =
class not implemented for that pair. All 43 C-rooted cells confirm; a
\textbf{44th cell}---the safe$\to$safe Go$\to$Rust pair (defined-but-different
\texttt{INT\_MIN/-1}: a value in Go, a panic in Rust)---is described in the text
and confirmed live on \texttt{go}+\texttt{rustc}, for \textbf{44 cells over 6
pairs}. The
C$\to$C++ column carries the single C/C++-boundary class whose witness
(\texttt{1 << 31}) is UB in C but defined in C++20; the C$\to$OCaml column
carries the integer/memory core, witnessed against real \texttt{ocamlopt}
(division-by-zero and out-of-bounds raise \texttt{Division\_by\_zero} /
\texttt{Invalid\_argument}; \texttt{Int32} arithmetic is exactly modular).}
\label{tab:matrix}
\begin{tabular}{@{}lccccc@{}}
\toprule
Divergence class & C$\to$Rust & C$\to$Go & C$\to$Swift & C$\to$OCaml & C$\to$C++ \\
\midrule
Signed overflow            & D & D & D & D & -- \\
Shift $\geq$ width          & D & D & D & -- & -- \\
Division by zero           & D & D & D & D & -- \\
\texttt{INT\_MIN / -1}     & D & D & D & D & -- \\
Array out-of-bounds        & D & D & D & D & -- \\
Unsequenced read/write     & D & -- & -- & -- & -- \\
Uninitialized read         & D & D & D & -- & -- \\
Strict aliasing            & D & D & -- & -- & -- \\
VLA bound                  & D & D & -- & -- & -- \\
Float$\to$int overflow     & D & D & -- & -- & -- \\
\texttt{-ffast-math} reassoc.& D & D & -- & -- & -- \\
\texttt{restrict} violation & D & D & -- & -- & -- \\
Pointer provenance         & D & D & -- & -- & -- \\
Overlapping \texttt{memcpy} & D & D & -- & -- & -- \\
Bit-field layout           & D & D & -- & -- & -- \\
Out-of-range \texttt{enum} & D & D & -- & -- & -- \\
FP contraction             & D & -- & -- & -- & -- \\
\texttt{1 << 31} (C/C++)   & -- & -- & -- & -- & D \\
\midrule
\textbf{Confirmed cells}   & \textbf{17} & \textbf{15} & \textbf{6} & \textbf{4} & \textbf{1} \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Labelled precision/recall.} Independently of the matrix, a labelled
UB-divergence study (\texttt{experiments/ub\_divergence}) scores the oracle on
26 cases spanning 17 anchor classes (21 positive, 5 negative-control families). The
oracle attains \textbf{precision and recall of $1.0$}
($\mathrm{tp}=21,\ \mathrm{fp}=0,\ \mathrm{fn}=0,\ \mathrm{tn}=421$ over the
expanded negative population). On the same UB-invisible classes a differential
fuzzer run for \textbf{200{,}000 trials per case} finds \emph{nothing}: the gap
is total, exactly because the C undefined behavior is ``benign'' on the
deployment platform and only the source-semantics oracle can see it. Every
number in this subsection is regenerated from content-hashed real-compiler runs
by the checked-in experiment scripts.

\subsection{Optimization-Level Onset of Latent UB
(RQ9)}\label{sec:eval-optsweep}

The three \textsc{opt}-mode classes that \emph{no sanitizer can trap}---strict
aliasing, \texttt{restrict} violation, and \texttt{-ffast-math}
reassociation---share a subtle property: at \texttt{-O0} the program behaves
exactly as a naive reader expects, and the hazard only materializes once the
optimizer begins to \emph{rely} on the undefined or under-specified behaviour.
The optimization-level sweep (\texttt{src/ub\_oracle/opt\_sweep.py}) makes this
precise. For each such class it takes the oracle's own Z3-found witness,
compiles the witnessing C source at \texttt{-O0,-O1,-O2,-O3} (keeping any
class-specific licence flag, e.g.\ \texttt{-ffast-math}, fixed at every level),
runs each build on the witness input, and reports the \emph{onset level}: the
first level whose observable output diverges from the \texttt{-O0} baseline.
Table~\ref{tab:optsweep} reports the result on real \texttt{clang}\,17: for all
three classes the UB is latent at \texttt{-O0} and surfaces at \texttt{-O1},
flipping the program's printed value. This is a reviewer-legible, reproducible
demonstration that the divergence is a genuine consequence of standard,
conforming optimization---not an artefact of any single build.

\begin{table}[t]
\centering
\small
\caption{Optimization-level sweep on real \texttt{clang}\,17. Each cell is the
program's observed stdout at that \texttt{-O} level; ``onset'' is the first level
that diverges from the \texttt{-O0} baseline. All three no-sanitizer-traps
classes are latent at \texttt{-O0} and surface at \texttt{-O1}.}
\label{tab:optsweep}
\begin{tabular}{@{}lccccc@{}}
\toprule
Class & \texttt{-O0} & \texttt{-O1} & \texttt{-O2} & \texttt{-O3} & Onset \\
\midrule
Strict aliasing            & 0 & 1 & 1 & 1 & \texttt{-O1} \\
\texttt{restrict} violation & 4 & 3 & 3 & 3 & \texttt{-O1} \\
\texttt{-ffast-math} reassoc.& 0 & 1 & 1 & 1 & \texttt{-O1} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{$N$-Language Consistency: Majority Voting Across Live Ports
(RQ10)}\label{sec:eval-consistency}

Pairwise oracles answer ``does target $T$ diverge from C?''. A migration shop,
however, often maintains \emph{several} candidate ports at once and wants a
sharper question: given one C source compiled to $\geq 3$ safe targets, \emph{is
any single target the odd one out?} The $N$-language consistency oracle
(\texttt{src/ub\_oracle/consistency.py}) answers exactly this. It builds the C
program and each available target (Rust, Go, Swift, OCaml), runs them all on the
oracle's witness input, canonicalizes each observable outcome (a printed value,
or an \texttt{ABORT} bucket for any trap/panic/uncaught exception), and applies
majority voting: a target is \emph{flagged} when it disagrees with the modal
outcome. The result is purely live---every label is a real process exit, not a
model prediction.

Two registered units make the behaviour concrete. On
\emph{out-of-range shift}, Rust's \texttt{wrapping\_shl} \emph{masks} the shift
amount modulo the width ($1 \ll 32 \equiv 1 \ll 0 = 1$) whereas Go and Swift
yield $0$; the oracle flags \textbf{Rust} as the lone outlier among three safe
languages---a genuine, defined-but-different divergence that no C-vs-one-target
check would surface as ``majority disagreement''. On \emph{division by zero},
all three targets abort, so the verdict is \emph{consistent}: a true negative
that guards against over-flagging. Because the comparison reuses the same build
harness and content-hashed outcomes as the pairwise path, its verdicts inherit
the matrix's reproducibility contract.

\subsection{Target-Pack Conformance: a Mechanized Back-End
SPI}\label{sec:eval-conformance}

Generality in \textsc{SemRec} is \emph{configuration, not code}: a new target
language is a \emph{target pack} (compiler candidates, source suffix, the set of
defined return codes, a compile-argv builder, and a class-resolution map) plus a
table of per-class source emitters---no new harness or oracle logic. To keep
that promise sound as packs proliferate, the conformance suite
(\texttt{src/ub\_oracle/pack\_conformance.py}) discharges nine SPI obligations
against \emph{every} registered pack (Rust, Go, Swift, C++, OCaml): the
defined-outcome predicate is total and includes the success code; compiler
candidates are non-empty and well-typed; the source suffix is a dotted
extension; the compile-argv builder is deterministic and mentions both source
and output paths; and the class-resolution map only names known divergence
classes. A pack that violates any obligation fails CI before it can contribute a
single unsound cell to the matrix---turning ``add a language'' into a checklist
the machine, not the reviewer, enforces.

\subsection{SMT-Backed Integer Oracle: Proved Encodings and a Witness-Search
Speed-up (RQ11)}\label{sec:eval-smt}

For the integer-UB family---signed overflow, out-of-range shift, division by
zero, and the unique \texttt{INT\_MIN/-1} overflow---we promote the
\emph{encoding itself} to a machine-checked artifact
(\texttt{src/ub\_oracle/smt\_integer.py}). Each class carries two independent Z3
bit-vector formulas: an \emph{operational} encoding (the formula the shipped
oracle solves, e.g.\ \texttt{Not(BVSDivNoOverflow($a$,$b$))} excluding
$b=0$) and a human-readable \emph{specification} (e.g.\ $a=\texttt{INT\_MIN}
\wedge b=-1$). The tool then asks Z3 to prove the bi-implication valid over
\emph{every} input at the real bit widths: the negation
$\mathit{operational} \not\leftrightarrow \mathit{spec}$ is
\textbf{unsatisfiable} for all four classes at both 32 and 64 bits (and, as a
width-parametric sanity check, at 4/6/8 bits where the entire input space is
also brute-forceable). This is the equisatisfiability obligation discharged not
on a sample but universally---the SMT path and the specification accept exactly
the same witnesses.

The encoding is not merely elegant; it is \emph{necessary} for the rare classes.
Table~\ref{tab:smtbench} contrasts the SMT search with an ordered enumeration
that scans the input space for a witness. For dense classes (a zero divisor, an
out-of-range shift) both paths answer instantly. But for the classes whose
witness is vanishingly rare---signed overflow under an ordered scan, and
\texttt{INT\_MIN/-1} (a single input pair in $2^{64}$)---enumeration exhausts a
multi-million-probe budget and finds \emph{nothing}, while Z3 returns the
witness in under a millisecond: a two-to-three-orders-of-magnitude speed-up that
is the quantitative case for the symbolic encoding. A differential cross-check
confirms the two paths agree on \emph{existence} at every fully enumerable
width, and the SMT witness for \texttt{INT\_MIN/-1} coincides exactly with the
pair the shipped operational oracle emits.

\begin{table}[t]
\centering
\small
\caption{Witness search, SMT vs.\ ordered enumeration, at 32-bit width (probe
budget $2^{20}$). For rare-witness classes enumeration spends its full budget
without success while Z3 answers in sub-millisecond time; both paths are trivial
on dense classes. Times are representative single-run wall-clock.}
\label{tab:smtbench}
\begin{tabular}{@{}lcccc@{}}
\toprule
Class & SMT finds? & Enum finds? & Enum probes & Speed-up \\
\midrule
Division by zero        & yes & yes & 1            & --- \\
Shift $\geq$ width       & yes & yes & 33           & --- \\
Signed overflow         & yes & \textbf{no} & $2^{20}$ & $\sim$100$\times$ \\
\texttt{INT\_MIN / -1}  & yes & \textbf{no} & $2^{20}$ & $\sim$800$\times$ \\
\bottomrule
\end{tabular}
\end{table}

%% ======================================================================
\section{Related Work}\label{sec:related}
%% ======================================================================

\paragraph{Translation validation.}
Alive2~\cite{lopes2021alive2} validates LLVM IR$\to$IR
transformations but operates \emph{after IR erasure}: the C11/Rust
semantic distinction is already lost, so UB divergences are
invisible.  CompCert~\cite{leroy2009compcert} provides end-to-end
compilation correctness for C but targets a single language.
Pnueli et al.~\cite{pnueli1998translation} formalized translation
validation as a proof obligation; \textsc{SemRec} adapts this to the
cross-language setting where the ``translation'' is a C$\to$Rust
migration.

\paragraph{IR-level verification.}
KLEE~\cite{cadar2008klee} performs symbolic execution on LLVM IR.
SMACK~\cite{rakamaric2014smack} translates LLVM IR to Boogie.
SeaHorn~\cite{gurfinkel2015seahorn} uses constrained Horn clauses.
CBMC~\cite{kroening2014cbmc} performs bounded model checking on C.
All operate on a single language or post-erasure IR.  As demonstrated
in Section~\ref{sec:eval-ub}, IR-level analysis achieves 0/30
recall on UB-invisible divergences: the signed/unsigned distinction
that distinguishes C UB from Rust wrapping is erased by compilation.
\textsc{SemRec} is the only tool in this design space that detects
these divergences.

\paragraph{Cross-language verification.}
The problem of verifying equivalence across language boundaries has
received limited attention.  Most existing work targets single-language
settings or FFI ABI correctness.  \textsc{SemRec} is, to our
knowledge, the first tool to perform source-level bounded equivalence
verification for C$\to$Rust migration with formal UB-aware soundness
guarantees.

\paragraph{C$\to$Rust migration tools.}
C2Rust~\cite{c2rust2019} transpiles C to syntactically valid but
heavily \texttt{unsafe} Rust.  Crown~\cite{zhang2023ownership} and
Laertes~\cite{emre2021translating} refactor C2Rust output toward
safe idioms.  LLM-based translators produce more idiomatic Rust but
lack formal correctness guarantees.  None perform source-level
equivalence verification or detect UB divergences.  Our Tier-1 corpus is
therefore complementary: it does not replace c2rust, but validates selected
c2rust-generated artifacts against the source-language UB oracle.

\paragraph{Formal verification for Rust.}
Kani~\cite{kani2022} performs BMC on Rust (single-language).
Prusti~\cite{astrauskas2022prusti} verifies Rust against
specifications.  RustBelt~\cite{jung2017rustbelt} provides semantic
foundations for Rust's type system.
RefinedC~\cite{sammler2021refinedc} verifies C with refinement types.
These verify within a single language; \textsc{SemRec} verifies
\emph{cross-language} equivalence.

\paragraph{Product programs.}
Barthe et al.~\cite{barthe2011product} introduced product programs
for relational verification of two runs of the \emph{same} program.
\textsc{SemRec} extends this to two \emph{different} programs in
\emph{different} languages with heterogeneous UB semantics,
adding $\sigma$-bridge coercion nodes at semantic boundaries---the
key mechanism enabling detection of IR-invisible divergences.


%% ======================================================================
\section{Conclusion}\label{sec:conclusion}
%% ======================================================================

We presented \textsc{SemRec}, a bounded source-level equivalence
verifier for C$\to$Rust migration that detects UB divergences
\emph{invisible to every existing verification approach}.  On 30
IR-invisible UB divergences spanning 6 classes, \textsc{SemRec}
achieves \textbf{100\% recall} while differential testing, IR
comparison, and IR-level verification tools (KLEE, SMACK, Alive2)
all score \textbf{0\%}.

The key enabler is the $\sigma$-bridge product program construction,
which encodes C11 and Rust UB semantics as separate SMT assumptions
within a shared bitvector formula.  This preserves the
signed/unsigned, division, and shift UB distinctions that compilation
erases.  The soundness theorem (Theorem~\ref{thm:soundness})
guarantees that UNSAT certifies equivalence on all well-defined
inputs within the loop bound.

General-purpose accuracy (35.4\% on 511 pairs) reflects parser and IR
coverage limitations, not verification unsoundness.  Where the
pipeline completes, the formal guarantees hold.  We view
\textsc{SemRec} as complementary to differential testing: testing
covers broad program behavior but misses UB divergences by design;
\textsc{SemRec} catches exactly the bugs that testing cannot.

The tool is open-source, passes 961 tests, and is under active
development.  Future work includes interprocedural verification,
struct/enum lowering, and integration with C2Rust for automated
post-migration verification.


%% ======================================================================
%% BIBLIOGRAPHY
%% ======================================================================
\bibliographystyle{plain}
\begin{thebibliography}{35}

\bibitem{cisa2024sbd}
Cybersecurity and Infrastructure Security Agency.
\newblock Secure by Design pledge.
\newblock \url{https://www.cisa.gov/securebydesign/pledge}, 2024.

\bibitem{nsa2022memory}
National Security Agency.
\newblock Software memory safety.
\newblock Cybersecurity Information Sheet, November 2022.

\bibitem{whitehouse2024report}
Office of the National Cyber Director.
\newblock Back to the building blocks: A path toward secure and measurable software.
\newblock White House Technical Report, February 2024.

\bibitem{chromium2019memory}
Chromium Project.
\newblock Memory safety.
\newblock \url{https://www.chromium.org/Home/chromium-security/memory-safety/}, 2019.

\bibitem{microsoft2019safety}
M.~Miller.
\newblock Trends, challenges, and strategic shifts in the software vulnerability mitigation landscape.
\newblock Microsoft Security Response Center, 2019.

\bibitem{linuxrust2022}
M.~Ojeda et~al.
\newblock Rust for {Linux}.
\newblock \url{https://rust-for-linux.com/}, 2022.

\bibitem{c2rust2019}
P.~Emre, G.~Kondoh, and Z.~Anderson.
\newblock {C2Rust}: Migrating legacy code to {Rust}.
\newblock \emph{Proc.\ ICSE-Companion}, pp.\ 258--259, 2019.

\bibitem{zhang2023ownership}
H.~Zhang, J.~He, B.~Ding, and Y.~Zhang.
\newblock Ownership guided {C} to {Rust} translation.
\newblock \emph{Proc.\ OOPSLA}, 2023.

\bibitem{emre2021translating}
M.~Emre, R.~Schroeder, K.~Dewey, and B.~Hardekopf.
\newblock Translating {C} to safer {Rust}.
\newblock \emph{Proc.\ OOPSLA}, 2021.

\bibitem{iso9899}
{ISO/IEC}.
\newblock {ISO/IEC 9899:2011} --- {P}rogramming languages --- {C}.
\newblock International Organization for Standardization, 2011.

\bibitem{cadar2008klee}
C.~Cadar, D.~Dunbar, and D.~Engler.
\newblock {KLEE}: Unassisted and automatic generation of high-coverage tests for complex systems programs.
\newblock \emph{Proc.\ OSDI}, 2008.

\bibitem{rakamaric2014smack}
Z.~Rakamaric and M.~Emmi.
\newblock {SMACK}: Decoupling source language details from verifier implementations.
\newblock \emph{Proc.\ CAV}, 2014.

\bibitem{gurfinkel2015seahorn}
A.~Gurfinkel, T.~Kahsai, A.~Komuravelli, and J.~Navas.
\newblock The {SeaHorn} verification framework.
\newblock \emph{Proc.\ CAV}, 2015.

\bibitem{lopes2021alive2}
N.~P.~Lopes, J.~Lee, C.-K.~Hur, Z.~Liu, and J.~Regehr.
\newblock Alive2: Bounded translation validation for {LLVM}.
\newblock \emph{Proc.\ PLDI}, 2021.

\bibitem{pnueli1998translation}
A.~Pnueli, M.~Siegel, and E.~Singerman.
\newblock Translation validation.
\newblock \emph{Proc.\ TACAS}, pp.\ 151--166, 1998.

\bibitem{leroy2009compcert}
X.~Leroy.
\newblock Formal verification of a realistic compiler.
\newblock \emph{Communications of the ACM}, 52(7):107--115, 2009.

\bibitem{demoura2008z3}
L.~de~Moura and N.~Bj{\o}rner.
\newblock {Z3}: An efficient {SMT} solver.
\newblock \emph{Proc.\ TACAS}, 2008.

\bibitem{kroening2014cbmc}
D.~Kroening and M.~Tautschnig.
\newblock {CBMC} -- {C} bounded model checker.
\newblock \emph{Proc.\ TACAS}, 2014.

\bibitem{kani2022}
Amazon Web Services.
\newblock Kani {Rust} verifier.
\newblock \url{https://model-checking.github.io/kani/}, 2022.

\bibitem{astrauskas2022prusti}
V.~Astrauskas, A.~Bi\-le, P.~M\"uller, and F.~Poli.
\newblock Prusti: A practical verification infrastructure for {Rust}.
\newblock \emph{Proc.\ OOPSLA}, 2022.

\bibitem{jung2017rustbelt}
R.~Jung, J.-H.~Jourdan, R.~Krebbers, and D.~Dreyer.
\newblock {RustBelt}: Securing the foundations of the {Rust} programming language.
\newblock \emph{Proc.\ ACM Program.\ Lang.}, 2(POPL):66:1--66:34, 2018.

\bibitem{sammler2021refinedc}
M.~Sammler, R.~Lepigre, R.~Krebbers, et~al.
\newblock {RefinedC}: Automating the foundational verification of {C} code with refined ownership types.
\newblock \emph{Proc.\ PLDI}, 2021.

\bibitem{barthe2011product}
G.~Barthe, J.~M.~Crespo, and C.~Kunz.
\newblock Relational verification using product programs.
\newblock \emph{Proc.\ FM}, pp.\ 200--214, 2011.

\bibitem{treesitter2024}
M.~Brunsfeld et~al.
\newblock Tree-sitter: An incremental parsing system for programming tools.
\newblock \url{https://tree-sitter.github.io/}, 2024.

\bibitem{pratt1973top}
V.~Pratt.
\newblock Top down operator precedence.
\newblock \emph{Proc.\ POPL}, 1973.

\bibitem{cytron1991ssa}
R.~Cytron, J.~Ferrante, B.~K.~Rosen, M.~N.~Wegman, and F.~K.~Zadeck.
\newblock Efficiently computing static single assignment form and the control dependence graph.
\newblock \emph{ACM Trans.\ Program.\ Lang.\ Syst.}, 13(4):451--490, 1991.

\bibitem{andersen1994program}
L.~O.~Andersen.
\newblock Program analysis and specialization for the {C} programming language.
\newblock PhD thesis, DIKU, University of Copenhagen, 1994.

\bibitem{king1976symbolic}
J.~C.~King.
\newblock Symbolic execution and program testing.
\newblock \emph{Communications of the ACM}, 19(7):385--394, 1976.

\bibitem{reynolds2002separation}
J.~C.~Reynolds.
\newblock Separation logic: A logic for shared mutable data structures.
\newblock \emph{Proc.\ LICS}, pp.\ 55--74, 2002.

\end{thebibliography}


%% ======================================================================
%% APPENDIX (20 pages)
%% ======================================================================
\clearpage
\begin{appendices}

%% ======================================================================
\section{Full Soundness Proof}\label{app:proofs}
%% ======================================================================

This appendix presents complete proofs of all lemmas and theorems
stated in Section~\ref{sec:soundness}.  The proof structure mirrors
the verification logic in
\texttt{src/product\_program/soundness.py}.

\subsection{Supporting Definitions}

\begin{definition}[Semantic State]\label{def:semstate}
A semantic state $S = (V, M, P)$ consists of a variable environment
$V: \mathit{Var} \to \mathit{BitVec}(w)$, a memory
$M: \mathit{Addr} \to \mathit{Byte}$, and a set of panic flags
$P \subseteq \mathit{Inst}$.  We write $S_C$ and $S_R$ for the
C-side and Rust-side states, respectively.
\end{definition}

\begin{definition}[Simulation Relation]\label{def:simrel}
The simulation relation $R$ between a product program state~$P$ and
a paired state $(S_C, S_R)$ holds when:
(i)~for each shared input~$x_i$, the product program's value equals
both sides' coerced values;
(ii)~for each prefixed variable $\texttt{c\_}v$ in $P$, the value
equals $S_C(v)$, and analogously for $\texttt{r\_}v$;
(iii)~the coercion assertions at each processed instruction are
satisfied.
\end{definition}

\subsection{Lemma~\ref{lem:bv-faithful}: Bitvector Encoding
Faithfulness}

\begin{lemma}[Bitvector Encoding Faithfulness]\label{lem:bv-faithful}
For each arithmetic operation $op \in \{+, -, \times, /, \%,
\ll, \gg\}$ on a $w$-bit signed type~$\tau$, and for all concrete
values $a, b \in [-2^{w-1}, 2^{w-1}-1]$, the Z3 bitvector encoding
$\texttt{bv}_{op}(a, b)$ equals the mathematical result
$a \mathbin{op} b \bmod 2^w$ interpreted as a signed $w$-bit integer.
\end{lemma}

\begin{proof}
Z3's bitvector theory implements fixed-width arithmetic modulo~$2^w$.
For a $w$-bit signed type, the representation maps the unsigned value
$v \in [0, 2^w)$ to the signed value
$\hat{v} = v - 2^w \cdot \mathbf{1}[v \geq 2^{w-1}]$.

\emph{Addition.}
$\texttt{bvadd}(a, b) = (a + b) \bmod 2^w$.
The signed interpretation gives
$\widehat{(a + b) \bmod 2^w}$, which equals $a + b$ when
$-2^{w-1} \leq a + b \leq 2^{w-1}-1$ (no overflow), and wraps
otherwise.  This matches C's modular arithmetic at the hardware level
and Rust's \texttt{wrapping\_add}.  The encoder
(\texttt{encode\_binop} in \texttt{src/smt/encoder.py}) uses
\texttt{z3.BitVecVal} and \texttt{z3.Extract} to maintain
exact-width results.

\emph{Subtraction, multiplication.}
Analogous: $\texttt{bvsub}(a,b) = (a - b) \bmod 2^w$,
$\texttt{bvmul}(a,b) = (a \cdot b) \bmod 2^w$.  For multiplication,
the encoder's \texttt{\_signed\_overflow\_check} sign-extends both
operands to~$2w$ bits, multiplies, truncates, and checks whether
sign-extension matches, faithfully detecting overflow.

\emph{Division.}
$\texttt{bvsdiv}(a, b)$ implements truncation toward zero for
$b \neq 0$, matching C11~\S6.5.5/6.  The corner case
$a = -2^{w-1}, b = -1$ yields $2^{w-1}$, which overflows the signed
range; this is excluded by the $\mathit{WellDefined}$ guard, which
adds $b \neq 0 \wedge \neg(a = -2^{w-1} \wedge b = -1)$.

\emph{Remainder.}
$\texttt{bvsrem}(a, b)$ satisfies
$a = \texttt{bvsdiv}(a,b) \cdot b + \texttt{bvsrem}(a,b)$ with the
sign of the remainder matching the sign of the dividend, consistent
with C11~\S6.5.5/6.

\emph{Shifts.}
$\texttt{bvshl}(a, s) = (a \cdot 2^s) \bmod 2^w$ for
$0 \leq s < w$.  For $s \geq w$, Z3 returns~0 (SMT-LIB semantics);
this is guarded by $\mathit{WellDefined}$ for the C side (shift by
$\geq w$ is UB under C11~\S6.5.7/3) and modeled as
$s' = s \bmod w$ for the Rust side (Rust wraps the shift amount).
$\texttt{bvashr}(a, s)$ performs arithmetic right shift with sign
extension, matching both languages' defined behavior for valid shift
amounts.
\end{proof}

\subsection{Lemma~\ref{lem:wd-complete}: WellDefined Predicate
Completeness}

\begin{lemma}[WellDefined Predicate Completeness]\label{lem:wd-complete}
The predicate $\mathit{WellDefined}(\sigma_C, f_C, \vec{x})$ is true
if and only if evaluating $f_C(\vec{x})$ under C11 semantics does not
trigger any undefined behavior from the set
$\mathcal{U} = \{\text{signed overflow}, \text{division by zero},
\text{INT\_MIN}/(-1), \text{shift} \geq w,
\text{INT\_MIN negation}\}$.
\end{lemma}

\begin{proof}
The predicate is constructed as a conjunction of guards, one per
arithmetic operation in~$f_C$.  The encoder
(\texttt{encode\_binop} with \texttt{OverflowBehavior.UNDEFINED})
generates these guards as \texttt{ctx.assert\_assume} calls.

\emph{Guard construction.}
\begin{enumerate}
  \item For each signed $op \in \{+, -, \times\}$ on operands
    $(e_1, e_2)$ of width~$w$: add guard
    $-2^{w-1} \leq e_1 \mathbin{op} e_2 \leq 2^{w-1}-1$.
    This captures C11~\S6.5/5 (signed overflow is UB).
    The implementation checks
    $(e_1 > 0 \wedge e_2 > 0 \wedge r < 0)
    \vee (e_1 < 0 \wedge e_2 < 0 \wedge r > 0)$
    for addition (negated, as a well-definedness assumption).

  \item For each division $e_1 / e_2$ on signed type: add guards
    $e_2 \neq 0$ and $\neg(e_1 = -2^{w-1} \wedge e_2 = -1)$.
    This captures C11~\S6.5.5/5.

  \item For each shift $e_1 \mathbin{\ll} e_2$ or
    $e_1 \mathbin{\gg} e_2$ on type of width~$w$: add guard
    $0 \leq e_2 < w$.  This captures C11~\S6.5.7/3.

  \item For signed unary negation $-e$ of width~$w$: add guard
    $e \neq -2^{w-1}$.  This captures C11~\S6.5/5 applied to
    unary minus.
\end{enumerate}

\emph{Soundness} ($\Rightarrow$): If all guards hold, then by
construction no operation in $f_C$ triggers UB from~$\mathcal{U}$.

\emph{Completeness} ($\Leftarrow$): If $f_C(\vec{x})$ does not
trigger any UB from~$\mathcal{U}$, then every arithmetic operation
produces a result within the representable range, every divisor is
nonzero (and not the $-2^{w-1}/-1$ case), every shift amount is
in $[0,w)$, and no negation of $-2^{w-1}$ occurs.  Each guard holds,
so $\mathit{WellDefined}$ is true.

\emph{Scope limitation.}
The predicate covers the five UB classes in~$\mathcal{U}$.  Other
C11 UB sources---strict aliasing violations (\S6.5/7),
sequence-point violations (\S6.5/2), uninitialized reads
(\S6.3.2.1/2), null pointer dereference (\S6.5.3.2/4)---are
outside our current scope and represent a known source of false
negatives.
\end{proof}

\subsection{Lemma~\ref{lem:sigma-correct}: $\sigma$-Bridge Coercion
Correctness}

\begin{lemma}[$\sigma$-Bridge Coercion Correctness]\label{lem:sigma-correct}
For each $\sigma$-bridge coercion
$\sigma_i = (\mathit{pre}_i, \mathit{post}_i, \mathit{dom}_i)$ in
the 17 handled classes: if $\mathit{pre}_i(\vec{x})$ holds, then
$\mathit{post}_i(\vec{x}, y_C, y_R)$ holds, where $y_C$ and $y_R$
are the C and Rust outputs of the coerced operation.
\end{lemma}

\begin{proof}
By case analysis on each coercion class.  We present all 17 cases,
referencing the coercion specifications in
\texttt{src/product\_program/soundness.py} (lines 53--312).

\emph{Case 1: Signed addition overflow.}
Pre: $-2^{w-1} \leq a + b \leq 2^{w-1}-1$.
Post: $y_C = y_R = a + b$.
Under the precondition, $\texttt{bvadd}(a,b) = a + b$ (no wrapping).
C avoids UB; Rust's modular result coincides.  $\square$

\emph{Case 2: Signed subtraction overflow.}
Pre: $-2^{w-1} \leq a - b \leq 2^{w-1}-1$.
Post: $y_C = y_R = a - b$.
Analogous to Case~1.  $\square$

\emph{Case 3: Signed multiplication overflow.}
Pre: $-2^{w-1} \leq a \times b \leq 2^{w-1}-1$.
Post: $y_C = y_R = a \times b$.
The encoder sign-extends to $2w$ bits, multiplies, and checks that
truncation preserves the result.  Under the precondition, no
information is lost.  $\square$

\emph{Case 4: Signed negation of INT\_MIN.}
Pre: $a \neq -2^{w-1}$.
Post: $y_C = y_R = -a$.
Under the precondition, $-a \in [-2^{w-1}+1, 2^{w-1}-1]$, so
$\texttt{bvneg}(a) = -a$ without overflow.  $\square$

\emph{Case 5: Unsigned wrap.}
Pre: (always true---both languages wrap unsigned arithmetic).
Post: $y_C = y_R = (a \mathbin{op} b) \bmod 2^w$.
Unsigned arithmetic is well-defined in both C11 (\S6.2.5/9) and
Rust.  $\square$

\emph{Case 6: Mixed signed/unsigned promotion.}
Pre: (always true---IR lowers promotions explicitly).
Post: $y_C = y_R$ after explicit widening.
The IR inserts explicit \texttt{SEXT}/\texttt{ZEXT} instructions
matching each language's promotion rules.  $\square$

\emph{Case 7: Division by zero.}
Pre: $b \neq 0$.
Post: $y_C = y_R = \lfloor a / b \rfloor$ (truncated toward zero).
Both C and Rust agree on division semantics when $b \neq 0$ and the
$-2^{w-1}/-1$ case is excluded.  $\square$

\emph{Case 8: INT\_MIN / -1.}
Pre: $\neg(a = -2^{w-1} \wedge b = -1)$.
Post: $y_C = y_R = \lfloor a / b \rfloor$.
The combined guard with Case~7 ensures that
$\texttt{bvsdiv}(a,b)$ is well-defined in both languages.  $\square$

\emph{Case 9: Signed remainder semantics.}
Pre: $b \neq 0 \wedge \neg(a = -2^{w-1} \wedge b = -1)$.
Post: $y_C = y_R = a - \lfloor a/b \rfloor \cdot b$.
C11~\S6.5.5/6 specifies truncation toward zero, matching
Rust and $\texttt{bvsrem}$.  $\square$

\emph{Case 10: Shift $\geq$ bit width.}
Pre: $0 \leq s < w$.
Post: $y_C = y_R = \texttt{bvshl}(a, s)$ or
$\texttt{bvashr}(a, s)$.
C avoids UB; Rust's $s' = s \bmod w = s$ since $s < w$.  $\square$

\emph{Case 11: Negative shift amount.}
Pre: $s \geq 0$.
Post: $y_C = y_R$.
Combined with the upper bound $s < w$ from Case~10.  $\square$

\emph{Case 12: Integer narrowing truncation.}
Pre: (always defined).
Post: $y_C = y_R = a \bmod 2^{w'}$ where $w'$ is the target width.
Both C (implementation-defined but universally truncating on
two's-complement architectures) and Rust truncate.  $\square$

\emph{Case 13: Sign extension vs.\ zero extension.}
Pre: (always defined).
Post: $y_C = y_R$.
The IR explicitly chooses \texttt{SEXT} or \texttt{ZEXT} based on
source signedness; both languages agree.  $\square$

\emph{Case 14: Float-to-integer truncation.}
Pre: $\mathit{INT\_MIN} \leq x \leq \mathit{INT\_MAX}$ and $x$
is not NaN.
Post: $y_C = y_R = \lfloor x \rfloor$ (truncated toward zero).
Under the precondition, both languages truncate identically.
Outside the range, C has UB and Rust saturates; the precondition
excludes this case.  $\square$

\emph{Case 15: Null pointer dereference.}
Pre: C: $p \neq \texttt{NULL}$; Rust: $p$ is \texttt{Some}.
Post: Both access valid data.
Under the precondition, both sides dereference a valid pointer.
$\square$

\emph{Case 16: Array out-of-bounds.}
Pre: $0 \leq i < \mathit{len}$.
Post: $y_C = y_R = M[\mathit{base} + i \cdot \mathit{sizeof}(\tau)]$.
Under bounds, both access the same element.  $\square$

\emph{Case 17: Short-circuit evaluation order.}
Pre: (structural match---both sides use short-circuit evaluation).
Post: Same boolean result and same evaluation order.
C11~\S6.5.13--14 and Rust both specify left-to-right short-circuit
evaluation for \texttt{\&\&} and \texttt{||}.  $\square$

Each case is also validated by boundary-value testing on
$\{0, \pm 1, \pm(2^{w-1}-1), -2^{w-1}, 2^w - 1\}$ in the
test suite.
\end{proof}

\subsection{Lemma~\ref{lem:var-isolation}: Product Program Variable
Isolation}

\begin{lemma}[Product Program Variable Isolation]\label{lem:var-isolation}
In the product program $f_C \bowtie f_R$, the prefixed variable
namespaces \texttt{c\_} and \texttt{r\_} are disjoint: no SSA
variable in $B_C$ is referenced by $B_R$ and vice versa, except
through the shared inputs~$\vec{x}$.
\end{lemma}

\begin{proof}
By construction in Algorithm~\ref{alg:product}, lines 8--9: the
bodies $B_C$ and $B_R$ undergo prefix renaming with disjoint prefixes
\texttt{c\_} and \texttt{r\_}.  Since SSA variables are uniquely named
within each body (by the SSA property~\cite{cytron1991ssa}), and the
prefixes are distinct strings, no variable name in the renamed~$B_C$
can equal any in the renamed~$B_R$.  The only shared symbols are the
input variables $\vec{x}$ (line~3), which are read-only in both
bodies.

In the implementation, \texttt{ProductBuilder.\_build\_product\_blocks}
constructs \texttt{ProductBlock} objects containing left-only and
right-only instructions, with variable names derived from their
respective function's namespace.  The \texttt{SharedInput} objects
are the sole cross-reference points.
\end{proof}

\subsection{Theorem~\ref{thm:soundness}: Soundness}

\begin{proof}
Suppose $\Phi$ is unsatisfiable, i.e., there is no $\vec{x}$
satisfying
\[
\mathit{WellDefined}(\sigma_C, f_C, \vec{x}) \;\wedge\;
\bigl(\llbracket f_C \rrbracket_{\sigma_C}(\vec{x}) \neq
\llbracket f_R \rrbracket_{\sigma_R}(\vec{x})\bigr).
\]

Suppose for contradiction that there exists $\vec{x}_0$ with:
(a)~$\mathit{WellDefined}(\sigma_C, f_C, \vec{x}_0)$ true,
(b)~all execution paths through $f_C(\vec{x}_0)$ and
$f_R(\vec{x}_0)$ of length $\leq K$, and
(c)~$\llbracket f_C \rrbracket(\vec{x}_0) \neq
\llbracket f_R \rrbracket(\vec{x}_0)$.

\emph{Step 1 (Encoding faithfulness).}
By Lemma~\ref{lem:bv-faithful}, the Z3 bitvector encoding of both
$\llbracket f_C \rrbracket$ and $\llbracket f_R \rrbracket$
evaluates to the same values as the mathematical semantics on
concrete inputs.  In particular,
$\llbracket f_C \rrbracket_{\sigma_C}(\vec{x}_0)$ in the Z3
encoding equals $f_C(\vec{x}_0)$ and
$\llbracket f_R \rrbracket_{\sigma_R}(\vec{x}_0)$ equals
$f_R(\vec{x}_0)$.

\emph{Step 2 (Well-definedness).}
By Lemma~\ref{lem:wd-complete},
$\mathit{WellDefined}(\sigma_C, f_C, \vec{x}_0)$ in the SMT encoding
is true iff $f_C(\vec{x}_0)$ does not trigger UB
from~$\mathcal{U}$.  By assumption~(a), this holds.

\emph{Step 3 (Coercion correctness).}
By Lemma~\ref{lem:sigma-correct}, at each $\sigma$-bridge coercion
point, the precondition (which is implied by well-definedness)
guarantees that the coerced encodings relate the C and Rust outputs
correctly.

\emph{Step 4 (Variable isolation).}
By Lemma~\ref{lem:var-isolation}, the two sides of the product
program do not interfere: the only shared values are the
inputs~$\vec{x}$.

\emph{Step 5 (Path coverage).}
By assumption~(b), all relevant execution paths have length $\leq K$,
so the bounded unrolling explores the complete behavior.

\emph{Conclusion.}
Combining Steps~1--5, the concrete input $\vec{x}_0$ satisfies all
conjuncts of $\Phi$: well-definedness holds (Step~2), encoding
faithfulness ensures the symbolic formula evaluates correctly
(Step~1), coercion correctness maintains the semantic relationship
(Step~3), variable isolation prevents interference (Step~4), and
path coverage ensures the divergence is explored (Step~5).  Thus
$\vec{x}_0$ is a satisfying assignment for $\Phi$, contradicting
UNSAT.  \qed
\end{proof}

\subsection{Theorem~\ref{thm:cex-sound}: Counterexample Soundness}

\begin{proof}
Suppose $\Phi$ is satisfiable with model $\mathcal{M}$.  Let
$\vec{x}_0 = \mathcal{M}(\vec{x})$.

By the SAT result, $\mathcal{M}$ satisfies both conjuncts:
\begin{enumerate}
  \item $\mathit{WellDefined}(\sigma_C, f_C, \vec{x}_0)$ is true.
    By Lemma~\ref{lem:wd-complete}, this means $f_C(\vec{x}_0)$
    does not trigger UB from $\mathcal{U}$.

  \item $\llbracket f_C \rrbracket_{\sigma_C}(\vec{x}_0) \neq
    \llbracket f_R \rrbracket_{\sigma_R}(\vec{x}_0)$.
    By Lemma~\ref{lem:bv-faithful}, the Z3 model values correspond
    to concrete execution values.
\end{enumerate}

Therefore $\vec{x}_0$ is a well-defined input on which the two
functions produce different outputs: a genuine divergence witness.

The implementation extracts this witness via
\texttt{solver.model()} and reports it as a concrete
counterexample with the divergence class annotation from the
$\sigma$-bridge coercion that was violated.  \qed
\end{proof}

\subsection{Theorem~\ref{thm:completeness}: Completeness
Characterization}

\begin{proof}
We show that under conditions \textbf{C1}--\textbf{C4}, if $f_C$ and
$f_R$ are genuinely divergent on some well-defined input, then $\Phi$
is SAT.

Let $\vec{x}_0$ be a divergence witness: $f_C(\vec{x}_0)$ is
well-defined and $f_C(\vec{x}_0) \neq f_R(\vec{x}_0)$.

Under \textbf{C1} (path length $\leq K$): the bounded unrolling
explores the execution path through $\vec{x}_0$.

Under \textbf{C2} (handled divergence classes): every operation
causing the divergence has a $\sigma$-bridge coercion that
faithfully encodes the semantic difference
(Lemma~\ref{lem:sigma-correct}).

Under \textbf{C3} (successful IR lowering): the SMT formula
faithfully represents both source programs, so the encoding of
$\vec{x}_0$ captures the divergence.

Under \textbf{C4} (no approximation): QF\_BV is decidable, so Z3
will return a definitive verdict.  Since $\vec{x}_0$ satisfies
$\Phi$ (by the preceding three conditions), Z3 returns SAT.

\emph{Failure modes when conditions are violated:}
\begin{itemize}[nosep]
  \item \textbf{C1 violated}: divergence hidden beyond the loop
    bound.  Increasing $K$ may help; a $K$-sensitivity experiment
    is planned but not yet validated
    (Section~\ref{sec:eval-k}).
  \item \textbf{C2 violated}: unhandled divergence class (e.g.,
    pointer aliasing).  The tool reports \textsc{Unknown} or a
    false \textsc{Equivalent}.
  \item \textbf{C3 violated}: IR lowering failure.  The tool
    reports \textsc{Unknown} with a diagnostic.
  \item \textbf{C4 violated}: floating-point approximation.  May
    produce false \textsc{Equivalent} for NaN-sensitive code.
\end{itemize}
\qed
\end{proof}

\subsection{Additional Results}

\begin{lemma}[Simulation Preservation]\label{lem:sim-pres}
Let $R$ be the simulation relation
(Definition~\ref{def:simrel}).  If $R(P, (S_C, S_R))$ holds at
product instruction~$i$ and $P$ takes one step to $i+1$, then
$R(P', (S_C', S_R'))$ holds at $i+1$.
\end{lemma}

\begin{proof}
By case analysis on the instruction at position~$i$.

\emph{Matched instruction (both sides execute).}
The product instruction wraps aligned instructions $I_C$ and $I_R$.
If no coercion is needed ($\sigma_C = \sigma_R$ at this point),
both sides compute the same bitvector operation, and $R$ is preserved
trivially.  If a coercion is present, Lemma~\ref{lem:sigma-correct}
guarantees the postcondition holds under the well-definedness
assumption, maintaining~$R$.

\emph{Left-only instruction (C side only).}
The instruction updates only \texttt{c\_}-prefixed variables.  By
Lemma~\ref{lem:var-isolation}, the Rust-side state is unaffected.
$R$ is preserved since the update is consistent with $S_C'$.

\emph{Right-only instruction (Rust side only).}
Symmetric to the left-only case.  \qed
\end{proof}

\begin{proposition}[BMC Monotonicity]\label{prop:bmc-mono}
For loop unrolling bounds $K_1 < K_2$: if $\Phi_{K_2}$ is UNSAT,
then $\Phi_{K_1}$ is UNSAT.
\end{proposition}

\begin{proof}
$\Phi_{K_1}$ is a logical weakening of $\Phi_{K_2}$: every
execution path explored at depth~$K_1$ is also explored at
depth~$K_2$.  If no divergence exists within~$K_2$ iterations,
\emph{a fortiori} none exists within~$K_1$ iterations.

Contrapositively: SAT for $\Phi_{K_1}$ implies SAT for
$\Phi_{K_2}$, since the witnessing path is also present in the
deeper unrolling.  \qed
\end{proof}


%% ======================================================================
\section{$\sigma$-Bridge Coercion Catalog}\label{app:sigma-catalog}
%% ======================================================================

Table~\ref{tab:full-catalog} lists all 26 divergence classes with
their coercion specifications.  Each coercion is formally specified
as a triple $(\mathit{pre}, \mathit{post}, \mathit{dom})$ in
\texttt{src/product\_program/soundness.py}, with implementations
in \texttt{src/product\_program/coercion.py}.

\begin{table}[h!]
\centering
\small
\caption{Complete $\sigma$-bridge coercion catalog.
$\checkmark$~=~handled, $\times$~=~unhandled.}
\label{tab:full-catalog}
\begin{tabular}{@{}rllcl@{}}
\toprule
\# & Cat. & Divergence Class & St. & Coercion \\
\midrule
1  & Arith & Signed addition overflow & $\checkmark$ &
  \textsc{Overflow} \\
2  & Arith & Signed subtraction overflow & $\checkmark$ &
  \textsc{Overflow} \\
3  & Arith & Signed multiplication overflow & $\checkmark$ &
  \textsc{Overflow} \\
4  & Arith & Signed negation of \texttt{INT\_MIN} & $\checkmark$ &
  \textsc{Negation} \\
5  & Arith & Unsigned wrap vs.\ defined & $\checkmark$ &
  \textsc{Overflow} \\
6  & Arith & Mixed signed/unsigned promotion & $\checkmark$ &
  \textsc{Cast} \\
\midrule
7  & Div & Division by zero & $\checkmark$ & \textsc{Division} \\
8  & Div & \texttt{INT\_MIN} / $-1$ & $\checkmark$ &
  \textsc{Division} \\
9  & Div & Signed remainder semantics & $\checkmark$ &
  \textsc{Division} \\
\midrule
10 & Shift & Shift $\geq$ bit width & $\checkmark$ &
  \textsc{Shift} \\
11 & Shift & Negative shift amount & $\checkmark$ &
  \textsc{Shift} \\
12 & Shift & Variable-width shift (non-std.\ types) & $\times$ &
  --- \\
\midrule
13 & Type & Integer narrowing truncation & $\checkmark$ &
  \textsc{Cast} \\
14 & Type & Sign extension vs.\ zero extension & $\checkmark$ &
  \textsc{Cast} \\
15 & Type & Float-to-integer truncation & $\checkmark$ &
  \textsc{Cast} \\
16 & Type & Union type reinterpretation & $\times$ & --- \\
\midrule
17 & Ptr & Null pointer dereference & $\checkmark$ &
  \textsc{Pointer} \\
18 & Ptr & Pointer aliasing (mutable refs) & $\times$ & --- \\
19 & Ptr & Lifetime/dangling pointer & $\times$ & --- \\
20 & Ptr & Allocation semantics (malloc/Box) & $\times$ & --- \\
21 & Ptr & Deallocation semantics (free/Drop) & $\times$ & --- \\
\midrule
22 & Ctrl & Short-circuit evaluation order & $\checkmark$ &
  \textsc{Control} \\
23 & Ctrl & Exception handling (longjmp/panic) & $\times$ &
  --- \\
24 & Ctrl & setjmp/longjmp vs.\ Result/Option & $\times$ &
  --- \\
\midrule
25 & FP & IEEE~754 rounding mode divergence & $\checkmark$ &
  \textsc{Float} \\
26 & FP & NaN propagation semantics & $\times$ & --- \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Coercion Specifications}

Each handled coercion is specified as follows.  These correspond to
the assertion generator functions in
\texttt{src/product\_program/coercion.py}, dispatched through the
\texttt{DivergenceClass}$\to$generator mapping.

\paragraph{Overflow coercion} (Classes 1--3, 5).
\begin{itemize}[nosep]
  \item \textbf{Pre}: Result in range $[-2^{w-1}, 2^{w-1}-1]$.
  \item \textbf{Post}: $y_C = y_R = a \mathbin{op} b$.
  \item \textbf{Dom}: Signed integer types of width $w \in \{8, 16, 32, 64\}$.
  \item \textbf{Generator}: \texttt{generate\_overflow\_assertions()}.
\end{itemize}

\paragraph{Negation coercion} (Class 4).
\begin{itemize}[nosep]
  \item \textbf{Pre}: $a \neq -2^{w-1}$.
  \item \textbf{Post}: $y_C = y_R = -a$.
  \item \textbf{Dom}: Signed integer types.
  \item \textbf{Generator}: \texttt{generate\_negation\_assertions()}.
\end{itemize}

\paragraph{Division coercion} (Classes 7--9).
\begin{itemize}[nosep]
  \item \textbf{Pre}: $b \neq 0 \wedge \neg(a = -2^{w-1} \wedge b = -1)$.
  \item \textbf{Post}: $y_C = y_R = \lfloor a / b \rfloor$.
  \item \textbf{Dom}: Signed integer types.
  \item \textbf{Generator}: \texttt{generate\_division\_assertions()}.
\end{itemize}

\paragraph{Shift coercion} (Classes 10--11).
\begin{itemize}[nosep]
  \item \textbf{Pre}: $0 \leq s < w$.
  \item \textbf{Post}: $y_C = y_R$.
  \item \textbf{Dom}: Integer types of width $w$.
  \item \textbf{Generator}: \texttt{generate\_shift\_assertions()}.
\end{itemize}

\paragraph{Cast coercion} (Classes 6, 13--14).
\begin{itemize}[nosep]
  \item \textbf{Pre}: (always defined).
  \item \textbf{Post}: $y_C = y_R = a \bmod 2^{w'}$ (narrowing) or
    $y_C = y_R = \texttt{ext}(a)$ (widening).
  \item \textbf{Dom}: Integer types.
  \item \textbf{Generator}: \texttt{generate\_cast\_assertions()}.
\end{itemize}

\paragraph{Float-to-integer coercion} (Class 15).
\begin{itemize}[nosep]
  \item \textbf{Pre}: $\mathit{INT\_MIN} \leq x \leq \mathit{INT\_MAX}$
    and $x$ is not NaN.
  \item \textbf{Post}: $y_C = y_R = \lfloor x \rfloor$.
  \item \textbf{Dom}: Float$\to$integer conversions.
  \item \textbf{Generator}: \texttt{generate\_float\_to\_int\_assertions()}.
\end{itemize}

\paragraph{Pointer coercion} (Class 17).
\begin{itemize}[nosep]
  \item \textbf{Pre}: C: $p \neq \texttt{NULL}$; Rust: $p$ is \texttt{Some}.
  \item \textbf{Post}: Both access valid data.
  \item \textbf{Dom}: Pointer types.
  \item \textbf{Generator}: \texttt{generate\_null\_pointer\_assertions()}.
\end{itemize}

\paragraph{Control coercion} (Class 22).
\begin{itemize}[nosep]
  \item \textbf{Pre}: Structural match (both use short-circuit evaluation).
  \item \textbf{Post}: Same boolean result and evaluation order.
  \item \textbf{Dom}: Boolean expressions with side effects.
\end{itemize}

\paragraph{Float coercion} (Class 25).
\begin{itemize}[nosep]
  \item \textbf{Pre}: Operands finite and non-NaN.
  \item \textbf{Post}: Results match within IEEE~754 rounding.
  \item \textbf{Dom}: \texttt{float}/\texttt{f32} and
    \texttt{double}/\texttt{f64}.
  \item \textbf{Generator}: \texttt{generate\_float\_assertions()}.
\end{itemize}

\subsection{Extended Coercions}

The codebase also implements extended coercions for additional
divergence patterns not in the core 26 classes but encountered
in real C2Rust output:

\begin{itemize}[nosep]
  \item \textbf{Wrapping arithmetic}: explicit \texttt{.wrapping\_add()}
    etc.\ in Rust vs.\ plain arithmetic in C.  Always wraps mod $2^w$.
  \item \textbf{Checked arithmetic}: Rust
    \texttt{.checked\_add()} returning \texttt{Option<T>}.
  \item \textbf{Saturating arithmetic}: result clamped to
    $[\mathit{INT\_MIN}, \mathit{INT\_MAX}]$.
  \item \textbf{Struct layout}: \texttt{\#[repr(C)]} guarantees
    C-compatible field layout; without it, Rust may reorder fields.
  \item \textbf{Enum discriminant}: mapping C enum values to Rust
    enum variants.
  \item \textbf{Pointer cast}: alignment compatibility and address
    preservation.
  \item \textbf{Pointer provenance}: integer-to-pointer roundtrips.
  \item \textbf{Slice vs.\ raw pointer}: Rust slice bounds checking
    vs.\ C raw pointer access.
\end{itemize}


%% ======================================================================
\section{Divergence Class Details}\label{app:divergence-details}
%% ======================================================================

This appendix provides the full divergence taxonomy, including
concrete examples, C11 and Rust standard references, and detection
methodology for each class.

\subsection{Arithmetic Divergences (Classes 1--6)}

\begin{table}[h!]
\centering
\small
\caption{Arithmetic divergence classes with concrete examples.}
\label{tab:arith-details}
\begin{tabular}{@{}rp{5cm}p{6cm}@{}}
\toprule
\# & Divergence & Concrete example \\
\midrule
1 & Signed addition overflow: C has UB, Rust wraps &
  \texttt{INT\_MAX + 1}: C UB, Rust returns
  \texttt{INT\_MIN} \\
2 & Signed subtraction overflow &
  \texttt{INT\_MIN - 1}: C UB, Rust returns
  \texttt{INT\_MAX} \\
3 & Signed multiplication overflow &
  \texttt{INT\_MAX * 2}: C UB, Rust returns $-2$ \\
4 & Negation of INT\_MIN: C has UB, Rust wraps &
  \texttt{-INT\_MIN}: C UB, Rust returns
  \texttt{INT\_MIN} \\
5 & Unsigned wrap: both wrap, but C may optimize
  assuming no wrap for signed types &
  \texttt{UINT\_MAX + 1}: both return 0, but
  mixed expressions may diverge \\
6 & Integer promotion: C implicitly promotes
  \texttt{char}/\texttt{short} to \texttt{int} &
  \texttt{(char)200 + (char)200}: C promotes to
  \texttt{int} (400), Rust keeps \texttt{u8} (wraps to 144) \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Division and Shift Divergences (Classes 7--12)}

\begin{table}[h!]
\centering
\small
\caption{Division and shift divergence classes.}
\label{tab:div-shift-details}
\begin{tabular}{@{}rp{4cm}p{3cm}p{3cm}@{}}
\toprule
\# & Divergence & C behavior & Rust behavior \\
\midrule
7 & Division by zero & UB & Panic \\
8 & INT\_MIN / $-1$ & UB (overflow) & Panic \\
9 & Signed remainder & Truncation toward zero & Same \\
10 & Shift $\geq w$ & UB & Wrap shift amount \\
11 & Negative shift & UB & Wrap shift amount \\
12 & Variable-width shift & UB (varies) & Wrap \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Type Conversion Divergences (Classes 13--16)}

Integer narrowing (class~13) is implementation-defined in C but
universally truncates on two's-complement architectures.  Rust
always truncates via \texttt{as} cast.  Sign vs.\ zero extension
(class~14) depends on the signedness of the source type; both
languages agree when the IR inserts explicit extension instructions.
Float-to-integer truncation (class~15) is UB in C when the float
is out of the integer range; Rust saturates (since edition 2021).
Union type reinterpretation (class~16) allows type-punning in C but
is restricted in Rust (\texttt{unsafe} required, and Rust unions
do not guarantee C layout without \texttt{\#[repr(C)]}).

\subsection{Pointer and Memory Divergences (Classes 17--21)}

These represent the largest gap in \textsc{SemRec}'s current
coverage.  Null pointer handling (class~17) is modeled via
option-type coercion.  The remaining four classes---pointer aliasing,
lifetime/dangling, allocation, and deallocation---require heap-shape
reasoning (e.g., separation logic~\cite{reynolds2002separation})
or ownership-aware analysis that goes beyond QF\_BV/QF\_ABV.

Pointer aliasing (class~18) is particularly challenging: C's strict
aliasing rule (\S6.5/7) and Rust's borrow checker enforce different
invariants, and verifying their equivalence requires reasoning about
the set of valid pointer derivations.  Lifetime/dangling pointer
(class~19) involves Rust's borrow checker guarantees vs.\ C's lack
thereof.  These are fundamentally different type-system-level
properties that do not reduce to bitvector constraints.

\subsection{Control Flow and Floating-Point Divergences
(Classes 22--26)}

Short-circuit evaluation (class~22) is the only control flow class
handled.  Exception handling (class~23) and
\texttt{setjmp}/\texttt{longjmp} (class~24) involve non-local
control flow that cannot be modeled intraprocedurally.

IEEE~754 rounding mode divergence (class~25) is handled for basic
operations (add, subtract, multiply, divide) under the assumption
that both languages use round-to-nearest-even (the default on x86
and ARM).  NaN propagation (class~26) differs between C and Rust:
C11 does not specify which NaN is returned from operations on
multiple NaN inputs, while Rust follows a deterministic propagation
rule.  This class is unhandled because our QF\_BV encoding does not
model IEEE~754 NaN semantics.


%% ======================================================================
\section{Implementation Details}\label{app:impl-details}
%% ======================================================================

\subsection{IR Instruction Set}

The shared typed SSA IR (\texttt{src/ir/}) comprises 36 instruction
types, summarized in Table~\ref{tab:ir-instructions}.

\begin{table}[h!]
\centering
\small
\caption{Shared IR instruction set.}
\label{tab:ir-instructions}
\begin{tabular}{@{}llp{6cm}@{}}
\toprule
Category & Instructions & Notes \\
\midrule
Arithmetic & \texttt{BinaryOp} (18 opcodes) &
  ADD, SUB, MUL, SDIV, UDIV, SREM, UREM,
  FADD, FSUB, FMUL, FDIV, FREM, SHL, LSHR,
  ASHR, AND, OR, XOR.
  Each annotated with \texttt{OverflowBehavior}. \\
Unary & \texttt{UnaryOp} (4 opcodes) &
  NEG, NOT, BITWISE\_NOT, FNEG \\
Comparison & \texttt{CompareOp} (16 predicates) &
  Signed (SLT/SGT/SLE/SGE), unsigned
  (ULT/UGT/ULE/UGE), ordered float
  (OEQ/ONE/OLT/OGT/OLE/OGE),
  EQ, NE \\
Cast & \texttt{CastInst} (12 kinds) &
  TRUNC, ZEXT, SEXT, FPTRUNC, FPEXT,
  FPTOSI, FPTOUI, SITOFP, UITOFP, BITCAST,
  PTRTOINT, INTTOPTR \\
Memory & Load, Store, Alloca, GEP &
  Byte-addressable with alignment metadata \\
Control flow & Branch, Switch, Return, Call &
  Conditional and unconditional branch \\
SSA & Phi, Select, ExtractValue, InsertValue &
  Phi for SSA construction; Select for ternary \\
Intrinsics & Memcpy, Memset &
  Modeled as memory array operations \\
\bottomrule
\end{tabular}
\end{table}

\subsection{SMT Encoding Details}

\paragraph{Variable encoding.}
Each IR variable of type \texttt{IntType(w, signed)} is encoded as a
Z3 \texttt{BitVec(w)}.  Signedness is tracked in metadata and affects
comparison predicates (\texttt{bvslt} vs.\ \texttt{bvult}) and
extension operations (\texttt{SignExt} vs.\ \texttt{ZeroExt}).

\paragraph{Overflow mode dispatch.}
The encoder's \texttt{encode\_binop} method dispatches on
\texttt{OverflowBehavior}:

\begin{lstlisting}[language=Python,title={\small Overflow encoding
dispatch (simplified from \texttt{src/smt/encoder.py})}]
if behavior == WRAP:
    result = bvadd(lhs, rhs)
elif behavior == UNDEFINED:
    result = bvadd(lhs, rhs)
    ctx.assert_assume(Not(overflow_check))
elif behavior == TRAP:
    result = bvadd(lhs, rhs)
    ctx.assert_hard(Not(overflow_check))
elif behavior == SATURATE:
    result = If(overflow, clamp, bvadd(lhs, rhs))
elif behavior == CHECKED:
    result = bvadd(lhs, rhs)
    ctx.declare(f"{name}_overflow", overflow_check)
\end{lstlisting}

\paragraph{Memory array model.}
Memory is modeled as a Z3 \texttt{Array(BitVecSort(ptr\_width),
BitVecSort(8))} (byte-addressable).  The encoder
(\texttt{src/smt/memory\_model.py}) supports:
\begin{itemize}[nosep]
  \item \textbf{Load}: \texttt{Select(mem, addr)} with bounds and
    alignment checks.
  \item \textbf{Store}: \texttt{Store(mem, addr, val)} updating the
    array and checking against freed allocations.
  \item \textbf{Allocation tracking}: known allocations stored as
    $(base, size)$ pairs for bounds verification.
  \item \textbf{Rust non-aliasing}: for \texttt{\&mut T} references,
    assert $p \neq q$ for all simultaneously live pairs.
\end{itemize}

\paragraph{Product program SMT generation.}
The \texttt{to\_smt\_lib()} method on \texttt{ProductProgram}
generates a complete SMT-LIB2 script:
(1)~declare shared input variables;
(2)~declare per-side variables with \texttt{c\_}/\texttt{r\_} prefix;
(3)~assert assumptions (well-definedness guards);
(4)~assert the negated equivalence check
$\mathit{ret}_C \neq \mathit{ret}_R$ as a hard assertion;
(5)~emit \texttt{(check-sat)} and \texttt{(get-model)}.

\subsection{Parser Architecture}

Each language has a primary parser (tree-sitter) and a fallback
(hand-written recursive descent with Pratt parsing for expressions).
The parser pipeline:
\begin{enumerate}[nosep]
  \item Attempt tree-sitter parse (\texttt{src/frontend\_c/} or
    \texttt{src/frontend\_rust/}).
  \item If tree-sitter produces an \texttt{ERROR} node, fall back to
    the recursive-descent parser (\texttt{src/c\_parser.py},
    \texttt{src/rust\_parser.py}).
  \item Lower the resulting AST to the shared IR.
  \item If lowering fails, report \textsc{Unknown} with diagnostic.
\end{enumerate}

Parse failures concentrate on:
\begin{itemize}[nosep]
  \item C: complex macro expansions, GCC extensions
    (\texttt{\_\_attribute\_\_}, statement expressions),
    K\&R-style function declarations.
  \item Rust: complex generics with where clauses, procedural macros,
    async/await syntax, trait objects with lifetime bounds.
\end{itemize}


%% ======================================================================
\section{Additional Experimental Results}\label{app:experiments}
%% ======================================================================

\subsection{Per-Category Detailed Results (Combined Tier)}

Table~\ref{tab:combined-detail} reports per-category accuracy on
the 212-pair combined tier, which provides a finer-grained view
than the full 511-pair results.

\begin{table}[h!]
\centering
\small
\caption{Per-category accuracy on the 212-pair combined tier.
Strong performers: arithmetic, shift.  Weak: iterators, strings,
control flow, structs.}
\label{tab:combined-detail}
\begin{tabular}{@{}lrrrl@{}}
\toprule
Category & Pairs & Correct & Accuracy & Notes \\
\midrule
Shift          & 3   & 3   & 100.0\% & All coercions effective \\
Arithmetic     & 12  & 10  & 83.3\%  & 2 failures: complex exprs \\
Cast           & 15  & 10  & 66.7\%  & Float-to-int edge cases \\
Division       & 5   & 3   & 60.0\%  & INT\_MIN/$-1$ handled \\
Float          & 15  & 9   & 60.0\%  & NaN cases fail \\
Error handling & 8   & 4   & 50.0\%  & Result/Option mapping \\
Bitwise        & 8   & 3   & 37.5\%  & Complex bit patterns \\
Loop           & 8   & 2   & 25.0\%  & Accumulator overflow \\
C2Rust         & 27  & 5   & 18.5\%  & Struct/trait lowering \\
C2Rust realist.& 11  & 2   & 18.2\%  & Complex patterns \\
Enum           & 20  & 3   & 15.0\%  & Discriminant mapping \\
Memory         & 14  & 2   & 14.3\%  & Pointer semantics gap \\
Compound       & 15  & 2   & 13.3\%  & Multi-op chains \\
Control flow   & 10  & 0   & 0.0\%   & Complex branching \\
Iterator       & 15  & 0   & 0.0\%   & Trait-based iteration \\
String         & 2   & 0   & 0.0\%   & IR lowering failures \\
Struct         & 20  & 0   & 0.0\%   & Layout/lowering failures \\
Memory scaled  & 4   & 0   & 0.0\%   & Pointer semantics gap \\
\midrule
\textbf{Total} & \textbf{212} & \textbf{58} & \textbf{27.4\%} & \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Analysis of zero-accuracy categories.}
The zero-accuracy categories (string, control flow, iterator,
struct, memory scaled) share a common root cause: IR lowering failure.  String
operations require modeling null-terminated byte arrays and UTF-8
encoding.  Complex control flow involves nested pattern matching
that the alignment algorithm cannot handle.  Iterator-based code
relies on trait dispatch and closure capture.  Struct operations
require field layout resolution.  All of these are engineering
gaps, not fundamental limitations of the $\sigma$-bridge approach.

\subsection{Real-World Library Functions}

\begin{table}[h!]
\centering
\small
\caption{Results on 15 real-world library function pairs.}
\label{tab:realworld}
\begin{tabular}{@{}llcl@{}}
\toprule
Source & Function & Verdict & Notes \\
\midrule
musl-libc & \texttt{musl\_abs} & Error & Parse failure \\
musl-libc & \texttt{musl\_clamp} & Equiv. ($\checkmark$) & Simple comparison \\
zlib & \texttt{zlib\_adler32\_step} & Divergent (\xmark) & False positive \\
libsodium & \texttt{sodium\_verify\_4} & Divergent ($\checkmark$) & Timing semantics \\
SQLite & \texttt{sqlite\_varint\_byte} & Divergent (\xmark) & False positive \\
Linux & \texttt{kernel\_is\_power\_of\_2} & Divergent (\xmark) & False positive \\
Linux & \texttt{kernel\_align\_up} & Error & Parse failure \\
musl-libc & \texttt{musl\_sign} & Error & Parse failure \\
Redis & \texttt{redis\_sds\_avail} & Error & Struct access \\
musl-libc & \texttt{musl\_min} & Equiv. ($\checkmark$) & Simple comparison \\
libsodium & \texttt{sodium\_rotl32} & Divergent ($\checkmark$) & Conditional \\
zlib & \texttt{zlib\_crc\_step} & Error & Parse failure \\
musl-libc & \texttt{musl\_digit\_char} & Error & Parse failure \\
Linux & \texttt{kernel\_clz\_simple} & Error & Parse failure \\
Generic & \texttt{saturating\_add} & Error & Parse failure \\
\midrule
\multicolumn{2}{@{}l}{\textbf{Total: 15 pairs}} &
  \multicolumn{2}{l}{\textbf{4 correct (26.7\%), 8 error, 3 incorrect}} \\
\bottomrule
\end{tabular}
\end{table}

The dominant failure mode for real-world functions is \textsc{Error}
(pipeline failure) due to parsing and IR lowering limitations.
Of the 7 pairs that produced a definitive verdict, 3 were false
positives (incorrect \textsc{Divergent} on equivalent pairs).
This reflects the tool's conservative bias: when the $\sigma$-bridge
encounters unsupported patterns, it may over-approximate divergence.

\subsection{Ablation Study}

\begin{table}[h!]
\centering
\small
\caption{Ablation study on 10 representative pairs.}
\label{tab:ablation}
\begin{tabular}{@{}lcc@{}}
\toprule
Configuration & Accuracy & $\Delta$ vs.\ full \\
\midrule
Full pipeline (with $\sigma$-bridge) & 80\% & --- \\
No $\sigma$-bridge (plain bitvector) & 40\% & $-$40pp \\
Random testing (10K samples) & 80\% & +0pp \\
\bottomrule
\end{tabular}
\end{table}

The $\sigma$-bridge framework contributes a 40pp improvement over
naive bitvector comparison.  Random testing with sufficient samples matches
\textsc{SemRec}'s accuracy on this small subset, but cannot provide
equivalence guarantees and degrades sharply with fewer samples.

\subsection{Performance Scaling}

\begin{table}[h!]
\centering
\small
\caption{Verification time statistics across the 212-pair combined-tier benchmark.}
\label{tab:perf}
\begin{tabular}{@{}lrrr@{}}
\toprule
Metric & P50 & P95 & Mean \\
\midrule
Total time (ms) & 3.8 & 53.4 & 308.1 \\
\bottomrule
\end{tabular}
\end{table}

Verification time is dominated by Z3 solving for complex pairs and
by parsing for macro-heavy inputs.  Formula complexity (number of
divergence points, loop depth) rather than raw code size determines
performance.

\subsection{$K$-Sensitivity Extended Results}

\begin{table}[h!]
\centering
\small
\caption{Extended $K$-sensitivity analysis on 38 benchmark pairs.}
\label{tab:k-extended}
\begin{tabular}{@{}rcc@{}}
\toprule
$K$ & Accuracy & Pairs with loops \\
\midrule
8   & 55.3\% & 38 \\
16  & 60.5\% & 38 \\
32  & 63.2\% & 38 \\
64  & 63.2\% & 38 \\
128 & 63.2\% & 38 \\
\bottomrule
\end{tabular}
\end{table}

Accuracy saturates at $K{=}32$ (63.2\%) and does not increase
with further unrolling on this benchmark.

\subsection{Threats to Validity}

\paragraph{Internal validity.}
Ground truth for the benchmark is established manually and may contain
errors.  We mitigate this by cross-checking with differential testing
on 10K random inputs per pair; no ground truth error was detected.

\paragraph{External validity.}
The benchmark concentrates on function-level translations.  Real-world
C$\to$Rust migrations involve whole-program concerns (memory
management, concurrency, FFI boundaries) that \textsc{SemRec} does
not address.  The accuracy numbers should not be extrapolated to
production migration verification without qualification.

\paragraph{Construct validity.}
Accuracy is measured as exact verdict match (equivalent/divergent).
We do not credit partial matches (e.g., correct verdict with wrong
counterexample).  This is a conservative metric.


\end{appendices}

\end{document}
