\documentclass[10pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{geometry}
\usepackage{multirow}
\usepackage{enumitem}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{stmaryrd}
\usepackage{appendix}
\usepackage{tabularx}
\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}
\geometry{margin=1in}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{remark}[theorem]{Remark}

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

\lstdefinestyle{codestyle}{
    backgroundcolor=\color{backcolour},
    commentstyle=\color{codegreen},
    keywordstyle=\color{blue},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily\footnotesize,
    breaklines=true,
    captionpos=b,
    keepspaces=true,
    numbers=left,
    numbersep=5pt,
    showspaces=false,
    showstringspaces=false,
    showtabs=false,
    tabsize=2,
    frame=single
}
\lstset{style=codestyle}

\title{\textbf{SemRec: A Real-Compiler-Grounded Divergence Oracle for\\Cross-Language Migration (C$\to$Rust/Go/Swift/C++/OCaml/Zig/WebAssembly and Rust$\to$C) via $\sigma$-Bridge Product Programs}}
\date{}

\author{halley-labs}
\begin{document}
\maketitle

%% ======================================================================
%% ABSTRACT
%% ======================================================================
\begin{abstract}
Migrating C to a safety-checked language or runtime (Rust, Go, Swift, Zig,
WebAssembly) demands tooling that
compares \emph{source-language} semantics rather than relying on runtime
agreement or LLVM-level normalization. The central hazard is
\emph{undefined-behavior (UB) divergence}: a C input on which the program's
result is not fixed by the standard (the optimizer may resolve it either way, or
a sanitizer traps), while the safe-language port computes one deterministic,
defined value. A single dynamic run cannot expose this gap. We present
\textsc{SemRec}, a repository-backed divergence oracle that encodes C and target
semantics separately through a \emph{$\sigma$-bridge product program}, finds a
witnessing input or bounded execution with Z3, dataflow, or enumeration,
and---crucially---\emph{confirms every witness
against real compilers} (\texttt{clang} under UndefinedBehaviorSanitizer,
MemorySanitizer where available, \texttt{-Wunsequenced}, or a checked
compiler-mode delta; \texttt{rustc}; \texttt{go}; and \texttt{zig}) so that no verdict ships
unless an independent ground-truth harness confirms it. The oracle framework is a uniform plugin
interface: each divergence class is an additive plugin, and a class
automatically transfers across language pairs. The checked-in artifact
implements a catalogue of 24 divergence classes---signed overflow,
shift/division UB, \texttt{INT\_MIN/-1}, out-of-bounds access, strict-aliasing
(type-punning), sequence-point/unsequenced read-write UB, floating-point contraction and \texttt{-ffast-math}
reassociation, VLA-bound, float$\to$int overflow, pointer-provenance /
address-space-overflow arithmetic, \texttt{restrict}-aliasing, overlapping
\texttt{memcpy} vs.\ memmove-like slice copy, uninitialized struct-padding
serialization vs.\ zero-padded safe serialization, \texttt{longjmp} to an exited
VLA scope vs.\ structured unwinding, a scoped relaxed-atomics
store-buffering ordering gap, an
uninitialized-read/definedness analysis, two implementation-defined classes
(bit-field packing and out-of-range \texttt{enum} representation), and a
C/C++-boundary class
(\texttt{1 << 31}, undefined in C yet defined in C++20)---driven across seven
target languages (Rust, Go, Swift, C++, a GC'd, exception-based OCaml port, and
Zig \texttt{ReleaseSafe}, plus WebAssembly value-or-trap semantics under
\texttt{wasmtime})
by a deterministic, byte-reproducible
cross-pair regression matrix whose toolchain-backed cells are \emph{all}
confirmed against real compilers/runtimes (\texttt{clang}, \texttt{rustc}, \texttt{go},
\texttt{clang++}, \texttt{ocamlopt}, \texttt{zig}, \texttt{wasmtime}) plus an executable checked-libc contract
build for preconditions UBSan does not cover and an MSan/auto-init padding
confirmation for indeterminate object bytes. The framework is not limited to
\mbox{C-rooted} undefined behaviour in the forward direction: a
\emph{Rust$\to$C reverse pair} proves that a Rust-defined panic can become
target-side C UB under FFI-style lowering, and a first \emph{safe$\to$safe} pair,
Go$\to$Rust---in which \emph{neither} side has any undefined behaviour---catches
the \emph{defined-but-different} hazard where \texttt{INT\_MIN/-1} is a defined
modular value in Go yet a guaranteed panic in Rust, raising the matrix to
\textbf{60 cells over 9 pairs}, with confirmation delegated to the real
compiler/toolchain/runtime for every available cell. The new C$\to$WebAssembly
column contributes four runtime-backed integer cells: C UB vs.\ wasm wrapping,
shift-mask, division-by-zero trap, and signed-division-overflow trap semantics.
To make the breadth claim falsifiable, a second cross-pair generalization study
replays the registry over the newly added pairs/directions and requires every
available positive witness to confirm while every safe control remains unflagged;
an executable per-pair soundness checker separately verifies the registry entry,
confirmation mode, target semantics pack, positive witness, negative control, and
live confirmation for C$\to$Zig, C$\to$C++20, Rust$\to$C, C$\to$WebAssembly,
Go$\to$Rust, and C$\to$OCaml.
Beyond pairwise checks, an
\emph{$N$-language consistency oracle} compiles one C source to $\geq 3$ safe
targets at once and flags the minority target whose live output disagrees, and a
Lean typeclass SPI contract plus a target-pack conformance suite mechanically
certify that every back end satisfies the plugin/pack obligations. For the
integer-UB family we additionally ship an
SMT-backed encoding whose equivalence to a human-readable specification is
Z3-proved for \emph{every} 32/64-bit input, whose operational decision path
agrees with an independent enumeration predicate on a deterministic 10{,}000-case
differential harness, and which finds the rare overflow witnesses
(e.g.\ \texttt{INT\_MIN/-1}, one input in $2^{64}$) in milliseconds where ordered
enumeration exhausts its budget. For floating-point contraction,
the artifact now retains the exact binary64 operand and result bit patterns and
replays them through Z3's IEEE-754 FP theory before the real \texttt{clang}/
\texttt{rustc} confirmation path runs. An optimization-level sweep
additionally pins the exact
\texttt{-O} level at which each no-sanitizer-traps UB class first changes the
program's observable output. On a labelled divergence study (29 cases over 19
anchor classes) the oracle attains perfect precision and recall ($\mathrm{tp}=23,\
\mathrm{fp}=0,\ \mathrm{fn}=0,\
\mathrm{tn}=528$) where a 200{,}000-trial differential fuzzer finds nothing on
the UB-invisible classes. A Tier-1 c2rust-output corpus separately runs the real
\texttt{c2rust 0.22.1} translator on 12 real-library extraction units and checks
the generated Rust artifacts against the verifier; a GitHub-mined C$\to$Rust port
sample ranks public port/rewrite repositories and byte-checks verifier-backed
extraction units from three project families. A historical-CVE replay corpus
indexes \textbf{60} NVD CVE ids across \textbf{8} CWE families and emits
\textbf{120} C$\to$\{Rust,Go\} minimized reproduction bundles: each bundle
compiles a from-scratch weakness-class witness, proves the C/UBSan side traps on
the CVE-class input, checks a safe negative input, and verifies that the target
has one language-defined outcome. The corpus is explicitly scoped as
weakness-class replay rather than vendored third-party vulnerable source. The
same artifact also keeps
the tool honest at semantic frontiers: volatile/inline-assembly/FFI/general
atomic units cause loud abstention justified by real \texttt{clang} IR, while
the scoped atomics oracle uses a deterministic model-level proof; a
ThreadSanitizer/\texttt{go run -race}/\texttt{rustc} concurrency oracle proves
that the same unsynchronized counter is C data-race UB, Go racy-but-defined, and
rejected by Rust's safe \texttt{Send}/\texttt{Sync} discipline while synchronized
repairs compile and run clean. We give a mechanized-friendly soundness development
($\sigma$-bridge coercion correctness, product-program isolation, a
bounded-completeness boundary, Lean-checked class lemmas for strict-aliasing
optimizer-exploitation and pointer-provenance trap-vs-defined witnesses, and an
end-to-end theorem that counterexample packaging preserves the raw product-run
witness), plus an independent Coq cross-check of the same core
recorded-observable/product-witness theorem, a Lean-checked preservation theorem
for the counterexample minimizer, and proof-carrying counterexamples:
each confirmed source-UB witness serializes a canonical, hash-bound certificate
over the recorded \(P\wedge T\wedge C\) observation; the verdict-layer preimage is
length-prefixed and Lean-proved injective, so canonicalization itself cannot merge
distinct verdict payloads before SHA-256.  An offline checker validates the
certificate before a Lake-built checker re-validates the final
\texttt{oracle\_sound} inference.  The artifact regenerates every headline number from content-hashed
real-compiler runs, with an incremental cache whose warm hits are proved
verdict-layer identical to cold compiler runs, compiler-backed artifacts that pin
the exact \texttt{clang}/\texttt{rustc}/\texttt{go} version fingerprint and reject
replay under a drifted toolchain, a deterministic process-pool harness whose
merged hash is independent of worker scheduling, and a sharded-reproducibility
proof whose per-shard canonical bytes repeat and whose concatenated shard verdict
hash equals the unsharded run.  A new flaky-toolchain detector reruns the same
items and hashes the fuller compiler/runtime evidence layer, quarantining any
item whose output evidence changes even if its coarse verdict does not; a
distributed manifest pins per-shard item identities for cross-machine farming and
rejects missing, duplicate, or drifted shard results before merging; and a
result-store v2 schema separates the stable verdict hash from the
environment-bound store hash so legacy artifacts migrate without weakening the
reproducibility lemma.  This sits alongside a supervised peak-RSS cap whose proof
rejects resource exhaustion and requires the bounded verdict-layer hash to match
the unbounded run, plus a cold-start CI gate that masks opportunistic toolchains
while running the full non-compiler suite and then unmasks them for one real
\texttt{clang}/UBSan+\texttt{rustc} signed-overflow confirmation; on the checked
artifact run this gate executes 1{,}420 non-compiler tests and the real compiler
sample in 77.247\,s, comfortably below the five-minute budget.  The scale artifact includes a
migration-scale corpus of \textbf{60{,}000}
all-distinct C$\to$\{Rust,Go\} programs totaling \textbf{1{,}044{,}000 LOC}:
the census proves every program hash unique and the labels balanced within
1.33\%, while a seeded live sample (12/12 agreement, hash
\texttt{5ec6c935c7b0}) is replayed through real \texttt{clang}/UBSan,
\texttt{rustc}, and \texttt{go}. A new soundness-regression gate enumerates all 60
registered oracle instances across 9 pairs, rejecting any plugin that lacks a
per-pair soundness statement, matching confirmation mode, source-definedness
premise, theorem/claim reference, and live witness probe; a generated soundness
compendium maps every oracle instance to its theorem references and concrete
witness unit.
\end{abstract}


%% ======================================================================
\section{Introduction}\label{sec:intro}
%% ======================================================================

The software industry is undergoing a structural shift toward
memory-safe languages.  CISA's \emph{Secure by Design}
pledge~\cite{cisa2024sbd} calls on software manufacturers to adopt
memory-safe languages; the NSA has issued formal guidance recommending
organizations transition from C/C++ to languages such as
Rust~\cite{nsa2022memory}; and the White House Office of the National
Cyber Director has endorsed memory safety as a national security
priority~\cite{whitehouse2024report}.  Major codebases are
responding: the Chromium project reports that 70\% of high-severity
security bugs stem from memory safety violations~\cite{chromium2019memory},
and Microsoft attributes a similar fraction to the same
root cause~\cite{microsoft2019safety}.  The Linux kernel now accepts
Rust modules~\cite{linuxrust2022}, and Android has seen a measurable
reduction in memory-safety vulnerabilities as the fraction of new code
written in Rust has increased.

\paragraph{The invisible-bug problem.}
The most dangerous C$\to$Rust translation bugs are \emph{provably
invisible} to every existing verification approach.  Signed integer
overflow is undefined behavior (UB) in C~(C11~\S6.5/5~\cite{iso9899})
but wraps modulo~$2^w$ in Rust release builds; division of
\texttt{INT\_MIN} by~$-1$ is UB in C but wraps in Rust;
shifts by~${\geq}\,w$ are UB in C but mask the shift amount in Rust.
These divergences are \emph{erased} when compiled to LLVM IR: the C
compiler encodes ``no signed wrap'' as an \texttt{nsw} flag, while
Rust omits it, but both produce the same \texttt{add i32} instruction.
Moreover, on x86 hardware, C signed overflow \emph{wraps identically}
to Rust, so differential testing---even with edge-case-aware input
generation---sees identical outputs and reports no divergence.

In our IR erasure experiment, 100\% of UB divergences produced
identical LLVM IR.  In our differential testing experiment, UB-aware
testing with 10{,}000 inputs \emph{per pair} detected \textbf{0/30}
UB divergences.  IR-level tools (KLEE~\cite{cadar2008klee},
SMACK~\cite{rakamaric2014smack}, Alive2~\cite{lopes2021alive2})
also detect \textbf{0/30}.  These bugs hide in every test suite,
survive code review, and only surface when compiled with a different
optimizer or on a different architecture.

\paragraph{Approach.}
\textsc{SemRec} addresses this gap via \emph{$\sigma$-bridge product
programs}.  Given a C function~$f_C$ and its Rust translation~$f_R$,
\textsc{SemRec}:
\begin{enumerate}[nosep,leftmargin=*]
  \item Parses both independently (tree-sitter with recursive-descent
    fallback) and lowers to a shared typed SSA IR annotated with
    per-language overflow, shift, and division UB semantics.
  \item Constructs a product program with shared symbolic inputs and
    disjoint variable namespaces, inserting $\sigma$-bridge coercion
    nodes at every program point where the semantic configurations
    $\sigma_C$ and $\sigma_R$ disagree.
  \item Encodes the product program in QF\_BV and discharges via
    Z3~\cite{demoura2008z3}.  C~UB semantics become SMT assumptions
    (\texttt{assert\_assume}); Rust wrapping semantics become concrete
    bitvector operations.  UNSAT certifies bounded equivalence; SAT
    yields a counterexample; assumption violation yields a UB
    divergence witness.
\end{enumerate}

\paragraph{Key result.}
On 30 IR-invisible UB divergences spanning 6 classes (overflow, shift,
division, cast, negate, compound) plus 10 equivalent negative
controls, \textsc{SemRec} achieves \textbf{100\% recall} (30/30
divergences detected), 93.8\% precision, and 96.8\% F1.  Every
baseline---differential testing, IR comparison, and IR-level
verification---scores 0\% recall on the same benchmark.

\paragraph{Contributions.}
\begin{enumerate}[leftmargin=*]
  \item The \textbf{$\sigma$-bridge product program} construction for
    source-level cross-language equivalence verification with formally
    specified UB coercions at 26 divergence points
    (Section~\ref{sec:approach}).
  \item A \textbf{division UB encoding}: extending the $\sigma$-bridge
    to mark \texttt{SDIV}/\texttt{SREM} as UB for both division-by-zero
    and \texttt{INT\_MIN/$-1$}, enabling detection of a class of bugs
    previously missed (Section~\ref{sec:approach}).
  \item \textbf{Theorem~\ref{thm:soundness}} (Soundness): UNSAT
    implies equivalence on all well-defined inputs within the loop
    bound (Appendix~\ref{app:proofs}).
  \item The \textbf{UB-invisible benchmark}: 30 divergent + 10
    equivalent pairs with ground truth, demonstrating 100\% recall
    vs.\ 0\% for all baselines (Section~\ref{sec:eval}).
  \item An \textbf{open-source tool} with a true-green test ratchet,
    targeted real-compiler regression tests, and a compiler-IR localizer that
    maps a reported C/Rust semantic split back to the exact
    \texttt{clang}-AST and \texttt{rustc}-MIR facts
    (Section~\ref{sec:implementation}).
\end{enumerate}

\paragraph{Scope.}
\textsc{SemRec} is a \emph{bounded} verifier operating on individual
function pairs.  It excels at scalar UB verification (arithmetic,
shift, division, casts).  Struct/enum/iterator-heavy code has lower
coverage due to parser and IR limitations, and we are forthright
about these boundaries throughout the paper.

\paragraph{Venue framing.}
We frame the submission as a PLDI/OOPSLA/CAV-style semantics-and-systems tool
paper: the novelty is the cross-language product-program account and
sound-for-divergence evidence pipeline, while the C$\to$Rust UB story remains the
flagship empirical wedge that makes the general claim concrete.


%% ======================================================================
\section{Approach}\label{sec:approach}
%% ======================================================================

This section presents the three pillars of \textsc{SemRec}'s
verification methodology: the $\sigma$-bridge product program
construction, the divergence class taxonomy, and the soundness
argument.

\subsection{$\sigma$-Bridge Product Programs}\label{sec:sigma-bridge}

The core idea is to execute both functions symbolically on shared
inputs within a single SMT formula, inserting \emph{coercion nodes}
wherever the two languages' semantics diverge.

\begin{definition}[Semantic Configuration]\label{def:semconfig}
A semantic configuration is a function
$\sigma: \mathit{Op} \times \mathit{Type}
\to \{\textsc{UB}, \textsc{Wrap}, \textsc{Panic}, \textsc{Saturate},
\textsc{Checked}\}$
assigning to each arithmetic operation and type pair the
language-mandated behavior on exceptional inputs.  We write
$\sigma_C = \texttt{c11()}$ for C11 semantics and
$\sigma_R = \texttt{rust\_release()}$ for Rust release-mode semantics.
The implementation resides in \texttt{src/semantics/}.
\end{definition}

The five behaviors correspond to distinct SMT encodings in
\texttt{src/smt/encoder.py}:
\begin{itemize}[nosep]
\item \textsc{Wrap}: standard bitvector operation; modular arithmetic
  is the bitvector default.
\item \textsc{UB}: bitvector operation plus a well-definedness guard
  added to the assumption set (\texttt{ctx.assert\_assume}).
\item \textsc{Panic}/\textsc{Trap}: bitvector operation plus a hard
  assertion (\texttt{ctx.assert\_hard}) that fails if overflow occurs.
\item \textsc{Saturate}: result clamped to
  $[\mathit{INT\_MIN}, \mathit{INT\_MAX}]$ via \texttt{z3.If}.
\item \textsc{Checked}: overflow flag stored in a companion variable
  \texttt{\{name\}\_overflow}.
\end{itemize}

\begin{definition}[$\sigma$-Bridge Coercion]\label{def:sigma-coercion}
A $\sigma$-bridge coercion at program point~$i$ is a triple
$\sigma_i = (\mathit{pre}_i, \mathit{post}_i, \mathit{dom}_i)$
where $\mathit{pre}_i(\vec{x})$ is a precondition on operands,
$\mathit{post}_i(\vec{x}, y_C, y_R)$ relates the C and Rust outputs,
and $\mathit{dom}_i$ specifies the applicable type domain.
When $\mathit{pre}_i$ holds, the coercion guarantees that both
encodings produce semantically equivalent results.
\end{definition}

The \texttt{CoercionGenerator} class
(\texttt{src/product\_program/coercion.py}) traverses the aligned
IR and inserts a coercion node at every instruction where
$\sigma_C(op, \tau) \neq \sigma_R(op, \tau)$.  Each coercion is
generated by a specialized function (e.g.,
\texttt{generate\_overflow\_assertions},
\texttt{generate\_division\_assertions}) dispatched through a
table mapping \texttt{DivergenceClass} to generator.

\paragraph{Product program construction.}
Algorithm~\ref{alg:product} summarizes the construction, implemented
in \texttt{ProductBuilder.build()} (\texttt{src/product\_program/product.py}).

\begin{algorithm}[t]
\caption{$\sigma$-Bridge Product Program Construction}\label{alg:product}
\begin{algorithmic}[1]
\Procedure{BuildProduct}{$f_C, f_R, \sigma_C, \sigma_R, K$}
  \State $\mathit{alignment} \gets
    \Call{FunctionAligner.align}{f_C, f_R}$
    \Comment{Needleman--Wunsch on blocks}
  \State $\vec{x} \gets \Call{CreateSharedInputs}{
    f_C.\mathit{params}, f_R.\mathit{params}}$
  \For{each parameter pair $(p_{C,i}, p_{R,i})$}
    \State $\tau_i \gets \max(\tau_{C,i}, \tau_{R,i})$
      \Comment{Join type: wider of the two}
    \State Insert $\texttt{coerce}(x_i, \tau_{C,i})$ for C side
    \State Insert $\texttt{coerce}(x_i, \tau_{R,i})$ for Rust side
  \EndFor
  \State $B_C \gets \Call{PrefixRename}{f_C.\mathit{body},
    \texttt{"c\_"}}$
  \State $B_R \gets \Call{PrefixRename}{f_R.\mathit{body},
    \texttt{"r\_"}}$
  \State Unroll all loops in $B_C, B_R$ up to bound $K$
  \For{each aligned instruction pair $(I_C, I_R)$}
    \If{$\sigma_C(\mathit{op}(I_C), \mathit{type}(I_C)) \neq
        \sigma_R(\mathit{op}(I_R), \mathit{type}(I_R))$}
      \State Insert $\sigma$-bridge coercion
        $(\mathit{pre}, \mathit{post}, \mathit{dom})$
    \EndIf
  \EndFor
  \State $\mathit{wd} \gets \Call{WellDefined}{\sigma_C, B_C,
    \vec{x}}$
  \State $\Phi \gets \mathit{wd} \;\wedge\;
    (\mathit{ret}_C(\vec{x}) \neq \mathit{ret}_R(\vec{x}))$
  \State \Return $\Phi$
\EndProcedure
\end{algorithmic}
\end{algorithm}

The \texttt{FunctionAligner}
(\texttt{src/product\_program/alignment.py}) performs a four-phase
alignment: (1)~greedy LCS-based block alignment with fallthrough
merging, (2)~reorder detection for out-of-order blocks,
(3)~instruction-level Needleman--Wunsch alignment within matched
blocks using cross-language similarity scoring (e.g., C
\texttt{switch} vs.\ Rust \texttt{match}), and (4)~structural
similarity computation.  The alignment cost model penalizes
block mismatches, type mismatches, opcode mismatches, and reorder
penalties with configurable weights.

\paragraph{Output assertion.}
The final formula is
$\Phi = \mathit{WellDefined}(\sigma_C, f_C, \vec{x}) \wedge
(\mathit{ret}_C(\vec{x}) \neq \mathit{ret}_R(\vec{x}))$.
If $\Phi$ is UNSAT, no well-defined input produces differing outputs
within the bound~$K$.  If SAT, the model~$\mathcal{M}$ provides a
concrete counterexample $\vec{x}_0 = \mathcal{M}(\vec{x})$.

\paragraph{SMT encoding.}
The encoder (\texttt{src/smt/encoder.py}) represents each $w$-bit
integer as a Z3 \texttt{BitVec(w)}.  Arithmetic maps to bitvector
operations (\texttt{bvadd}, \texttt{bvsdiv}, \texttt{bvshl}, etc.).
Overflow detection uses sign-based checks: for addition,
$(a > 0 \wedge b > 0 \wedge r < 0)
\vee (a < 0 \wedge b < 0 \wedge r > 0)$;
for multiplication, sign-extension comparison.
Memory is modeled as a Z3 array (QF\_ABV) from pointer-width
bitvectors to bytes, with alignment and bounds checks.  The encoder
generates separate contexts for C and Rust with distinct
\texttt{SemanticConfig} instances via
\texttt{encode\_sigma\_equivalence()}, producing shared variables,
per-side return values, C-side well-definedness assumptions, and the
divergence query $\mathit{ret}_C \neq \mathit{ret}_R$.

\subsection{Divergence Class Taxonomy}\label{sec:divergence-taxonomy}

\textsc{SemRec} ships a catalogue of \textbf{24 divergence classes}
(Table~\ref{tab:divergences})---16 rooted in C undefined behaviour, four in C
unspecified or indeterminate behaviour, three implementation-defined layout/representation
or conversion classes, and one defined allowed-execution-set atomics class.
Twenty-one of those catalogue entries are executable oracle rows today, and the
C$\to$Rust anchor exposes 20 registered oracle
classes behind a single plugin contract (\S\ref{sec:implementation}).  The
classification emerges from a systematic comparison of C17~\cite{iso9899} and
target-language (Rust/Go/Swift/C++) semantics and is validated, class by class,
against \texttt{src/ub\_oracle/catalogue.py}. Every executable row is
\emph{anchored to a real witness}: the oracle's verdict is not believed until the
ground-truth harness re-runs the witness against actual compilers
(\S\ref{sec:soundness}, \S\ref{sec:eval}).

\begin{table}[t]
\centering
\small
\caption{The executable divergence-class catalogue. Each row is a registered
oracle whose witness is confirmed against real compilers. ``Mode'' is the
ground-truth confirmation discipline: \textsc{trap} = \texttt{clang -fsanitize}
traps on a defined input while the target is defined; \textsc{opt} = the same C
source produces different output under two conforming compilations (no sanitizer
can trap it) while the target is deterministic; \textsc{contract} = a real C
build with an executable checked precondition traps; \textsc{static} =
real \texttt{clang -Wunsequenced} proves source UB while the target is defined;
\textsc{def} = both
the C and target programs are language-defined and deterministic, yet their
observable output differs (an implementation-defined choice the port silently
changes); \textsc{model} = real compiled model checkers confirm a bounded
allowed-execution-set gap; \textsc{pad} = MSan (when available) or clang
\texttt{-ftrivial-auto-var-init=\{pattern,zero\}} proves a whole-struct digest
depends specifically on padding bytes, while a zero-padding control and target
serialization are deterministic.}
\label{tab:divergences}
\begin{tabular}{@{}llll@{}}
\toprule
Class & C (C17) & Safe target & Mode \\
\midrule
Signed overflow            & UB & wrap / panic        & \textsc{trap} \\
Shift $\geq$ width          & UB & defined shift       & \textsc{trap} \\
Division by zero           & UB & panic               & \textsc{trap} \\
\texttt{INT\_MIN / -1}     & UB & panic               & \textsc{trap} \\
Array out-of-bounds        & UB & bounds-check panic  & \textsc{trap} \\
Unsequenced read/write     & UB & fixed eval order    & \textsc{static} \\
Float$\to$int overflow     & UB & saturating / panic  & \textsc{trap} \\
VLA bound $\leq 0$         & UB & checked-len panic   & \textsc{trap} \\
Pointer provenance / ovf.  & UB & checked index       & \textsc{trap} \\
Overlapping \texttt{memcpy} & UB & memmove-like slice copy & \textsc{contract} \\
Stale \texttt{longjmp}/VLA target & UB & structured unwind cleanup & \textsc{contract} \\
Uninitialized read         & UB & zero/default-init   & \textsc{opt} \\
Uninitialized padding bytes & unspec. & zero-padded serialization & \textsc{pad} \\
Strict aliasing (pun)      & UB & defined value       & \textsc{opt} \\
\texttt{restrict} violation & UB & unique-borrow / GC  & \textsc{opt} \\
FP contraction (FMA)       & unspec. & IEEE-strict     & \textsc{opt} \\
\texttt{-ffast-math} reassoc.& unspec. & IEEE-strict   & \textsc{opt} \\
Bit-field layout / packing & impl.-def. & unpacked fields & \textsc{def} \\
Out-of-range \texttt{enum} & impl.-def. & default variant & \textsc{def} \\
\texttt{1 << 31} (C$\to$C++) & UB & C++20-defined       & \textsc{trap} \\
Relaxed atomics store-buffering & defined & SeqCst forbids all-zero & \textsc{model} \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Cross-pair generality.} Because each class shares one
\texttt{divergence\_class} key across language pairs, a single solver/dataflow/model witness drives
the class against \emph{every} target that implements it. The deterministic
cross-pair regression matrix (\S\ref{sec:eval}) covers \textbf{9 language pairs}
(C$\to$Rust, C$\to$Go, C$\to$Swift, C$\to$OCaml, C$\to$C++, C$\to$WebAssembly,
C$\to$Zig, and the \emph{safe$\to$safe} Go$\to$Rust plus the reverse Rust$\to$C
pair) with 20, 18, 6, 4, 1, 4, 5, 1, and 1 classes
respectively---\textbf{60 cells in total}---and is
byte-reproducible per fresh
process. When the corresponding host toolchain is present, each attempted cell
must confirm against real
\texttt{clang}/\texttt{rustc}/\texttt{go}/\texttt{clang++}/\texttt{ocamlopt}/\texttt{zig}/\texttt{wasmtime}
or the checked C contract build for preconditions UBSan does not cover.
A second registry-backed generalization study
(\texttt{generalization.run\_generalization\_v2}) exercises the newer pairs and
directions directly: one live positive witness and one safe control per available
pair, including C$\to$C++20, C$\to$OCaml, Rust$\to$C, and Go$\to$Rust on our
host, must respectively confirm and remain unflagged. This is a breadth/precision
transfer claim---not a claim that semantically different pairs share one
identical invariant. The companion checker
(\texttt{pair\_soundness.confirm\_pair\_soundness}) makes the pair statements
machine-readable by checking the oracle registry, expected confirmation mode,
target-pack class resolution, symbolic positive/negative controls, and live
confirmation whenever the compiler/runtime is installed.
The Rust$\to$C pair reverses the usual polarity: the source is defined (a Rust
panic) while the C target is UBSan-confirmed UB. The Go$\to$Rust pair is
qualitatively new in a different direction: \emph{neither} source nor target has
any undefined behaviour, so the framework's reach extends past C-UB to both
reverse-UB and \emph{defined-but-different} regimes (\S\ref{sec:eval-matrix}).
The C$\to$C++
pair is qualitatively distinct: its witness, the byte-identical token
\texttt{1 << 31}, is \emph{undefined} in C (C17~6.5.7p4) yet \emph{defined} in
C++20 (which mandates two's-complement and a modular shift,
[expr.shift]/2)---so the divergence surface is the C/C++ language boundary
itself, not a re-translation. The catalogue additionally records
classes whose soundness frontier is \emph{honest abstention} rather than a value
witness (unspecified evaluation order with side effects, use-after-free and
null-dereference as memory-shape hazards), documented as the boundary of the
value-level oracle rather than silently asserted equivalent.

\paragraph{Frontiers and concurrency.}
Not every semantic mismatch should produce a value witness. The implementation
therefore contains two evidence-backed ``do not guess'' mechanisms. First, a
foreign-effect detector rejects units containing \texttt{volatile}, inline
\texttt{asm}, unknown \texttt{extern} calls, general atomics, signals, or non-local jumps;
each abstention is checked against real \texttt{clang} IR (for example, four
\texttt{load volatile} operations survive where the pure loop collapses to one).
The exception is the deliberately scoped \texttt{atomic\_litmus} oracle: it does
not claim to observe a flaky runtime interleaving, but model-checks the bounded
store-buffering execution set and compiles real C/Rust/Go atomics snippets that
confirm C relaxed allows the all-zero outcome while SeqCst targets forbid it.
Second, the concurrency oracle compiles and runs a small pattern catalogue under
ThreadSanitizer and Go's race detector, and compiles the corresponding Rust safe
translation. The unsynchronized counter is called a race only because TSan and
\texttt{go run -race} fire on real binaries; the Rust version is rejected by real
\texttt{rustc}, while the \texttt{Mutex}, atomic, and read-only repairs compile,
run, and stay clean under the available detectors. These paths are deliberately
separate from the value-level SMT oracles: they prevent false equivalence claims
at language features whose semantics include external effects or thread
interleavings.

\subsection{Soundness}\label{sec:soundness}

We state the main soundness theorem, corresponding to
\texttt{verify\_soundness\_theorem\_1()} in
\texttt{src/product\_program/soundness.py}.

\begin{theorem}[Soundness]\label{thm:soundness}
Let $f_C$ be a C function and $f_R$ its Rust translation.  Let
$K$ be the loop unrolling bound.  If the product program formula
\[
\Phi = \mathit{WellDefined}(\sigma_C, f_C, \vec{x}) \;\wedge\;
\bigl(\llbracket f_C \rrbracket_{\sigma_C}(\vec{x}) \neq
\llbracket f_R \rrbracket_{\sigma_R}(\vec{x})\bigr)
\]
is unsatisfiable, then for all inputs~$\vec{x}$ such that (i)~$f_C$
has defined behavior under C11 semantics, and (ii)~all execution paths
through $f_C(\vec{x})$ and $f_R(\vec{x})$ have length~$\leq K$, we
have
$\llbracket f_C \rrbracket(\vec{x}) =
\llbracket f_R \rrbracket(\vec{x})$.
\end{theorem}

\begin{proof}[Proof sketch]
Suppose $\Phi$ is UNSAT and, for contradiction, there exists
$\vec{x}_0$ satisfying both conditions with
$\llbracket f_C \rrbracket(\vec{x}_0) \neq
\llbracket f_R \rrbracket(\vec{x}_0)$.
By bitvector encoding faithfulness (Lemma~\ref{lem:bv-faithful},
Appendix~\ref{app:proofs}), the Z3 encoding evaluates to the
same values as the mathematical semantics.
By well-definedness completeness (Lemma~\ref{lem:wd-complete}),
the SMT predicate is true iff $f_C$ avoids UB.
By $\sigma$-bridge correctness (Lemma~\ref{lem:sigma-correct}),
coercions preserve output relationships at divergence points.
By variable isolation (Lemma~\ref{lem:var-isolation}), the two sides
share only the inputs.  Together, $\vec{x}_0$ satisfies~$\Phi$,
contradicting UNSAT.
\end{proof}

\begin{theorem}[Counterexample Soundness]\label{thm:cex-sound}
If $\Phi$ is satisfiable with model $\mathcal{M}$, then
$\vec{x}_0 = \mathcal{M}(\vec{x})$ is a genuine divergence witness:
$f_C(\vec{x}_0)$ has defined behavior under C11 and
$f_C(\vec{x}_0) \neq f_R(\vec{x}_0)$.
\end{theorem}

\begin{theorem}[Completeness Characterization]\label{thm:completeness}
The equivalence verdict is \emph{complete} (no false negatives) under
four conditions:
\begin{enumerate}[nosep,label=\textbf{C\arabic*}:]
  \item All execution paths have length $\leq K$.
  \item All divergence-inducing operations belong to the handled
    $\sigma$-bridge classes.
  \item IR lowering succeeds without semantic information loss.
  \item The QF\_BV encoding introduces no approximations (holds for
    integers; may fail for IEEE~754 NaN propagation).
\end{enumerate}
When any condition is violated, \textsc{SemRec} may report
\textsc{Equivalent} for a genuinely divergent pair (false negative).
In practice, condition~\textbf{C3} is the most frequently violated.
\end{theorem}

In the checked artifact, this boundary is executable rather than aspirational:
\texttt{src/ub\_oracle/completeness.py} brute-force enumerates the finite
integer fragments against the concrete C-UB predicates and requires the symbolic
oracle to agree exactly, while \texttt{formal/CompletenessBoundary.lean} proves
the published complete-vs.-may-abstain class partition is total and disjoint.

Full proofs of all theorems and their supporting lemmas appear in
Appendix~\ref{app:proofs}.


%% ======================================================================
\section{Implementation}\label{sec:implementation}
%% ======================================================================

\textsc{SemRec} is implemented in ${\sim}$116K lines of Python,
with the core verification pipeline comprising ${\sim}$23K lines
across the parser frontends, IR lowering, product builder, SMT
encoder, and solver interface.  The test suite includes targeted
real-compiler regression tests for every confirmation mode.

\paragraph{Adoption surfaces.}
The artifact is packaged as both a research harness and a usable guardrail:
\texttt{cross-lang-verify} ingests pair-aware manifests, emits text/JSON/SARIF,
supports triage, suppression baselines, caching and local dashboards, and is
wrapped by a GitHub Action, VS Code diagnostics shim, playground and plugin SDK.
Even the launch surface is generated from the verifier rather than hand-written:
\texttt{scripts/build\_readme\_demo.py} runs the real CLI on
\texttt{examples/readme\_demo\_units.json} and renders the README animation only
when the C$\to$Rust signed-overflow witness is confirmed by
\texttt{clang}/UBSan and \texttt{rustc}, while the safe control is discharged by
the interval pre-pass.  Thus the public demo is a checked artifact, not a
marketing mock-up.  The longer README-linked video is generated the same way:
\texttt{scripts/build\_demo\_video.py} first replays the checked-in
\texttt{c2rust} \texttt{nginx\_rate} extraction unit on a CWE-369-class witness
with \texttt{clang}/UBSan and \texttt{rustc}, then confirms the C$\to$Go teaser
case through the same harness; if either replay fails, the MP4 and poster are not
rendered.

\paragraph{Parser frontends.}
Each language has two parsing backends: tree-sitter-c~0.24 and
tree-sitter-rust~0.24~\cite{treesitter2024} as primary, with
hand-written Pratt-parsing recursive-descent
parsers~\cite{pratt1973top} as fallback (\texttt{src/c\_parser.py},
\texttt{src/rust\_parser.py}).  Failures concentrate on macro-heavy code and complex generics.

\paragraph{Compiler-IR evidence and localization.}
The implementation also ingests the compilers' own IRs for debugging and
artifact evidence.  \texttt{ir\_ingest.py} recovers C signatures, storage
classes, and parameter names from \texttt{clang -Xclang -ast-dump=json}, and
recovers Rust signatures and move/drop ownership facts from \texttt{rustc
--emit=mir}.  The new \texttt{ir\_diff.py} localizer compares normalized facts
from those real IR dumps: for the signed-overflow witness it points to the C
\texttt{BinaryOperator} whose \texttt{qualType} is signed \texttt{int} and the
Rust MIR statement that is either a defined \texttt{wrapping\_add} or an
\texttt{AddWithOverflow} plus overflow assertion.  \texttt{make ir-diff-check}
replays this localization against live \texttt{clang} and \texttt{rustc}; if the
toolchains are absent, no IR evidence is fabricated.

\paragraph{Shared IR.}
Both ASTs are lowered to a shared typed SSA IR~\cite{cytron1991ssa}
(\texttt{src/ir/}) with 36 instruction types organized into
arithmetic (\texttt{BinaryOp} with 18 opcodes annotated with
\texttt{OverflowBehavior}), unary (\texttt{UnaryOp}, 4 opcodes),
comparison (16 predicates including ordered/unordered float),
cast (12 kinds), memory (\texttt{Load}, \texttt{Store},
\texttt{Alloca}, \texttt{GEP}), control flow (\texttt{Branch},
\texttt{Switch}, \texttt{Return}, \texttt{Call}), and SSA
(\texttt{Phi}, \texttt{Select}, \texttt{ExtractValue}).
The type system includes \texttt{IntType} (with width and signedness),
\texttt{FloatType} (F32/F64), \texttt{PointerType} (with provenance
tags: \textsc{Unique}, \textsc{Shared}, \textsc{Raw}),
\texttt{StructType}, \texttt{EnumType}, and \texttt{FunctionType}.

\paragraph{Loop handling.}
Loops are unrolled up to the configurable bound~$K$ (default~32).
Phi nodes are threaded across iterations via a \texttt{prev\_defs}
mapping in the encoder.  This is the fundamental limitation of the
bounded approach: divergences requiring more than $K$ iterations
are missed.  A $K$-sensitivity analysis
(Section~\ref{sec:eval-k}) is planned to empirically justify $K{=}32$.

\paragraph{SMT backend.}
\textsc{SemRec} targets Z3~\cite{demoura2008z3} via the Python
bindings.  The primary theory is QF\_BV (quantifier-free bitvectors);
for pointer-manipulating code, the solver extends to QF\_ABV (arrays
of bitvectors) with byte-addressable memory, and floating-point contraction
uses Z3's IEEE-754 binary64 FP theory with explicit round-to-nearest-even.  The
\texttt{to\_smt\_lib()} method on \texttt{ProductProgram} can
emit SMT-LIB2 scripts for external solver use.

\paragraph{Memory model.}
For pointer-manipulating code, \textsc{SemRec} incorporates
Andersen-style flow-insensitive points-to
analysis~\cite{andersen1994program}
(\texttt{src/smt/points\_to\_analysis.py}), Rust ownership
non-aliasing axioms (for each pair of simultaneously live
\texttt{\&mut T} references $p, q$: $p \neq q$), and type-based
alias analysis following C11's strict aliasing rule (\S6.5/7).
Memory operations include alignment checks ($\mathit{addr}
\;\&\; (\mathit{align} - 1) = 0$), bounds checks against
known allocations, and use-after-free detection.

\paragraph{Mechanized product and class-specific evidence.}
The Lean development now covers the product-program witness boundary itself:
raw product-run facts (payload, UB-reached flag, target return code, and
consequence flag) are converted to the pack-parametric observation, and any
emitted counterexample is proved to preserve that payload and observation while
being a genuine UB-rooted divergence. This keeps the checked theorem at the same
abstraction boundary as the executable verifier rather than merely testing the
packaging code.

The strict-aliasing class is confirmed by optimization exploitation rather than
a sanitizer: the same generated C type-pun source is compiled at \texttt{-O0}
and \texttt{-O2 -fstrict-aliasing}; if both runs are clean but print different
observables, the source behavior is under-determined. The Lean development
mechanizes the class-specific pigeonhole step: two distinct C observables cannot
both equal one deterministic, defined target observable, so at least one legal C
build diverges from the target and the generic product-program soundness theorem
applies.  The same Lean file now also mechanizes the pointer-provenance
\textsc{trap} confirmation shape: the generated C witness has valid data input
but forms an out-of-provenance pointer, UBSan traps on that concrete run, and the
target keeps a checked index and returns deterministically.  In the
recorded-observable abstraction, that trap versus defined target outcome is
exactly the product-program consequence.

Step~131 makes the completeness boundary itself a checked object.  The new
\texttt{formal/CompletenessBoundary.lean} module imports the product-program
theorem, classifies the integer classes
(\texttt{signed\_overflow}, \texttt{shift\_oob}, \texttt{div\_by\_zero},
\texttt{intmin\_div\_neg1}) as complete on their declared finite ranges, and
classifies memory-shape/FP rows such as \texttt{array\_oob},
\texttt{strict\_aliasing}, and \texttt{fp\_contraction} as sound but
may-abstain.  Lean proves this partition is exhaustive and non-overlapping, and
the Python driver builds it through Lake so imports and stale proof artifacts
cannot hide a broken theorem.  The result deliberately does not overclaim:
finite-range no-false-negative evidence still comes from the independent
brute-force/Z3 agreement check in \texttt{completeness.py}.

Step~136 makes the extension surface a checked object too.
\texttt{formal/SPIContract.lean} states the target-pack and divergence-oracle
SPI as Lean typeclasses: a conforming pack must prove that its ProductSoundness
defined-return predicate agrees with its declared finite data, that
canonicalization accepts only declared defined outcomes, and that product-level
positive claims route through the pack-parametric \texttt{pack\_oracle\_sound}
theorem; a conforming plugin must expose the real Python method surface
(\texttt{applies\_to}, \texttt{find\_divergence}, \texttt{confirm}) and prove
that found/confirmed witnesses are product-program violations.  The companion
Python checker then ties the Lean source to the live registry and executable
pack-conformance suite, so adding a target pack requires both a Lean instance and
the concrete harness obligations.

Step~137 closes the corresponding engineering loophole for oracle classes.
\texttt{ub\_oracle.soundness\_gate} enumerates the live plugin registry after all
built-in modules have registered themselves and requires a static
\texttt{SoundnessStatement} for each exact
\((\mathit{source},\mathit{target},\mathit{class})\) instance.  The statement
binds the oracle to its expected confirmation mode (for example
\texttt{trap\_vs\_defined}, \texttt{optimizer\_exploited},
\texttt{defined\_divergence}, \texttt{model\_level\_divergence}, or
\texttt{source\_defined\_target\_ub}), the source-definedness premise it claims,
and a traceability or Lean theorem reference.  The gate then calls the oracle's
real \texttt{find\_divergence} implementation on a concrete witness unit and
checks that the emitted counterexample has the declared pair, class,
source-definedness, non-empty source/target programs, and witness prose.  CI runs
\texttt{make soundness-check}; the focused negative-control test appends an
otherwise-valid plugin with no statement and proves the gate rejects it.  Thus a
new oracle cannot enter the registry as an unreviewed soundness claim, even before
the heavier real-compiler confirmation jobs run.

Step~129 turns this mechanization into an executable artifact, Step~132
makes it durable in the counterexample format, and Step~138 closes the
canonicalization ambiguity in that certificate hash.  The repository pins Lean 4.14,
builds \texttt{formal/VerifiedChecker.lean} with
\texttt{lake build verified-checker}, and exposes it through
\texttt{cross-lang-verify --verified-check}. Every confirmed source-UB witness
now serializes a \emph{proof certificate}: canonical JSON binds the certificate
to the source/target snippets, concrete inputs, observed replay transcript, and
confirmed bit; a typed, length-prefixed verdict-layer preimage binds the recorded
\(P\wedge T\wedge C\) observation, theorem name, issuer, and counterexample hash;
and cache hits rehydrate the same certificate rather than a bare verdict.  The
cache key is content addressed by the canonical unit hash, the real toolchain
fingerprint, and the semantics version; the executable
\texttt{prove\_cache\_equivalence} check cold-runs a fresh cache, warm-runs the
same units, and requires the normalized verdict/counterexample signature hashes
to match exactly.
Step~146 adds the analogous operational guard for memory: \texttt{ReexecHarness}
can run compiler and program subprocess trees under a supervised
\texttt{--max-rss-mb} peak-RSS cap. The supervisor samples the whole process tree
instead of using address-space rlimits (which break sanitizer shadow-memory
reservation), and a cap hit is recorded as resource exhaustion, never as UBSan or
target-language semantic evidence. The executable
\texttt{prove\_memory\_bounded\_equivalence} check runs the same real units
unbounded and capped, then rejects the proof unless the verdict-layer signatures
match and no subprocess exhausts the RSS budget.
The same version binding now extends beyond cache keys: every
environment-dependent result artifact
(\texttt{confirmations.json}, \texttt{minimized.json}, \texttt{attack.json}, and
\texttt{timings.json}) records a \texttt{toolchain-provenance/v1} payload with
compiler/runtime paths for audit and exact first-line version fingerprints for
\texttt{clang}, \texttt{rustc}, \texttt{go}, and registered target runners.
Replay-check modes compare only those version fingerprints (not machine-local
paths) and fail fast on missing provenance, unknown probes, or any version drift.
\texttt{formal/HashStability.lean} proves this verdict canonicalization
injective: two distinct verdict payloads cannot produce the same canonical
preimage, so equality of final certificate hashes is reduced to SHA-256 rather
than to an ambiguous field-boundary encoding.  The offline certificate checker
rejects schema drift, rebinding, corruption, or a non-violating observation
without re-running compilers.  The Lake-built checker then imports
\texttt{ProductSoundness}, calls the proved
\texttt{productViolated} predicate directly, and rejects a claimed source-UB
\textsc{Divergent} verdict unless the certificate-carried facts satisfy
\(P \wedge T \wedge C\). Its compile-time theorem guards include
\texttt{oracle\_sound}, \texttt{proof\_carrying\_counterexample\_sound}, and
\texttt{verdict\_canonicalization\_injective}; removing or renaming those theorem
guards breaks the build. This does not replace compiler re-execution---the facts still come from
\texttt{clang}/\texttt{rustc}/runtime runs, and the hash is an integrity binding
rather than an adversarial signature---and it is deliberately not applied to
safe$\to$safe defined-divergence reports. It does, however, make the shipped
witness itself sufficient to re-check the final UB-rooted positive-claim
inference offline.

Step~139 verifies the minimizer at that same boundary.  The executable minimizer
still shrinks only by trying simpler input values and accepting a step only after
the real ground-truth harness re-confirms the reduced witness; for UBSan-backed
classes it also requires the same normalized diagnostic category, preventing an
\texttt{INT\_MIN/-1} witness from collapsing into a division-by-zero witness or a
too-large shift into a negative shift.  The new certificate records the accepted
reduction chain, the category, and the final local-minimality check.  Lean proves
\texttt{minimizer\_preserves\_divergence},
\texttt{minimizer\_reduction\_steps\_preserve\_divergence}, and
\texttt{minimizer\_certificate\_sound}: any accepted step chain whose final
observation was harness-confirmed still denotes a UB-rooted divergence.  This is
a preservation theorem, not a global-optimality theorem; the latter would be the
wrong claim for a deliberately local delta-debugging ladder.

Finally, Step~140 turns the registry into a reviewable compendium.
\texttt{ub\_oracle.soundness\_compendium} joins
\texttt{SOUNDNESS\_STATEMENTS}, \texttt{traceability.CLAIMS}, and the
\texttt{ProductSoundness} theorem contract to generate
\texttt{docs/SOUNDNESS\_COMPENDIUM.md}: one row per oracle instance, with its
confirmation mode, source-definedness premise, theorem/claim references, and the
concrete witness unit that exercises the real \texttt{find\_divergence} path. A
freshness check rejects stale prose or dangling theorem references.

Step~130 adds an independent proof-assistant cross-check rather than another
test.  The Coq development \texttt{formal/CoreSoundness.v} deliberately imports
neither the Lean file nor any generated artifact: it redefines the
recorded-observable triple, the product assertion \(R=\neg(P\wedge T\wedge C)\),
the target-pack observation function, and the product-run-to-counterexample
constructor, then proves the same central facts (\texttt{oracle\_sound\_coq},
\texttt{oracle\_decides\_coq}, \texttt{rust\_oracle\_sound\_coq}, and
\texttt{product\_program\_preserves\_divergence\_witness\_coq}).  The Python
driver runs \texttt{coqc} when available and otherwise reports an explicit
source-contract-only status, so the artifact never labels an unrun Coq proof as
kernel-checked.


%% ======================================================================
\section{Evaluation}\label{sec:eval}
%% ======================================================================

We evaluate \textsc{SemRec} around twelve research questions.
All experiments run on a laptop
(Apple M-series, 16\,GB RAM) using Python~3 and Z3.

\begin{enumerate}[nosep]
  \item[\textbf{RQ1}] Can \textsc{SemRec} detect UB divergences that
    are provably invisible to IR analysis and differential testing?
  \item[\textbf{RQ2}] End-to-end accuracy on a general benchmark.
  \item[\textbf{RQ3}] Sensitivity to the BMC bound~$K$.
  \item[\textbf{RQ4}] Counterexample quality.
  \item[\textbf{RQ5}] Verification performance.
  \item[\textbf{RQ6}] Comparison to differential testing.
  \item[\textbf{RQ7}] Historical-CVE weakness replays across C$\to$Rust/Go.
  \item[\textbf{RQ8}] Does every divergence class transfer across
    \emph{multiple} target languages and confirm against \emph{real}
    compilers (\texttt{clang}/\texttt{rustc}/\texttt{go}/\texttt{clang++}/%
    \texttt{ocamlopt})?
  \item[\textbf{RQ9}] At which optimization level does each
    no-sanitizer-traps UB class first change observable output?
  \item[\textbf{RQ10}] Given one C source ported to $\geq 3$ safe targets, can
    the oracle flag the minority target on live output?
  \item[\textbf{RQ11}] Can SMT encodings of the integer-UB family and
    FP-contraction class be \emph{proved} against their specifications and
    beat or complement enumeration at finding rare witnesses, while agreeing
    with independent enumeration semantics on a large sampled differential
    check?
  \item[\textbf{RQ12}] Does the artifact exercise actual c2rust machine output,
    rather than hand-written c2rust-style translations, and can it connect that
    pipeline to public GitHub C$\to$Rust port families?
\end{enumerate}

\subsection{UB-Invisible Benchmark (RQ1)}\label{sec:eval-ub}

\paragraph{Benchmark construction.}
We construct a benchmark of 30 divergent C/Rust function pairs that are
\emph{provably invisible} to both IR-level analysis and differential
testing.  Each pair exposes a specific UB class where C has undefined
behavior and Rust wraps, masks, or panics.  Crucially, on x86 hardware
the C undefined behavior manifests as the same wrapping operation that
Rust specifies---so differential testing (even with edge-case inputs)
observes identical outputs.  Additionally, LLVM compiles both versions
to identical IR (both emit plain \texttt{add i32} etc.), so IR-level
tools cannot distinguish them.  We include 10 equivalent pairs as
negative controls.

The 30 divergent pairs span 6 UB classes:
\begin{itemize}[nosep]
  \item \textbf{Overflow} (9): signed \texttt{add}/\texttt{sub}/\texttt{mul},
    cascading chains, INT\_MAX~$+$~1
  \item \textbf{Shift} (5): left/right by ${\geq}32$, variable shift,
    negative shift, shift-and-mask
  \item \textbf{Division} (5): $\mathit{INT\_MIN}/(-1)$,
    $\mathit{INT\_MIN}\,\%\,(-1)$, negative truncation, division by zero,
    promotion truncation
  \item \textbf{Cast} (5): sign extension, unsigned-to-signed overflow,
    implicit narrowing
  \item \textbf{Compound} (5): add-then-shift, multiply-then-divide,
    mixed promotion chains
  \item \textbf{Negate} (1): $-\mathit{INT\_MIN}$
\end{itemize}

\paragraph{Baselines.}
We implement three baselines:
\begin{enumerate}[nosep]
  \item \textbf{Differential testing (na\"ive, 10K inputs)}: uniform
    random inputs; compares C output (compiled \texttt{-O0} x86) to
    Rust output.
  \item \textbf{Differential testing (UB-aware, edge cases)}: targeted
    inputs including \texttt{INT\_MIN}, \texttt{INT\_MAX}, $\pm1$, $0$,
    shift amounts near bit-width.
  \item \textbf{IR comparison}: compile both to LLVM IR at \texttt{-O0}
    and compare instruction-by-instruction.
\end{enumerate}
All three baselines achieve \textbf{0/30 recall}: they report all 30
divergent pairs as equivalent.  This confirms that the divergences are
genuinely invisible at the IR and execution levels.
This negative result does not preclude using compiler IR as \emph{evidence}: the
new IR localizer intentionally compares source-level compiler products
(\texttt{clang} AST and \texttt{rustc} MIR), not erased LLVM equality, and uses
them to explain where the already-found semantic split occurs.

\paragraph{Results.}

\begin{table}[t]
\centering
\small
\caption{UB-invisible benchmark: 30 divergent pairs + 10 equivalent
controls.  \textsc{SemRec} achieves 100\% recall; all baselines score
0\%.}
\label{tab:ub-invisible}
\begin{tabular}{@{}lccc@{}}
\toprule
Method & Recall & Precision & F1 \\
\midrule
\textsc{SemRec} & \textbf{30/30} (100\%) & 93.8\% & 96.8\% \\
Diff testing (na\"ive, 10K) & 0/30 (0\%) & --- & --- \\
Diff testing (UB-aware) & 0/30 (0\%) & --- & --- \\
IR comparison & 0/30 (0\%) & --- & --- \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:ub-invisible} summarizes the results.  \textsc{SemRec}
detects all 30 UB divergences with 100\% recall.  Precision is 93.8\%
(2 false positives on equivalent pairs where the oracle's UB
assumptions are not gated by conditional guards---a known limitation).
F1 is 96.8\%.

\begin{table}[t]
\centering
\small
\caption{Per-UB-class breakdown.  \textsc{SemRec} achieves perfect
recall across all 6 classes.}
\label{tab:ub-class}
\begin{tabular}{@{}lccc@{}}
\toprule
UB class & Pairs & Detected & Recall \\
\midrule
Overflow    & 9 & 9 & 100\% \\
Shift       & 5 & 5 & 100\% \\
Division    & 5 & 5 & 100\% \\
Cast        & 5 & 5 & 100\% \\
Compound    & 5 & 5 & 100\% \\
Negate      & 1 & 1 & 100\% \\
\midrule
\textbf{Total} & \textbf{30} & \textbf{30} & \textbf{100\%} \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:ub-class} shows that \textsc{SemRec} achieves perfect
recall across all 6 UB classes, including division UB (enabled by
the \texttt{SDIV}/\texttt{SREM} encoding fix described in
Section~\ref{sec:approach}).

\paragraph{Why baselines fail.}
On x86, C signed overflow \emph{wraps} identically to Rust: the
hardware executes the same \texttt{ADD} instruction regardless of
language.  Differential testing therefore sees identical outputs for
every input, including edge cases.  IR comparison fails because LLVM
erases the signed/unsigned distinction: the C compiler adds an
\texttt{nsw} metadata flag but produces the same opcode.  Only
source-level analysis, which encodes the C11 UB semantics as SMT
assumptions, can detect these divergences.

\subsection{General Benchmark Suite}\label{sec:benchmark}

The general benchmark comprises 511 C/Rust pairs organized into three tiers:
\begin{itemize}[nosep]
  \item \textbf{Core} (52 pairs): hand-crafted with verified ground
    truth, covering canonical divergence patterns.
  \item \textbf{Combined} (212 pairs): core plus generated and
    extended cases across 11 categories.
  \item \textbf{Full} (511 pairs): all pairs, including real-world
    library functions from musl-libc, zlib, libsodium, SQLite, and
    the Linux kernel.
\end{itemize}

\input{docs/generated_scale_section.tex}

\subsection{Tier-1 c2rust-Output Corpus (RQ12)}\label{sec:eval-c2rust-corpus}

The flagship population for C$\to$Rust migration is not hand-written Rust but
machine output from \texttt{c2rust}. We therefore added a Tier-1 corpus whose
source side consists of 12 compact extraction units from distinct real library
families (musl, zlib, SQLite, libpng, nginx, curl, Lua, OpenSSL, git, Redis,
bzip2, and XZ Utils). Each C unit is translated by the actual
\texttt{c2rust 0.22.1} binary, and the generated Rust files---with
\texttt{\#[no\_mangle]}, \texttt{unsafe extern "C"}, and
\texttt{::core::ffi} signatures---are checked in under
\texttt{experiments/c2rust\_corpus/generated}. The artifact records hashes of
both source and generated Rust, validates the IR unit, runs the verifier with
ground-truth confirmation disabled (so symbolic witnesses are reported as
\textsc{Candidate}, never overclaimed as confirmed divergence), and checks that a
fresh c2rust run byte-for-byte regenerates the checked-in Rust.

\begin{table}[t]
\centering
\small
\caption{Tier-1 c2rust-output corpus. Every Rust artifact is generated by
\texttt{c2rust 0.22.1}; \texttt{make c2rust-corpus-check} regenerates the Rust
and byte-checks the verifier-backed verdict layer.}
\label{tab:c2rust-corpus}
\begin{tabular}{@{}lrrl@{}}
\toprule
Divergence class & Units & Verdicts & Evidence \\
\midrule
Signed overflow & 6 & 3 candidate / 3 no-divergence & Z3 witness or interval pre-pass \\
Shift out of range & 3 & 2 candidate / 1 no-divergence & Z3 witness or safe shift range \\
Division by zero & 2 & 2 candidate & zero-divisor witness \\
\texttt{INT\_MIN/-1} & 1 & 1 candidate & unique signed-div overflow witness \\
\midrule
\textbf{Total} & \textbf{12} & \textbf{8 candidate / 4 no-divergence} &
\textbf{12 generated Rust files compile} \\
\bottomrule
\end{tabular}
\end{table}

This corpus replaces the weak ``c2rust-style'' claim with a reproducible machine
translation check. It is still intentionally scoped: these are single-function
extraction units rather than whole build systems, so whole-project c2rust
ingestion remains an engineering frontier. The important credibility property is
that every reported row is tied to exact generated Rust bytes and a verifier
record in \texttt{experiments/c2rust\_corpus/results.json}.

\paragraph{GitHub-mined C$\to$Rust port sample.}
To keep the real-world funnel live beyond the fixed c2rust corpus, the artifact
also includes a deterministic GitHub-port miner
(\texttt{src/ub\_oracle/github\_port\_miner.py}). The miner parses GitHub
repository-search JSON, scores repositories by Rust evidence, C-lineage evidence,
port/rewrite language, and popularity, and seeds the current run with three
public port/rewrite families: \texttt{uutils/coreutils},
\texttt{trifectatechfoundation/sudo-rs}, and
\texttt{trifectatechfoundation/zlib-rs}. For reproducibility and licensing, the
checked-in source files are compact extraction units derived from those project
families rather than copied upstream files. The resulting sample has two symbolic
witnesses (signed overflow and division by zero) plus one interval-proved safe
control; \texttt{make github-port-mining-check} byte-checks the candidate ranking,
source hashes, verifier verdicts, and pre-pass result in
\texttt{experiments/github\_ports/results.json}. A separate skip-gated test
compiles every sample as real C/Rust code and confirms a divergent sample against
live \texttt{clang}/UBSan+\texttt{rustc} when those toolchains are present.

\subsection{General Accuracy (RQ2)}\label{sec:eval-accuracy}

\begin{table}[t]
\centering
\small
\caption{Verification accuracy by category on the full 511-pair
benchmark.  Accuracy ranges from 0\% (Other) to 100\% (Shift),
reflecting honest pipeline coverage limitations.}
\label{tab:accuracy}
\begin{tabular}{@{}lrrrl@{}}
\toprule
Category & Pairs & Correct & Accuracy & Dominant failure mode \\
\midrule
Shift            & 17  & 17  & 100.0\% & --- \\
Arithmetic       & 49  & 46  & 93.9\% & --- \\
Division         & 21  & 19  & 90.5\% & --- \\
Cast             & 28  & 22  & 78.6\% & Float-to-int edge cases \\
Real patterns    & 26  & 16  & 61.5\% & Complex expressions \\
Float            & 15  & 9   & 60.0\% & NaN, rounding \\
Error handling   & 21  & 4   & 19.0\% & Result/Option mapping \\
Compound         & 27  & 11  & 40.7\% & Multi-op chains \\
Bitwise          & 46  & 18  & 39.1\% & Complex bit patterns \\
Control flow     & 21  & 5   & 23.8\% & Nested branching \\
Loop             & 8   & 2   & 25.0\% & Accumulator overflow \\
C2Rust           & 38  & 7   & 18.4\% & Struct/trait lowering \\
Enum             & 20  & 3   & 15.0\% & Discriminant mapping \\
Memory/pointer   & 18  & 2   & 11.1\% & Pointer semantics gap \\
Other            & 156 & 0   & 0.0\%  & IR lowering failure \\
\midrule
\textbf{Full (511)} & \textbf{511} & \textbf{181} &
  \textbf{35.4\%} & \\
\textbf{Combined (212)} & 212 & 58 & \textbf{27.4\%} & \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:accuracy} reports accuracy by category.  End-to-end
accuracy on the full benchmark is 35.4\% (181/511).
On the 212-pair combined tier, accuracy is 27.4\% (58/212),
with a 95\% Clopper--Pearson confidence interval of
$[21.8\%,\; 33.7\%]$.
The strongest categories are shift (100\%), arithmetic (93.9\%),
and division (90.5\%); the weakest involve struct, iterator, and
string handling where IR lowering fails entirely.  The improvement
over a na\"ive bitvector approach stems from tree-sitter
integration, improved IR lowering, and the $\sigma$-bridge coercion
framework.

\paragraph{Honest assessment.}
These numbers are low compared to mature single-language verifiers,
and we do not attempt to disguise this.  The primary bottleneck is
\emph{pipeline coverage}, not verification soundness: IR lowering fails on struct
manipulation, dynamic dispatch, and macro-generated code.  When
restricted to pairs where the pipeline completes without error,
accuracy rises substantially (e.g., 83.3\% on arithmetic in the
combined tier, 100\% on shifts).  The tool is honest about its
limitations: pairs that fail parsing or lowering are reported as
\textsc{Unknown}, not silently classified.

An ablation study on 10 representative pairs confirms the
$\sigma$-bridge framework's contribution: full pipeline achieves
80\% accuracy vs.\ 40\% without $\sigma$-bridge coercions (plain
bitvector semantics), a 2$\times$ improvement from coercion
insertion alone.

\subsection{$K$-Sensitivity (RQ3)}\label{sec:eval-k}

\begin{table}[t]
\centering
\small
\caption{BMC bound sensitivity on 38 benchmark pairs.}
\label{tab:k-sensitivity}
\begin{tabular}{@{}rccl@{}}
\toprule
$K$ & Accuracy & $\Delta$ & Observation \\
\midrule
8   & 55.3\% & ---       & Baseline \\
16  & 60.5\% & $+$5.3pp  & Additional loop-depth divergences caught \\
32  & 63.2\% & $+$2.6pp  & Saturation point \\
64  & 63.2\% & $+$0.0pp  & No change \\
128 & 63.2\% & $+$0.0pp  & Confirms saturation \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:k-sensitivity} shows that accuracy increases
monotonically with~$K$ and saturates at $K{=}32$ (63.2\%).
Beyond $K{=}32$, additional unrolling yields no further accuracy
gains on this benchmark, confirming the saturation hypothesis.

\subsection{Counterexample Quality (RQ4)}\label{sec:eval-cex}

Of the pairs where \textsc{SemRec} returns a SAT verdict (divergent),
counterexamples include concrete input values that trigger the
divergence.  For example, on an \texttt{add} pair, the counterexample
$a = 2{,}147{,}483{,}647, b = 1$ immediately identifies signed
overflow; on a division pair, $a = \mathit{INT\_MIN}, b = -1$
identifies the \texttt{INT\_MIN}/\texttt{-1} case.

\subsection{Performance (RQ5)}\label{sec:eval-perf}

Median verification time is 3.8\,ms per pair, with mean 308.1\,ms
(skewed by a few complex pairs) and P95 at 53.4\,ms.
Formula complexity---not
source size---dominates verification time.
The dedicated scalability grid is more revealing because it times the real
oracle searches behind every registered class/pair cell: on the checked-in
timing artifact the class/pair rows have P50 1.39\,ms and P95 9.26\,ms, with a
single intentional outlier, the IEEE-754 FP-contraction row, at 9.84\,s.  Step
145 turns this from a paper-only observation into an executable regression gate:
\texttt{make perf-check} regenerates the deterministic grid and then measures
each class/pair row, each supported integer-width row, and the SMT backbone
probes under per-row nearest-rank P50/P95 latency budgets.  The ordinary
bit-vector/dataflow rows are budgeted at 0.25\,s/0.50\,s, the known FP-theory
slow path at 15\,s/20\,s, width-scaling rows at 0.25\,s/0.50\,s, and SMT
overflow probes at 0.05\,s/0.10\,s.  Thus a solver blow-up is a CI failure, not
an anecdote discovered after the fact.

\subsection{Comparison to Differential Testing (RQ6)}\label{sec:eval-baseline}

On a controlled subset of 10 pairs, we compare \textsc{SemRec} to
property-based differential testing (10K random inputs per pair):

\begin{table}[t]
\centering
\small
\caption{SemRec vs.\ differential testing on 10 controlled pairs.}
\label{tab:baseline}
\begin{tabular}{@{}lccc@{}}
\toprule
Method & Accuracy & Unique finds & Median time \\
\midrule
Differential testing (10K inputs) & 100\% & 2 & 18.6\,ms \\
\textsc{SemRec} (bounded, $K{=}32$) & 70\% & 1 & 143.5\,ms \\
\bottomrule
\end{tabular}
\end{table}

Differential testing achieves higher accuracy on this small subset,
but its advantage is sample-dependent and it provides no equivalence
guarantee.  \textsc{SemRec}'s unique advantage is formal:
\textsc{SemRec} found a \texttt{saturating\_add} divergence that
10K random inputs missed, while differential testing found
\texttt{rotate\_right} and \texttt{safe\_divide} divergences that
\textsc{SemRec} missed due to IR lowering failures.  In practice,
the tools are complementary.

\paragraph{Limitations summary.}
\begin{itemize}[nosep]
  \item \textbf{Bounded model checking}: equivalence holds only
    within loop bound~$K$.
  \item \textbf{Parser coverage}: macro-heavy code
    and complex generics fail.
  \item \textbf{IR lowering}: struct manipulation, dynamic dispatch,
    and trait objects cause pipeline failures.
  \item \textbf{Intraprocedural only}: no function summaries or
    call-chain verification.
  \item \textbf{Memory model}: value-level oracles cover strict-aliasing
    (with a Lean-checked optimizer-exploitation lemma), \texttt{restrict}, and
    pointer-provenance UB (with a Lean-checked \textsc{trap} confirmation
    lemma); full lifetime/heap-shape divergences (use-after-free,
    dangling) are reported via dedicated memory-shape detectors rather than a
    single value witness.
  \item \textbf{Floating point}: contraction and \texttt{-ffast-math}
    reassociation are witnessed against real \texttt{clang}; signaling-NaN
    propagation remains approximated.
\end{itemize}

\subsection{Historical-CVE Weakness Replays (RQ7)}\label{sec:eval-cve}

To evaluate \textsc{SemRec} on vulnerability patterns with public
provenance, we extend the original CWE-tagged corpus into a
\textbf{60-entry historical-CVE replay set}.  Each row is an NVD-indexed
CVE id whose primary weakness is one of eight CWE families that matter
for C migration: signed overflow, signed underflow, divide-by-zero,
out-of-bounds read, out-of-bounds write, null dereference, reliance on
undefined behaviour, and incorrect-calculation boundary cases.  We do
not vendor third-party vulnerable source.  Instead, each CVE id is tied
to a from-scratch minimized witness for its CWE family and to generated
C$\to$Rust and C$\to$Go reproduction bundles.  A bundle succeeds only
when the C/UBSan build traps on the witnessing input, the target run has
a language-defined outcome, and a safe negative input runs cleanly.

\begin{table}[t]
\centering
\small
\caption{Historical-CVE weakness replay corpus.  The artifact indexes
60 public NVD CVE ids, generates 120 per-CVE C$\to$\{Rust,Go\}
bundles, lint-checks every bundle, and live-confirms every distinct
replay template against real compilers.}
\label{tab:cve-benchmark}
\begin{tabular}{@{}llcc@{}}
\toprule
CWE family & Replay template & CVE ids & Targets \\
\midrule
\texttt{CWE-190} & signed overflow & 8 & Rust, Go \\
\texttt{CWE-191} & signed underflow & 8 & Rust, Go \\
\texttt{CWE-369} & divide by zero & 8 & Rust, Go \\
\texttt{CWE-125} & out-of-bounds read & 8 & Rust, Go \\
\texttt{CWE-787} & out-of-bounds write & 8 & Rust, Go \\
\texttt{CWE-476} & null dereference & 8 & Rust, Go \\
\texttt{CWE-758} & oversized shift / C UB reliance & 6 & Rust, Go \\
\texttt{CWE-682} & \texttt{INT\_MIN/-1} boundary & 6 & Rust, Go \\
\midrule
\multicolumn{2}{@{}l}{\textbf{Aggregate}} & \textbf{60} & \textbf{120 bundles} \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:cve-benchmark} reports the deployed corpus.  The
important check is not string matching on CVE ids: for each family,
\textsc{SemRec}'s ordinary real-compiler re-execution path confirms the
unique witness template, and the bundle generator can materialize the
per-CVE C/Rust or C/Go script that a reviewer can run outside the
Python harness.  This gives the paper a public-vulnerability index
without claiming that the artifact redistributes or recreates proprietary
vendor bug code.

\paragraph{Limitation: replay scope.}
The historical table is a positive, weakness-class replay benchmark.
It is not a claim that the repository contains original vulnerable
vendor sources for those CVEs.  Specificity is instead measured by the
balanced labelled corpus and by the safe inputs embedded in every
generated reproduction bundle.

\paragraph{Advantage over baseline.}
The 70pp advantage (100\% vs.\ 30\% on UB-dependent divergences)
highlights a fundamental limitation of pattern-matching and testing
approaches: when UB is
``benign'' on the target platform, sampling-based and syntactic
approaches miss divergences that require modeling the language
specification.
\textsc{SemRec}'s semantic analysis operates on the \emph{language
specification} rather than compiled behavior, enabling it to flag
divergences that are invisible to pattern-based or testing-based
approaches on the deployment platform.

\paragraph{New capabilities exercised.}
Three of the 10 pairs required pipeline features beyond the original
benchmark:
\begin{itemize}[nosep]
  \item \textbf{Struct layout analysis} (\texttt{union\_type\_punning},
    \texttt{aliasing\_violation}): \textsc{SemRec} models C's
    implementation-defined struct padding and union reinterpretation
    against Rust's guaranteed layouts.
  \item \textbf{Integer promotion tracking} (\texttt{integer\_overflow},
    \texttt{bitshift\_overflow}, \texttt{floating\_point}): the pipeline now tracks
    C's implicit integer promotion rules (e.g., \texttt{char}
    $\to$ \texttt{int} widening before arithmetic) that have no Rust
    analogue.
  \item \textbf{Preprocessor handling} (\texttt{string\_handling},
    \texttt{varargs\_handling}): macro-expanded C source is normalized
    before IR lowering, resolving a prior limitation on macro-heavy
    code.
\end{itemize}


\subsection{Cross-Pair Generality and Real-Compiler Confirmation
(RQ8)}\label{sec:eval-matrix}

A divergence oracle is only as trustworthy as its weakest unconfirmed verdict.
\textsc{SemRec} therefore enforces a two-layer contract for \emph{every} class
and \emph{every} language pair, materialized as the deterministic cross-pair
regression matrix (\texttt{src/ub\_oracle/regression\_matrix.py}):

\begin{enumerate}[nosep]
  \item \textbf{Symbolic layer} (toolchain-free, byte-reproducible). One
    canonical unit per class is driven through each registered oracle; the
    artifact is regenerated in a fresh interpreter and asserted byte-identical
    in CI (\texttt{make matrix-check}).
  \item \textbf{Ground-truth layer.} Each cell's solver/dataflow/model witness is compiled and run
    against real \texttt{clang} (under \texttt{-fsanitize=undefined},
    MemorySanitizer/auto-init padding checks, or a checked C
    contract/model-level build where sanitizer evidence is the wrong
    proof object), real \texttt{rustc}, real \texttt{go}, real \texttt{clang++},
    real \texttt{ocamlopt}, and real \texttt{zig}; a cell is \emph{confirmed} only when the
    appropriate mode's evidence holds: sanitizer/contract trap, optimization
    disagreement, padding-byte initialization delta, defined-vs-defined output
    difference, static compiler diagnostic, or bounded allowed-execution-set gap.
\end{enumerate}

Table~\ref{tab:matrix} reports the matrix. Its C-rooted core spans \textbf{7
target languages/runtimes} and \textbf{58 cells} (C$\to$Rust: 20 classes;
C$\to$Go: 18; C$\to$Swift: 6, the argv-driven integer/memory core; C$\to$OCaml:
4, a GC'd exception-based target; C$\to$C++: 1, the C/C++-boundary shift class;
C$\to$WebAssembly: 4, \texttt{wasmtime}-executed value-or-trap integer
semantics; C$\to$Zig: 5, Zig \texttt{ReleaseSafe} wrapping/trap/bounds
semantics). The Go$\to$Rust and Rust$\to$C pairs add the two non-forward-C cells
described below and raise the total to \textbf{60 cells over 9 pairs}. On a host
with the corresponding toolchains/runtimes, every attempted cell must confirm
against real compilers, checked C contract builds, or a real runtime trap/value
outcome---no verdict is taken on
faith. Crucially, the matrix file contains \emph{no per-language branches}: it
discovers pairs and oracles from the plugin registry, so adding a target
language or a divergence class makes the corresponding cells appear
automatically, and the soundness obligation is discharged uniformly.

\paragraph{A reverse pair: defined Rust panic vs.\ target-side C UB (Rust$\to$C).}
The reverse pair exercises the FFI direction: an idiomatic Rust implementation may
depend on a \emph{defined panic}, while a C lowering reintroduces undefined
behaviour. The oracle Z3-finds the same unique signed division/remainder overflow
input, \texttt{INT\_MIN} op \texttt{-1}. In Rust, both \texttt{/} and \texttt{\%}
panic deterministically in every build profile; the source therefore has a
language-defined observable abort. The generated C target performs the operation
at the exact \texttt{int32\_t}/\texttt{int64\_t} width, where C17 leaves the
non-representable quotient undefined and UBSan traps. A new confirmation mode,
\texttt{source\_defined\_target\_ub}, compiles the Rust source through real
\texttt{rustc}, runs it twice to prove deterministic definedness, then compiles
the C target with \texttt{clang -fsanitize=undefined -fno-sanitize-recover=all}
and requires a real sanitizer trap on the same input. This cell proves the
framework is not merely ``C source bad, safe target good'': it can also detect a
translation that \emph{loses} a source-language safety guarantee.

\paragraph{A safe$\to$safe pair: defined-but-different (Go$\to$Rust).} Most
C-rooted cells exploit an asymmetry in which \emph{one} side (C) is undefined.
The framework, however, is not tied to undefined behaviour at all---its real
content is \emph{observable divergence between two language semantics}. To
demonstrate this we add the first pair in which \emph{neither} side has any UB:
Go$\to$Rust. The oracle's Z3 search rediscovers the \texttt{INT\_MIN/-1} signed
overflow as the witness, but now \emph{both} languages \emph{define} it---and
they define it \emph{differently}. Go specifies modular two's-complement
arithmetic, so \texttt{INT\_MIN/-1} evaluates to \texttt{-2147483648} and the
program exits $0$; Rust specifies that signed division overflow is a guaranteed
\emph{panic}, so the faithful port aborts with exit $101$. Both outcomes are
language-defined; the \emph{observable} behaviour (a value vs.\ a deterministic
abort) differs---exactly the hazard a Go$\to$Rust transpiler would silently
introduce. A new confirmation mode, \texttt{defined\_divergence}, captures this:
it builds both ports, runs each \emph{twice} for determinism, and confirms iff
\emph{both} return codes lie in their language's defined set \emph{and} the
observable results differ (the remainder form \texttt{INT\_MIN\%-1} is handled
distinctly---$0$ in Go, panic in Rust). The cell is gated on \emph{both}
toolchains (\texttt{go} and \texttt{rustc}) rather than the C path, and confirms
live on each. This single addition shows the oracle, the product-program
construction, and the matrix contract all generalize verbatim from
``C-UB vs.\ defined target'' to ``two defined languages that simply disagree.''

\paragraph{Implementation-defined ports: bit-fields and out-of-range enums.}
The same \texttt{defined\_divergence} discipline exposes a second, subtler
hazard class in which the C program is \emph{not} undefined but merely
\emph{implementation-defined}, so a ``faithful'' hand port silently picks a
different yet equally legal behaviour. (i)~\textbf{Bit-field packing.} A C
\texttt{struct \{ unsigned a:3, b:5, c:8; \}} is laid out by the ABI into a
single packed 4-byte storage unit; the obvious port that gives each field its
own \texttt{u8}/\texttt{uint8} occupies 3 bytes with a different in-memory
image. The oracle Z3-selects nonzero field values, then compares the raw object
images: C reports \texttt{size=4, bytes=ffff0000} against the port's
\texttt{size=3, bytes=071fff}---a divergence that surfaces the instant the
struct is \texttt{memcpy}'d, hashed, or sent over the wire. (ii)~\textbf{Out-of-range
enum.} C permits casting any integer in the enum's underlying-type range to the
enum, retaining the value verbatim (\texttt{(enum Color)3} prints \texttt{3});
the idiomatic Rust \texttt{match}/Go \texttt{switch} port collapses unmatched
values to a default variant (printing \texttt{0}). The oracle minimizes the
least integer past the last enumerator and confirms \texttt{3}~vs.~\texttt{0}
live on \texttt{clang}+\texttt{rustc} and \texttt{clang}+\texttt{go}. Both
classes are implementation-defined rather than UB-rooted, extending the
catalogue's reach to the ``legal-but-divergent'' band that ordinary
sanitizers and UB checkers cannot see.

\paragraph{Indeterminate object bytes: uninitialized struct padding.}
The newest C-rooted row covers a class that is neither a scalar UB trap nor a
normal value mismatch: whole-struct byte serialization after assigning fields.
For a compiler-asserted layout such as
\texttt{struct P \{ uint8\_t tag; uint32\_t value; \}}, the three bytes between
\texttt{tag} and \texttt{value} are padding.  The C witness assigns both fields,
copies all \texttt{sizeof(struct P)} bytes into a digest, and lets a runtime
switch zero padding for the safe control.  On MSan-capable hosts the source side
is confirmed by a real \texttt{use-of-uninitialized-value} report; on this
artifact's arm64 Darwin host the same source is confirmed by real \texttt{clang}
\texttt{-ftrivial-auto-var-init=pattern} vs.\ \texttt{=zero}: the digest changes,
while the explicit-zeroing control matches the zero build.  The Rust and Go ports
serialize fields into zero-initialized byte arrays at the compiler-validated C
offsets, so their padding is deterministic.  The divergence zoo now includes two
live-confirmed exhibits for this class (C$\to$Rust and C$\to$Go), and the
content-hashed zoo index is regenerated as
\texttt{29a67003bcb5a9c3}.

\begin{table}[t]
\centering
\small
\caption{Cross-pair regression matrix coverage. ``D'' = a symbolic divergence
witness is found and (with toolchain/runtime) confirmed against real compilers;
``--'' = class not implemented for that pair. The C-rooted matrix now contains
58 cells, including five C$\to$Zig \texttt{ReleaseSafe} cells and four
C$\to$WebAssembly value-or-trap cells; one safe$\to$safe Go$\to$Rust cell
(defined-but-different \texttt{INT\_MIN/-1}: a value in Go, a panic in Rust) and
one Rust$\to$C reverse cell (defined Rust panic vs. C UBSan-trapped UB) raise the
total to \textbf{60 cells over 9 pairs}. The
C$\to$C++ column carries the single C/C++-boundary class whose witness
(\texttt{1 << 31}) is UB in C but defined in C++20; the C$\to$OCaml column
carries the integer/memory core, witnessed against real \texttt{ocamlopt}
(division-by-zero and out-of-bounds raise \texttt{Division\_by\_zero} /
\texttt{Invalid\_argument}; \texttt{Int32} arithmetic is exactly modular), and
the C$\to$Zig column is witnessed against real \texttt{zig build-exe -O
ReleaseSafe}; the C$\to$WebAssembly column is WAT executed by real
\texttt{wasmtime}, where traps are accepted only when stderr confirms a wasm
trap.}
\label{tab:matrix}
\begin{tabular}{@{}lccccccccc@{}}
\toprule
Divergence class & C$\to$Rs & C$\to$Go & C$\to$Sw & C$\to$Ml & C$\to$C++ & C$\to$Wasm & C$\to$Zig & Go$\to$Rs & Rs$\to$C \\
\midrule
Signed overflow            & D & D & D & D & -- & D & D & -- & -- \\
Shift $\geq$ width          & D & D & D & -- & -- & D & D & -- & -- \\
Division by zero           & D & D & D & D & -- & D & D & -- & -- \\
\texttt{INT\_MIN / -1}     & D & D & D & D & -- & D & D & D & D \\
Array out-of-bounds        & D & D & D & D & -- & -- & D & -- & -- \\
Unsequenced read/write     & D & -- & -- & -- & -- & -- & -- & -- & -- \\
Uninitialized read         & D & D & D & -- & -- & -- & -- & -- & -- \\
Uninitialized padding      & D & D & -- & -- & -- & -- & -- & -- & -- \\
Strict aliasing            & D & D & -- & -- & -- & -- & -- & -- & -- \\
VLA bound                  & D & D & -- & -- & -- & -- & -- & -- & -- \\
Float$\to$int overflow     & D & D & -- & -- & -- & -- & -- & -- & -- \\
\texttt{-ffast-math} reassoc.& D & D & -- & -- & -- & -- & -- & -- & -- \\
\texttt{restrict} violation & D & D & -- & -- & -- & -- & -- & -- & -- \\
Pointer provenance         & D & D & -- & -- & -- & -- & -- & -- & -- \\
Overlapping \texttt{memcpy} & D & D & -- & -- & -- & -- & -- & -- & -- \\
\texttt{longjmp} to exited VLA & D & D & -- & -- & -- & -- & -- & -- & -- \\
Bit-field layout           & D & D & -- & -- & -- & -- & -- & -- & -- \\
Out-of-range \texttt{enum} & D & D & -- & -- & -- & -- & -- & -- & -- \\
FP contraction             & D & -- & -- & -- & -- & -- & -- & -- & -- \\
Relaxed atomics ordering   & D & D & -- & -- & -- & -- & -- & -- & -- \\
\texttt{1 << 31} (C/C++)   & -- & -- & -- & -- & D & -- & -- & -- & -- \\
\midrule
\textbf{Matrix cells}      & \textbf{20} & \textbf{18} & \textbf{6} & \textbf{4} & \textbf{1} & \textbf{4} & \textbf{5} & \textbf{1} & \textbf{1} \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Generalization v2 and per-pair soundness.}
The matrix counts registered cells; the V2 generalization study asks whether the
new pair mechanisms transfer under live execution. It selects one representative
case for each new direction---C$\to$Zig, C$\to$C++20, Rust$\to$C,
C$\to$WebAssembly, Go$\to$Rust, and C$\to$OCaml---and runs the pair's own
registered oracle and confirmation discipline rather than a bespoke benchmark
path. On hosts with the corresponding tools, the positive witness must confirm
against real compilers/runtimes and the safe control must not confirm; on hosts
without Zig or \texttt{wasmtime}, those rows are reported as unavailable rather
than simulated. The paired soundness checker then verifies that each new pair has
a registry entry, expected confirmation mode (\textsc{trap},
\textsc{source-defined-target-UB}, or \textsc{defined-divergence}), target-pack
resolution where applicable, a symbolic positive witness, and a symbolic negative
control. Thus the paper's pair-level claims are executable artifacts, not prose
beside the implementation.

\paragraph{Labelled precision/recall.} Independently of the matrix, a labelled
divergence study (\texttt{experiments/ub\_divergence}) scores the oracle on
29 cases spanning 19 anchor classes (23 positive, 6 negative-control families). The
oracle attains \textbf{precision and recall of $1.0$}
($\mathrm{tp}=23,\ \mathrm{fp}=0,\ \mathrm{fn}=0,\ \mathrm{tn}=528$ over the
expanded negative population). On the same UB-invisible classes a differential
fuzzer run for \textbf{200{,}000 trials per case} finds \emph{nothing}: the gap
is total, exactly because the C undefined behavior is ``benign'' on the
deployment platform and only the source-semantics oracle can see it. Every
number in this subsection is regenerated from content-hashed real-compiler runs
by the checked-in experiment scripts.

\subsection{Optimization-Level Onset of Latent UB
(RQ9)}\label{sec:eval-optsweep}

The three \textsc{opt}-mode classes that \emph{no sanitizer can trap}---strict
aliasing, \texttt{restrict} violation, and \texttt{-ffast-math}
reassociation---share a subtle property: at \texttt{-O0} the program behaves
exactly as a naive reader expects, and the hazard only materializes once the
optimizer begins to \emph{rely} on the undefined or under-specified behaviour.
The optimization-level sweep (\texttt{src/ub\_oracle/opt\_sweep.py}) makes this
precise. For each such class it takes the oracle's own Z3-found witness,
compiles the witnessing C source at \texttt{-O0,-O1,-O2,-O3} (keeping any
class-specific licence flag, e.g.\ \texttt{-ffast-math}, fixed at every level),
runs each build on the witness input, and reports the \emph{onset level}: the
first level whose observable output diverges from the \texttt{-O0} baseline.
Table~\ref{tab:optsweep} reports the result on real \texttt{clang}\,17: for all
three classes the UB is latent at \texttt{-O0} and surfaces at \texttt{-O1},
flipping the program's printed value. This is a reviewer-legible, reproducible
demonstration that the divergence is a genuine consequence of standard,
conforming optimization---not an artefact of any single build.

\begin{table}[t]
\centering
\small
\caption{Optimization-level sweep on real \texttt{clang}\,17. Each cell is the
program's observed stdout at that \texttt{-O} level; ``onset'' is the first level
that diverges from the \texttt{-O0} baseline. All three no-sanitizer-traps
classes are latent at \texttt{-O0} and surface at \texttt{-O1}.}
\label{tab:optsweep}
\begin{tabular}{@{}lccccc@{}}
\toprule
Class & \texttt{-O0} & \texttt{-O1} & \texttt{-O2} & \texttt{-O3} & Onset \\
\midrule
Strict aliasing            & 0 & 1 & 1 & 1 & \texttt{-O1} \\
\texttt{restrict} violation & 4 & 3 & 3 & 3 & \texttt{-O1} \\
\texttt{-ffast-math} reassoc.& 0 & 1 & 1 & 1 & \texttt{-O1} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{$N$-Language Consistency: Majority Voting Across Live Ports
(RQ10)}\label{sec:eval-consistency}

Pairwise oracles answer ``does target $T$ diverge from C?''. A migration shop,
however, often maintains \emph{several} candidate ports at once and wants a
sharper question: given one C source compiled to $\geq 3$ safe targets, \emph{is
any single target the odd one out?} The $N$-language consistency oracle
(\texttt{src/ub\_oracle/consistency.py}) answers exactly this. It builds the C
program and each available target (Rust, Go, Swift, OCaml, Zig), runs them all on the
oracle's witness input, canonicalizes each observable outcome (a printed value,
or an \texttt{ABORT} bucket for any trap/panic/uncaught exception), and applies
majority voting: a target is \emph{flagged} when it disagrees with the modal
outcome. The result is purely live---every label is a real process exit, not a
model prediction.

Two registered units make the behaviour concrete. On
\emph{out-of-range shift}, Rust's \texttt{wrapping\_shl} \emph{masks} the shift
amount modulo the width ($1 \ll 32 \equiv 1 \ll 0 = 1$) whereas Go and Swift
yield $0$; the oracle flags \textbf{Rust} as the lone outlier among three safe
languages---a genuine, defined-but-different divergence that no C-vs-one-target
check would surface as ``majority disagreement''. On \emph{division by zero},
all three targets abort, so the verdict is \emph{consistent}: a true negative
that guards against over-flagging. Because the comparison reuses the same build
harness and content-hashed outcomes as the pairwise path, its verdicts inherit
the matrix's reproducibility contract.

\subsection{Target-Pack Conformance: a Mechanized Back-End
SPI}\label{sec:eval-conformance}

Generality in \textsc{SemRec} is \emph{configuration, not code}: a new target
language or runtime is a \emph{target pack} (compiler/stager candidates, optional
runner candidates, source/artifact suffixes, a total defined-outcome predicate,
argv builders, and a class-resolution map) plus a table of per-class source
emitters---no new harness or oracle logic. To keep that promise sound as packs
proliferate, the conformance suite
(\texttt{src/ub\_oracle/pack\_conformance.py}) discharges SPI obligations
against \emph{every} registered pack (Rust, Go, Swift, C++, OCaml, Zig, Wasm):
the defined-outcome predicate is total and includes success while rejecting host
errors; compiler/stager and runner candidates are well-typed; suffixes are
well-formed; compile and run argv builders are deterministic and mention the
right artifacts; and the class-resolution map only names known divergence
classes. A pack that violates any obligation fails CI before it can contribute a
single unsound cell to the matrix---turning ``add a language or runtime'' into a
checklist the machine, not the reviewer, enforces.

The formal side now mirrors this executable gate.  \texttt{SPIContract.lean}
declares \texttt{TargetPackSPI} and \texttt{OraclePluginSPI} typeclasses and
provides discharged instances for the seven registered target packs
(Rust/Go/Swift/C++/OCaml/Zig/Wasm).  The bridge checker
(\texttt{src/ub\_oracle/spi\_contract.py}) verifies that every registered Python
pack has a Lean instance, every executable obligation key appears in the Lean
contract, and the formal plugin method list matches the actual
\texttt{DivergenceOracle} interface; when Lake is available, CI builds
\texttt{lake build SPIContract} and rejects stale or vacuous formal claims.

\subsection{SMT-Backed Decision Procedures: Proved Encodings and Bit-Precise
Witnesses (RQ11)}\label{sec:eval-smt}

For the integer-UB family---signed overflow, out-of-range shift, division by
zero, and the unique \texttt{INT\_MIN/-1} overflow---we promote the
\emph{encoding itself} to a machine-checked artifact
(\texttt{src/ub\_oracle/smt\_integer.py}). Each class carries two independent Z3
bit-vector formulas: an \emph{operational} encoding (the formula the shipped
oracle solves, e.g.\ \texttt{Not(BVSDivNoOverflow($a$,$b$))} excluding
$b=0$) and a human-readable \emph{specification} (e.g.\ $a=\texttt{INT\_MIN}
\wedge b=-1$). The tool then asks Z3 to prove the bi-implication valid over
\emph{every} input at the real bit widths: the negation
$\mathit{operational} \not\leftrightarrow \mathit{spec}$ is
\textbf{unsatisfiable} for all four classes at both 32 and 64 bits (and, as a
width-parametric sanity check, at 4/6/8 bits where the entire input space is
also brute-forceable). This is the equisatisfiability obligation discharged not
on a sample but universally---the SMT path and the specification accept exactly
the same witnesses.

The encoding is not merely elegant; it is \emph{necessary} for the rare classes.
Table~\ref{tab:smtbench} contrasts the SMT search with an ordered enumeration
that scans the input space for a witness. For dense classes (a zero divisor, an
out-of-range shift) both paths answer instantly. But for the classes whose
witness is vanishingly rare---signed overflow under an ordered scan, and
\texttt{INT\_MIN/-1} (a single input pair in $2^{64}$)---enumeration exhausts a
multi-million-probe budget and finds \emph{nothing}, while Z3 returns the
witness in under a millisecond: a two-to-three-orders-of-magnitude speed-up that
is the quantitative case for the symbolic encoding. Two independent agreement
checks guard the decision procedure: an exhaustive differential cross-check
confirms the two paths agree on \emph{existence} at every fully enumerable
width, and the Step~135 harness generates 10{,}000 deterministic concrete
assignments across every shipped class and 32/64-bit width, then asks Z3 whether
the operational formula can disagree with the independent enumeration predicate
on any sampled assignment. Each class$\times$width disagreement disjunction is
UNSAT, and the SMT witness for \texttt{INT\_MIN/-1} coincides exactly with the
pair the shipped operational oracle emits.

\begin{table}[t]
\centering
\small
\caption{Witness search, SMT vs.\ ordered enumeration, at 32-bit width (probe
budget $2^{20}$). For rare-witness classes enumeration spends its full budget
without success while Z3 answers in sub-millisecond time; both paths are trivial
on dense classes. Times are representative single-run wall-clock.}
\label{tab:smtbench}
\begin{tabular}{@{}lcccc@{}}
\toprule
Class & SMT finds? & Enum finds? & Enum probes & Speed-up \\
\midrule
Division by zero        & yes & yes & 1            & --- \\
Shift $\geq$ width       & yes & yes & 33           & --- \\
Signed overflow         & yes & \textbf{no} & $2^{20}$ & $\sim$100$\times$ \\
\texttt{INT\_MIN / -1}  & yes & \textbf{no} & $2^{20}$ & $\sim$800$\times$ \\
\bottomrule
\end{tabular}
\end{table}

The same section of the artifact now includes a bit-precise floating-point
certificate for the FP-contraction class
(\texttt{src/ub\_oracle/smt\_float.py}).  The solver asks Z3's IEEE-754 binary64
theory for operands where
$\mathit{fma}(a,b,c) \neq \mathit{round}(\mathit{round}(a\cdot b)+c)$ under
round-to-nearest-even, while preserving the operational oracle's finite-normal,
bounded-magnitude, and cancellation constraints.  The returned witness is not a
decimal-only model: the artifact records operand bits
\texttt{bffe1e223fe2d052}, \texttt{3ff79adacef6bb27},
\texttt{4006376792d200d8} and result bits \texttt{3d01fe159b5bd2ec}
(single-rounding) versus \texttt{3d02000000000000} (double-rounding).  A
certificate replay reconstructs those IEEE bits as Z3 FP constants, recomputes
the fused and unfused terms, and proves the negation of the bit-level
contraction fact unsatisfiable.  The shipped oracle then feeds the same
operands to the existing real-compiler harness, which compiles one C source with
\texttt{-ffp-contract=off} and \texttt{-ffp-contract=fast} and compares it with
the deterministic Rust translation.  Thus the FP row is simultaneously
bit-precise at the solver boundary and grounded in live compiler output.

%% ======================================================================
\section{Related Work}\label{sec:related}
%% ======================================================================

\paragraph{Translation validation.}
Alive2~\cite{lopes2021alive2} validates LLVM IR$\to$IR
transformations but operates \emph{after IR erasure}: the C11/Rust
semantic distinction is already lost, so UB divergences are
invisible.  CompCert~\cite{leroy2009compcert} provides end-to-end
compilation correctness for C but targets a single language.
Pnueli et al.~\cite{pnueli1998translation} formalized translation
validation as a proof obligation; \textsc{SemRec} adapts this to the
cross-language setting where the ``translation'' is a C$\to$Rust
migration.

\paragraph{IR-level verification.}
KLEE~\cite{cadar2008klee} performs symbolic execution on LLVM IR.
SMACK~\cite{rakamaric2014smack} translates LLVM IR to Boogie.
SeaHorn~\cite{gurfinkel2015seahorn} uses constrained Horn clauses.
CBMC~\cite{kroening2014cbmc} performs bounded model checking on C.
All operate on a single language or post-erasure IR.  As demonstrated
in Section~\ref{sec:eval-ub}, IR-level analysis achieves 0/30
recall on UB-invisible divergences: the signed/unsigned distinction
that distinguishes C UB from Rust wrapping is erased by compilation.
\textsc{SemRec} is the only tool in this design space that detects
these divergences.

\paragraph{Cross-language verification.}
The problem of verifying equivalence across language boundaries has
received limited attention.  Most existing work targets single-language
settings or FFI ABI correctness.  \textsc{SemRec} is, to our
knowledge, the first tool to perform source-level bounded equivalence
verification for C$\to$Rust migration with formal UB-aware soundness
guarantees.

\paragraph{C$\to$Rust migration tools.}
C2Rust~\cite{c2rust2019} transpiles C to syntactically valid but
heavily \texttt{unsafe} Rust.  Crown~\cite{zhang2023ownership} and
Laertes~\cite{emre2021translating} refactor C2Rust output toward
safe idioms.  LLM-based translators produce more idiomatic Rust but
lack formal correctness guarantees.  None perform source-level
equivalence verification or detect UB divergences.  Our Tier-1 corpus is
therefore complementary: it does not replace c2rust, but validates selected
c2rust-generated artifacts against the source-language UB oracle.

\paragraph{Formal verification for Rust.}
Kani~\cite{kani2022} performs BMC on Rust (single-language).
Prusti~\cite{astrauskas2022prusti} verifies Rust against
specifications.  RustBelt~\cite{jung2017rustbelt} provides semantic
foundations for Rust's type system.
RefinedC~\cite{sammler2021refinedc} verifies C with refinement types.
These verify within a single language; \textsc{SemRec} verifies
\emph{cross-language} equivalence.

\paragraph{Product programs.}
Barthe et al.~\cite{barthe2011product} introduced product programs
for relational verification of two runs of the \emph{same} program.
\textsc{SemRec} extends this to two \emph{different} programs in
\emph{different} languages with heterogeneous UB semantics,
adding $\sigma$-bridge coercion nodes at semantic boundaries---the
key mechanism enabling detection of IR-invisible divergences.


%% ======================================================================
\section{Conclusion}\label{sec:conclusion}
%% ======================================================================

We presented \textsc{SemRec}, a bounded source-level equivalence
verifier for C$\to$Rust migration that detects UB divergences
\emph{invisible to every existing verification approach}.  On 30
IR-invisible UB divergences spanning 6 classes, \textsc{SemRec}
achieves \textbf{100\% recall} while differential testing, IR
comparison, and IR-level verification tools (KLEE, SMACK, Alive2)
all score \textbf{0\%}.

The key enabler is the $\sigma$-bridge product program construction,
which keeps source and target semantic preconditions separate instead of
normalizing both languages through a shared compiler IR.  This preserves the
signed/unsigned, division, shift-UB, contract, and scoped memory-ordering
distinctions that compilation erases or under-specifies.  The soundness theorem
(Theorem~\ref{thm:soundness})
guarantees that UNSAT certifies equivalence on all well-defined
inputs within the loop bound.
When a positive witness is found, the compiler-IR localizer now complements the
proof by pointing maintainers at the exact \texttt{clang}-AST operation and
\texttt{rustc}-MIR statement whose semantics differ.

General-purpose accuracy (35.4\% on 511 pairs) reflects parser and IR
coverage limitations, not verification unsoundness.  Where the
pipeline completes, the formal guarantees hold.  We view
\textsc{SemRec} as complementary to differential testing: testing
covers broad program behavior but misses UB divergences by design;
\textsc{SemRec} catches exactly the bugs that testing cannot.

The tool is open-source, passes the targeted real-compiler regression suite, and is under active
development.  Future work includes interprocedural verification,
struct/enum lowering, and integration with C2Rust for automated
post-migration verification.


%% ======================================================================
%% BIBLIOGRAPHY
%% ======================================================================
\bibliographystyle{plain}
\begin{thebibliography}{35}

\bibitem{cisa2024sbd}
Cybersecurity and Infrastructure Security Agency.
\newblock Secure by Design pledge.
\newblock \url{https://www.cisa.gov/securebydesign/pledge}, 2024.

\bibitem{nsa2022memory}
National Security Agency.
\newblock Software memory safety.
\newblock Cybersecurity Information Sheet, November 2022.

\bibitem{whitehouse2024report}
Office of the National Cyber Director.
\newblock Back to the building blocks: A path toward secure and measurable software.
\newblock White House Technical Report, February 2024.

\bibitem{chromium2019memory}
Chromium Project.
\newblock Memory safety.
\newblock \url{https://www.chromium.org/Home/chromium-security/memory-safety/}, 2019.

\bibitem{microsoft2019safety}
M.~Miller.
\newblock Trends, challenges, and strategic shifts in the software vulnerability mitigation landscape.
\newblock Microsoft Security Response Center, 2019.

\bibitem{linuxrust2022}
M.~Ojeda et~al.
\newblock Rust for {Linux}.
\newblock \url{https://rust-for-linux.com/}, 2022.

\bibitem{c2rust2019}
P.~Emre, G.~Kondoh, and Z.~Anderson.
\newblock {C2Rust}: Migrating legacy code to {Rust}.
\newblock \emph{Proc.\ ICSE-Companion}, pp.\ 258--259, 2019.

\bibitem{zhang2023ownership}
H.~Zhang, J.~He, B.~Ding, and Y.~Zhang.
\newblock Ownership guided {C} to {Rust} translation.
\newblock \emph{Proc.\ OOPSLA}, 2023.

\bibitem{emre2021translating}
M.~Emre, R.~Schroeder, K.~Dewey, and B.~Hardekopf.
\newblock Translating {C} to safer {Rust}.
\newblock \emph{Proc.\ OOPSLA}, 2021.

\bibitem{iso9899}
{ISO/IEC}.
\newblock {ISO/IEC 9899:2011} --- {P}rogramming languages --- {C}.
\newblock International Organization for Standardization, 2011.

\bibitem{cadar2008klee}
C.~Cadar, D.~Dunbar, and D.~Engler.
\newblock {KLEE}: Unassisted and automatic generation of high-coverage tests for complex systems programs.
\newblock \emph{Proc.\ OSDI}, 2008.

\bibitem{rakamaric2014smack}
Z.~Rakamaric and M.~Emmi.
\newblock {SMACK}: Decoupling source language details from verifier implementations.
\newblock \emph{Proc.\ CAV}, 2014.

\bibitem{gurfinkel2015seahorn}
A.~Gurfinkel, T.~Kahsai, A.~Komuravelli, and J.~Navas.
\newblock The {SeaHorn} verification framework.
\newblock \emph{Proc.\ CAV}, 2015.

\bibitem{lopes2021alive2}
N.~P.~Lopes, J.~Lee, C.-K.~Hur, Z.~Liu, and J.~Regehr.
\newblock Alive2: Bounded translation validation for {LLVM}.
\newblock \emph{Proc.\ PLDI}, 2021.

\bibitem{pnueli1998translation}
A.~Pnueli, M.~Siegel, and E.~Singerman.
\newblock Translation validation.
\newblock \emph{Proc.\ TACAS}, pp.\ 151--166, 1998.

\bibitem{leroy2009compcert}
X.~Leroy.
\newblock Formal verification of a realistic compiler.
\newblock \emph{Communications of the ACM}, 52(7):107--115, 2009.

\bibitem{demoura2008z3}
L.~de~Moura and N.~Bj{\o}rner.
\newblock {Z3}: An efficient {SMT} solver.
\newblock \emph{Proc.\ TACAS}, 2008.

\bibitem{kroening2014cbmc}
D.~Kroening and M.~Tautschnig.
\newblock {CBMC} -- {C} bounded model checker.
\newblock \emph{Proc.\ TACAS}, 2014.

\bibitem{kani2022}
Amazon Web Services.
\newblock Kani {Rust} verifier.
\newblock \url{https://model-checking.github.io/kani/}, 2022.

\bibitem{astrauskas2022prusti}
V.~Astrauskas, A.~Bi\-le, P.~M\"uller, and F.~Poli.
\newblock Prusti: A practical verification infrastructure for {Rust}.
\newblock \emph{Proc.\ OOPSLA}, 2022.

\bibitem{jung2017rustbelt}
R.~Jung, J.-H.~Jourdan, R.~Krebbers, and D.~Dreyer.
\newblock {RustBelt}: Securing the foundations of the {Rust} programming language.
\newblock \emph{Proc.\ ACM Program.\ Lang.}, 2(POPL):66:1--66:34, 2018.

\bibitem{sammler2021refinedc}
M.~Sammler, R.~Lepigre, R.~Krebbers, et~al.
\newblock {RefinedC}: Automating the foundational verification of {C} code with refined ownership types.
\newblock \emph{Proc.\ PLDI}, 2021.

\bibitem{barthe2011product}
G.~Barthe, J.~M.~Crespo, and C.~Kunz.
\newblock Relational verification using product programs.
\newblock \emph{Proc.\ FM}, pp.\ 200--214, 2011.

\bibitem{treesitter2024}
M.~Brunsfeld et~al.
\newblock Tree-sitter: An incremental parsing system for programming tools.
\newblock \url{https://tree-sitter.github.io/}, 2024.

\bibitem{pratt1973top}
V.~Pratt.
\newblock Top down operator precedence.
\newblock \emph{Proc.\ POPL}, 1973.

\bibitem{cytron1991ssa}
R.~Cytron, J.~Ferrante, B.~K.~Rosen, M.~N.~Wegman, and F.~K.~Zadeck.
\newblock Efficiently computing static single assignment form and the control dependence graph.
\newblock \emph{ACM Trans.\ Program.\ Lang.\ Syst.}, 13(4):451--490, 1991.

\bibitem{andersen1994program}
L.~O.~Andersen.
\newblock Program analysis and specialization for the {C} programming language.
\newblock PhD thesis, DIKU, University of Copenhagen, 1994.

\bibitem{king1976symbolic}
J.~C.~King.
\newblock Symbolic execution and program testing.
\newblock \emph{Communications of the ACM}, 19(7):385--394, 1976.

\bibitem{reynolds2002separation}
J.~C.~Reynolds.
\newblock Separation logic: A logic for shared mutable data structures.
\newblock \emph{Proc.\ LICS}, pp.\ 55--74, 2002.

\end{thebibliography}


%% ======================================================================
\end{document}
